```{r}
# Clear the environment and read the prepared dataset
rm(list = ls()) # Clear the environment

library(tidyverse)
library(broom)
library(cbbdata)
library(hoopR)
library(lme4) # For mixed effects models
library(gtsummary) # Display the results of mixed effects models (tidy doesn't work)
library(mgcv) # For GAMs in R

df <- read.csv("player_stats.csv")
df
```

```{r}
# Change necessary variable data types for the models to work properly
df <- df |>
  mutate(id = as.factor(id),
         player = as.factor(player),
         pos = as.factor(pos),
         exp = factor(exp, levels = c("Fr", "So", "Jr", "Sr", "SSr")))
```

```{r}
# Split the data to transfers and non-transfers based on the variable `changed_team`

transfers <- df |>
  filter(changed_team == 1) |>
  filter(year > 2022)

non_transfers <- df |>
  filter(changed_team != 1 | is.na(changed_team)) |>
  filter(year > 2022)

nrow(transfers)
nrow(non_transfers)
```

First, I will fit some classic linear regression models. The covariates that I will consider in my model, thanks to the advisor's help, are:

- Team ratings
- Player ratings *(need work)*
- Position *(wait for Kendall)*
- Conference
- Experience level (may be duplicated)
- Year/Season

```{r}
# The simplest linear regression model
linreg_1 <- lm(ppg ~ prev_ppg, data = non_transfers)
tidy(linreg_1, conf.int = TRUE)
```

```{r}
# Multiple linear regression model (note the assumptions)
# This will take a lot of time because there are too many parameters to fit
linreg_2 <- lm(ppg ~ prev_ppg + id + team + pos + exp + conf +
                 year + current_barthag,
               data = non_transfers)

# Filter for the necessary terms to avoid showing too many parameters
tidy(linreg_2, conf.int = TRUE) |>
  filter(term %in% c("(Intercept)", "prev_ppg", "year", "current_barthag"))
```

```{r}
# Check for heteroskedasticity; also take a while to run because of many parameters
# Code taken from Lecture 23 slides
augment(linreg_2) |>
  ggplot(aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_smooth() +
  labs(x = "Fitted value", y = "Residual")+
  geom_hline(linetype = "dashed", color = "red", yintercept = 0)
```

```{r}
# Is the slope in the residual plot significantly different from zero?
augmented_linreg <- augment(linreg_2)
linreg_check <- lm(.resid ~ .fitted, data = augmented_linreg)
tidy(linreg_check, conf.int = TRUE)
```

The p-value is very high (approximately equal to 1), so the residual trends in the dataset does not show significant deviation from zero.

```{r}
# ADDITIONAL DIAGNOSTICS WITH QQ PLOTS
```

#### 1. Mixed effects model with random intercepts

Random intercepts (random effects) in player and team level, fixed effects in conference level.

```{r}
# https://stackoverflow.com/questions/67123529/how-to-include-nested-effects-in-mixed-effect-model-with-lme4-package
mixed_effects_1 <- lmer(ppg ~ prev_ppg + pos + exp + conf +
                 year + current_barthag + (1|id/team),
               data = non_transfers)

tbl_regression(mixed_effects_1)
```

Is the random effects even necessary at all, using an ANOVA test?

```{r}
comparison_lm_1 <- lm(ppg ~ prev_ppg + pos + exp + conf + year + current_barthag + id + team,
                      data = non_transfers)
comparison_mixed_1 <- lmer(ppg ~ prev_ppg + pos + exp + conf +
                             year + current_barthag + (1|id/team),
                           data = non_transfers, REML = FALSE)
anova(comparison_mixed_1, comparison_lm_1)
```

```{r}
# Note that nested models in lme4 should be specified from top to bottom level
comparison_lm_1 <- lm(ppg ~ prev_ppg + pos + exp + year + current_barthag,
                      data = non_transfers)

comparison_mixed_1 <- lmer(ppg ~ prev_ppg + pos + exp +
                             year + current_barthag + (1|id),
                           data = non_transfers, REML = FALSE)

comparison_mixed_2 <- lmer(ppg ~ prev_ppg + pos + exp +
                             year + current_barthag + (1|team/id),
                           data = non_transfers, REML = FALSE)

comparison_mixed_3 <- lmer(ppg ~ prev_ppg + pos + exp +
                             year + current_barthag + (1|conf/team/id),
                           data = non_transfers, REML = FALSE)

# Extract p-value from ANOVA tests
print("ANOVA p-value results between mixed effects models and fixed effects models:")
anova(comparison_mixed_1, comparison_lm_1)$`Pr(>Chisq)`[2]
anova(comparison_mixed_2, comparison_lm_1)$`Pr(>Chisq)`[2]
anova(comparison_mixed_3, comparison_lm_1)$`Pr(>Chisq)`[2]
print("")
print("ANOVA p-value results between different mixed effects models together:")
anova(comparison_mixed_1, comparison_mixed_2)$`Pr(>Chisq)`[2]
anova(comparison_mixed_2, comparison_mixed_3)$`Pr(>Chisq)`[2]
```

See [this link](https://stats.stackexchange.com/questions/114901/should-i-exclude-random-effects-from-a-model-if-they-are-not-statistically-signi) for some explanation of whether random effects is valid.

```{r}
lme4:::anova.merMod(comparison_mixed_1)
```

#### 2. Mixed effects model with random slopes

For the initial set-up, I will add random slopes to player and team level, and still fixed effects in conference level.

```{r}
#random_slopes <- lmer(ppg ~ prev_ppg + pos + exp + conf +
                        #year + current_barthag + (1 + prev_ppg| id/team),
                        #data = non_transfers)
#tbl_regression(random_slopes)

# Error: number of observations (=2214) <= number of random effects (=3776) for term (1 +
# prev_ppg | team:id); the random-effects parameters and the residual variance (or scale
# parameter) are probably unidentifiable

random_slopes_1 <- lmer(ppg ~ prev_ppg + pos + exp + conf +
                          year + current_barthag + (1 | id/team) + (1 + prev_ppg | team),
                        data = non_transfers)
tbl_regression(random_slopes_1)
```

#### 3. Applying the mixed effects model for the transferred players

```{r}
# https://optimumsportsperformance.com/blog/making-predictions-from-a-mixed-model-using-r/
# Select necessary columns from the transfers dataset

mixed_effects_1 <- lmer(ppg ~ prev_ppg + pos + exp + conf +
                          year + current_barthag + (1|id/team),
                        data = non_transfers)

transfers_subset <- transfers |>
  select(ppg, prev_ppg, pos, exp, conf, year, current_barthag, id, team)

transfers_augment <- predict(mixed_effects_1, newdata = transfers_subset, re.form = NA)

error_dataset <- transfers_subset |>
  mutate(expected_ppg = transfers_augment) |>
  mutate(prediction_error = ppg - expected_ppg) |>
  select(id, year, ppg, expected_ppg, prediction_error) |>
  arrange(desc(prediction_error))

error_dataset
```

```{r}
# The errors closely follow a normal distribution
# https://statisticsglobe.com/normal-density-curve-on-top-of-histogram-ggplot2-r

# after_stat(density) is used to convert the histogram from raw counts to density

error_dataset |>
  ggplot(aes(x = prediction_error)) +
  geom_histogram(aes(y = after_stat(density))) +
  geom_vline(xintercept = 0, color = "red", linewidth = 1, linetype = "dashed") +
  stat_function(fun = dnorm,
                args = list(mean = mean(error_dataset$prediction_error),
                            sd = sd(error_dataset$prediction_error)),
                color = "darkred",
                size = 1.5)
```

For better demonstration, we can also draw a normal distribution curve on top of this histogram and use some statistical tests to demonstrate that the distribution is normal.

```{r}
# Test if the prediction errors actually follow a normal distribution
library(dgof)
ks.test(error_dataset$prediction_error, "pnorm")
```

```{r}
error_dataset |>
  ggplot(aes(x = ppg, y = expected_ppg)) +
  geom_point() + coord_fixed() +
  geom_smooth() +
  geom_abline(intercept = 0, slope = 1, color = "red", linewidth = 1.5)
```

```{r}
error_dataset |>
  ggplot(aes(x = ppg, y = prediction_error)) +
  geom_point(alpha = 0.3) + coord_fixed() +
  geom_smooth() +
  geom_hline(yintercept = 0, color = "red", linewidth = 1, linetype = "dashed")
```

Players with higher actual points per game tend to be over-predicted.

```{r}
error_dataset |>
  ggplot(aes(x = expected_ppg, y = prediction_error)) +
  geom_point(alpha = 0.3) + coord_fixed() +
  geom_smooth() +
  geom_hline(yintercept = 0, color = "red", linewidth = 1, linetype = "dashed")
```

We can also make a "reverse EDA" to analyze errors:

```{r}
transfers_prediction_errors <- error_dataset |>
  left_join(df, join_by(id, year, ppg))

transfers_prediction_errors
```

Box plot of prediction errors with different categorical variables:

```{r}
# Differences in experience level
transfers_prediction_errors |>
  ggplot(aes(x = exp, y = prediction_error)) +
  geom_boxplot() +
  geom_hline(yintercept = 0, color = "red", linewidth = 1, linetype = "dashed")
```

```{r}
# Differences in position
transfers_prediction_errors |>
  ggplot(aes(x = pos, y = prediction_error)) +
  geom_boxplot() +
  geom_hline(yintercept = 0, color = "red", linewidth = 1, linetype = "dashed")
```

```{r}
# Differences in conference classification
transfers_prediction_errors |>
  mutate(conf_class = factor(conf_class, levels = c("low-major", "mid-major", "high-major"))) |>
  ggplot(aes(x = conf_class, y = prediction_error)) +
  geom_boxplot() +
  geom_hline(yintercept = 0, color = "red", linewidth = 1, linetype = "dashed")
```

The difference in conference classification plot is very interesting! The prediction error is somewhat negatively biased when the players advance to high-major classifications.

#### Appendix: GAMs to estimate the direction of relationship between current points per game with other useful predictors

```{r}
gam_1 <- gam(ppg ~ s(prev_ppg) + s(current_barthag) + pos + exp + conf + year,
             data = non_transfers, family = gaussian(link = "identity"))
```

```{r}
library(gratia)
draw(gam_1)
```

#### Appendix: Summary of the mixed effects reports

```{r}
summary(mixed_effects_1)
```

```{r}
#non_transfers |>
  #ggplot(aes(x = ppg)) +
  #geom_histogram()

summary(non_transfers$ppg)
```

```{r}
mixed_effects_conf <- lmer(ppg ~ prev_ppg + pos + exp + team +
                             year + current_barthag + (1|id),
                           data = non_transfers)

summary(mixed_effects_conf)
```

