```{r}
# Clear the environment and read the prepared dataset
rm(list = ls()) # Clear the environment
library(tidyverse)
library(broom)
library(cbbdata)
library(hoopR)
# library(performance) # More advanced model fit metrics for glms
library(mgcv) # For tweedie regression and GAMs in R
library(caret) # For tree-based models

df <- read.csv("player_stats.csv")
df
```

#### 1. EDA before modeling

Distribution of the outcome variable (currently points per game):

```{r}
df |>
  mutate(transfer_type = case_when(
    changed_team == 1 ~ "transfer",
    TRUE ~ "non_transfer"
  )) |>
  ggplot(aes(x = ppg)) +
  geom_density() +
  facet_wrap(~ transfer_type, ncol = 1)
```

```{r}
df |>
  mutate(transfer_type = case_when(
    changed_team == 1 ~ "transfer",
    TRUE ~ "non_transfer"
  )) |>
  filter(ppg == 0) |>
  group_by(transfer_type) |>
  count()
```

The proportion of zero observations (zero points per game) in the dataset is: `(971+45)/20117` = approx. 5%.

Since the number of points per game of the players are strictly non-negative and right skewed, we can also use generalized linear models such as gamma regression and tweedie regression (as 5% of zeroes might be a substantial part of the data, depending on the use case).

*(Also, check the specific assumptions for gamma and tweedie regression!)*

#### 2. Data preprocessing dedicated to machine/statistical learning algorithms

Splitting the dataset into athletes who have just transferred (`changed_team = 1`) and the rest:

```{r}
transfers <- df |>
  filter(changed_team == 1) |>
  filter(year > 2022)

non_transfers <- df |>
  filter(changed_team != 1 | is.na(changed_team)) |>
  filter(year > 2022)

nrow(transfers)
nrow(non_transfers)
```

How many `NA` values are there in the dataset of non-transfers (essential variables only)?

```{r}
colSums(is.na(non_transfers |>
                select(ppg, prev_ppg, pos, exp, current_barthag,
                       conf_class, has_transferred)))
```

We will build a regression model with cross-validation. Our goal is to closely predict performance as much as possible, and no inference is needed at the current stage.

As a reminder, the previous points per game takes the points per game from the last season of the player in NCAA Division I. We also temporarily filter out players in the 2021-22 season (`year == 2022`) because we do not currently have their previous season's points per game.

A major note in this dataset is that observations might be temporally related, because certain players are tracked across multiple years. We will split the dataset into training and validation folds based on the season in which player data was recorded:

```{r}
non_transfers_2022 <- non_transfers |>
  filter(year == 2022)

non_transfers_2023 <- non_transfers |>
  filter(year == 2023)

non_transfers_2024 <- non_transfers |>
  filter(year == 2024)
```

We determine the possible training and validation sets as follows, based on the year in which the data was recorded:

| Training set  | Validation set |
|:-------------:|:--------------:|
|     2022      |      2023      |
|     2022      | 2023 and 2024  |
|     2023      |      2024      |
| 2022 and 2023 |      2024      |

For better splits, I may need to get more data from the 2020-21 season on certain metrics, such as previous points per game and previous team rating.

#### 3. Assessing models by deviance explained

According to Lecture 16 slides, with a classic linear regression model, the deviance equals the residual sum of squares (RSS). We can also access the adjusted R^2^ (adjusted pseudo-R^2^ or McFadden's R^2^ in GLMs) to determine the variance or deviance explained by the model.

[A paper](https://pmc.ncbi.nlm.nih.gov/articles/PMC8279135/) actually suggested R^2^ is a better evaluation metric than RSS-derived metrics, but we will go over it later.

Here are the most basic baseline regressions to answer the question:

```{r}
# https://stackoverflow.com/questions/69305380/extract-list-name-passed-to-function-as-string-in-r

model_result <- function(model_var) {
  
  # Extract the model's name as specified by the user as a variable name
  sp_model_name <- deparse(substitute(model_var))
  
  # Create a tibble storing model name and deviance of the current model
  current_tibble <- tibble(
    model_name = sp_model_name,
    deviance_explained = 1 - (model_var$deviance / model_var$null.deviance)
  )
  
  return(current_tibble)
  
}
```

```{r}
null_model <- glm(ppg ~ 1, data = non_transfers,
                  family = gaussian(link = "identity"))

linreg_1 <- glm(ppg ~ prev_ppg, data = non_transfers,
                family = gaussian(link = "identity"))

linreg_2 <- glm(ppg ~ prev_ppg + pos + exp, data = non_transfers,
                family = gaussian(link = "identity"))

linreg_3 <- glm(ppg ~ prev_ppg + pos + exp + current_barthag +
                  conf_class + has_transferred,
                data = non_transfers,
                family = gaussian(link = "identity"))

model_perf_board <- bind_rows(
  model_result(null_model), model_result(linreg_1),
  model_result(linreg_2), model_result(linreg_3)
  )

model_perf_board
```

We can also extend our regression configuration with Poisson regression, tweedie regression, and Generalized Additive Models (GAMs), like this:

```{r}
tweedie_1 <- gam(ppg ~ prev_ppg, data = non_transfers,
                 family = tw(link = "identity"))

tweedie_2 <- gam(ppg ~ prev_ppg + pos + exp, data = non_transfers,
                 family = tw(link = "identity"))

tweedie_3 <- gam(ppg ~ prev_ppg + pos + exp + current_barthag +
                  conf_class + has_transferred,
                 data = non_transfers,
                 family = tw(link = "identity"))

# It seemed like categorical variables are not allowed to be in a smoothing spline
# to avoid raising errors when implementing a GAM

non_transfers_new <- non_transfers |>
  mutate(pos = as.factor(pos),
         exp = factor(exp, levels = c("Fr", "So", "Jr", "Sr")),
         conf_class = factor(conf_class,
                             levels = c("low-major", "mid-major", "high-major")),
         has_transferred = as.factor(has_transferred))

non_transfers_new$prev_ppg <- unlist(non_transfers_new$prev_ppg)

gam_1 <- gam(ppg ~ s(prev_ppg), data = non_transfers_new,
             family = gaussian(link = "identity"))

gam_2 <- gam(ppg ~ s(prev_ppg) + pos + exp,
             data = non_transfers_new, family = gaussian(link = "identity"))

gam_3 <- gam(ppg ~ s(prev_ppg) + s(current_barthag) + pos + exp + conf_class,
             data = non_transfers_new, family = gaussian(link = "identity"))

model_perf_board_2 <- bind_rows(
  model_perf_board,
  model_result(tweedie_1), model_result(tweedie_2), model_result(tweedie_3),
  model_result(gam_1), model_result(gam_2), model_result(gam_3)
  )

model_perf_board_2
```

I randomly found [this link](https://library.virginia.edu/data/articles/getting-started-tweedie-models-0) as a good reference for tweedie regression and its implementation in R.

Now we want to know the deviance explained with tree-based models, including: single decision trees, bagging, random forests, and boosting. Model trainings are done with 10-fold cross-validation.

```{r}
# This function is adapted from the "get_best_result" function from this website:
# https://daviddalpiaz.github.io/r4sl/the-caret-package.html#regression

model_result_tree <- function(tree_model_var) {
  
  # caret_fit$bestTune means the complexity parameter that yields the best results
  # which() gets the index of that row
  best <- which(
    rownames(tree_model_var$results) == rownames(tree_model_var$bestTune)
    )
  
  # Extract the row with the best result using the index
  # Then, remove the row index number from the dataframe
  best_result <- tree_model_var$results[best, ]
  rownames(best_result) <- NULL
  
  # Return the performance and add this value into the model performance board
  perf <- best_result$Rsquared
  
  # The rest of the implementation is similar to the model_result function
  sp_model_name <- deparse(substitute(tree_model_var))
  current_tibble <- tibble(
    model_name = sp_model_name,
    deviance_explained = perf
  )
  
  return(current_tibble)
}
```


```{r}
# https://daviddalpiaz.github.io/r4sl/the-caret-package.html#regression

# Single decision trees
single_tree_1 <- train(ppg ~ prev_ppg, method = "rpart", tuneLength = 50,
                       trControl = trainControl(method = "cv", number = 10),
                       data = non_transfers,
                       na.action = na.omit)

single_tree_2 <- train(ppg ~ prev_ppg + pos + exp,
                       method = "rpart", tuneLength = 50,
                       trControl = trainControl(method = "cv", number = 10),
                       data = non_transfers,
                       na.action = na.omit)

single_tree_3 <- train(ppg ~ prev_ppg + pos + exp + current_barthag +
                         conf_class + has_transferred,
                       method = "rpart", tuneLength = 50,
                       trControl = trainControl(method = "cv", number = 10),
                       data = non_transfers_new,
                       na.action = na.omit)

# Bagging, random forest, and boosting - Coming soon

model_perf_board_3 <- bind_rows(
  model_perf_board_2,
  model_result_tree(single_tree_1), model_result_tree(single_tree_2),
  model_result_tree(single_tree_3)
  ) |>
  arrange(desc(deviance_explained))

model_perf_board_3
```

[COMING SOON]

#### 4. Assessing models by predictive performance

First, I am trying to make the model as accurate as possible in predicting current performance of a player based on their past performance and some other factors. Personally, *(disclaimer: this need to be verified by statistical knowledge)*, variable selection on its own does not matter that much at this stage.

[COMING SOON]

#### Appendix: Using McFadden's adjusted R^2^ to compare between GLMs (WORKING CODE)

```{r, eval = FALSE}
# https://stats.oarc.ucla.edu/other/mult-pkg/faq/general/faq-what-are-pseudo-r-squareds/
# https://stackoverflow.com/questions/67286343/r-code-to-get-log-likelihood-for-binary-logistic-regression
```

Here are the most basic baseline regressions to answer the question:

```{r, eval = FALSE}
# https://stackoverflow.com/questions/69305380/extract-list-name-passed-to-function-as-string-in-r

add_model <- function(model_var, model_perf_board) {
  
  # Extract the model's name as specified by the user as a variable name
  sp_model_name <- deparse(substitute(model_var))
  
  # Compute the adjusted McFadden's R^2 using log likelihood
  k <- model_var$rank
  null_model <- update(model_var, . ~ 1)
  
  result <- 1 - (as.numeric(logLik(model_var)) - k) / as.numeric(logLik(null_model))
  result <- -log(result)
  
  # Create a tibble storing model name and performance of the current model
  current_tibble <- tibble(
    model_name = sp_model_name,
    r2_mcfadden_adjusted = result
  )
  
  # Add a model to an existing model performance board
  # Note that the model_perf_board needs to be initialized first at this stage
  r2_mcfadden_tibble <- bind_rows(model_perf_board, current_tibble)
  return(r2_mcfadden_tibble)
  
}
```

```{r, eval = FALSE}
# Initialize the R^2 McFadden tibble of different models
model_perf_board <- tibble(
  model_name = character(),
  r2_mcfadden_adjusted = numeric()
)

linreg_1 <- glm(ppg ~ prev_ppg, data = non_transfers,
                family = gaussian(link = "identity"))

linreg_2 <- glm(ppg ~ prev_ppg + pos + exp, data = non_transfers,
                family = gaussian(link = "identity"))

linreg_3 <- glm(ppg ~ prev_ppg + pos + exp + current_barthag +
                  conf_class + has_transferred,
                data = non_transfers,
                family = gaussian(link = "identity"))

model_perf_board <- add_model(linreg_1, model_perf_board)
model_perf_board <- add_model(linreg_2, model_perf_board)
model_perf_board <- add_model(linreg_3, model_perf_board)
model_perf_board
```

What about applying a tweedie regression with these variables, also using `glm`?

See [this link](https://library.virginia.edu/data/articles/getting-started-tweedie-models-0) for reference.

```{r, eval = FALSE}
tweedie_1 <- gam(ppg ~ prev_ppg, data = non_transfers,
                 family = tw(link = "identity"))

tweedie_2 <- gam(ppg ~ prev_ppg + pos + exp, data = non_transfers,
                 family = tw(link = "identity"))

tweedie_3 <- gam(ppg ~ prev_ppg + pos + exp + current_barthag +
                  conf_class + has_transferred,
                 data = non_transfers,
                 family = tw(link = "identity"))

model_perf_board <- add_model(tweedie_1, model_perf_board)
model_perf_board <- add_model(tweedie_2, model_perf_board)
model_perf_board <- add_model(tweedie_3, model_perf_board)
model_perf_board
```