```{r}
rm(list = ls()) # Clear the environment
library(tidyverse)
library(cbbdata)
library(hoopR)
```

#### 1. Dataset for player statistics by season

```{r}
# Read the player stats by season dataset
df_2022 <- cbd_torvik_player_season(year = 2022)
df_2023 <- cbd_torvik_player_season(year = 2023)
df_2024 <- cbd_torvik_player_season(year = 2024)
df_2025 <- cbd_torvik_player_season(year = 2025)

df <- bind_rows(df_2022, df_2023, df_2024, df_2025)
df
```

Update the conference classification and previous conference to the dataset (code by Kendall):

```{r}
df_enhanced <- df |>
  # Move the unique player ID to the front for readability first
  relocate(id) |>
  mutate(conf_class = case_when(
    conf %in% c("B10", "SEC", "ACC", "B12", "BE") ~ "high-major",
    conf %in% c("Amer", "A10", "CAA", "MVC", "MWC", "P12", "WCC") ~ "mid-major",
    conf %in% c("AE", "ASun", "BSky", "BSth", "BW", "CUSA", "Horz",
              "Ivy", "MAAC", "MAC", "MEAC", "NEC", "OVC", "Pat",
              "SC", "Slnd", "SWAC", "Sum", "SB", "WAC", "ind") ~ "low-major"
  ))

df_enhanced
```

Kendall helped me with the code for the `has_transferred` and `changed_team` variables!

```{r}
# Exploring Kendall's code (adjusted to group by player IDs instead of player names)

transfer_df <- df_enhanced |>
  group_by(id) |>
  filter(n_distinct(team) > 1) |>
  ungroup()

transfer_df
```

```{r}
# Whether a player has transferred - make it a binary variable for easier coding
df_enhanced <- df_enhanced |>
  mutate(has_transferred = ifelse(id %in% transfer_df$id, 1, 0))

df_enhanced
```

I have refined the code for the `changed_team` variable originally proposed by Kendall to account for players who transferred out of a NCAA Division I institution, or redshirted due to certain factors.

I also created new columns to better identify the transfer patterns for each student-athlete:

- `cumulative_career_transfers`: Cumulative number of college career transfers for each player. Each time a player transfers to a new team, the number increases by 1.

- `num_transfers`: The total number of transfers that a particular player has made throughout their entire college career (some people can have more than one transfers).

```{r}
career_transfers_by_player <- transfer_df |>
  select(id, year, team, exp, conf, conf_class, ppg) |>
  arrange(id, year) |>
  group_by(id) |>
  mutate(prev_team = lag(team),
         prev_exp = lag(exp),
         prev_conf = lag(conf),
         prev_conf_class = lag(conf_class),
         prev_ppg = lag(ppg),
         # Player's transfer history
         changed_team = ifelse(!(is.na(prev_team)) & (team != prev_team),
                                1, 0),
         cumulative_career_transfers = cumsum(changed_team),
         num_transfers = max(cumulative_career_transfers)) |>
  
  select(-team, -exp, -conf, -conf_class, -ppg)

career_transfers_by_player
```

Note that the previous points per game takes the points per game from the last season of the player in NCAA Division I.

Now rejoin these columns back into the dataset:

```{r}
df_enhanced <- df_enhanced |>
  left_join(career_transfers_by_player, join_by(id, year))

df_enhanced
```

I have also created another indicator variable, `skipped_year_before`, that denotes players who skipped one or more years in the Division I based on the lack of data. Some possible reasons for this are redshirting or transferring to non-DI institutions.

```{r}
year_presence_d1 <- df_enhanced |>
  select(id, year) |>
  arrange(id, year) |>
  group_by(id) |>
  mutate(
    
    # Comparing the player's previous year in NCAA DI and the year immediately
    # before to understand which player skipped one or more years in NCAA
         prev_d1_year = lag(year),
         year_before = year - 1,
         skipped_year_before = case_when(
           prev_d1_year == year_before ~ 0,
           is.na(prev_d1_year) ~ 0,
           prev_d1_year != year_before ~ 1
         )) |>
  
  # Separate mutate command to specify that we don't know whether a player has
  # skipped NCAA DI the year before if the year before is 2021
  
  mutate(skipped_year_before = case_when(
    year_before == 2021 ~ NA,
    TRUE ~ skipped_year_before
  ))

year_presence_d1
```

Again, join the new columns that I have created above back into the full dataset:

```{r}
df_enhanced <- df_enhanced |>
  left_join(year_presence_d1, join_by(id, year))

df_enhanced
```

We can check the code's accuracy by filtering the current `df_enhanced` dataset for a particular player ID that transfers the most, such as `65627` and `70342`.

Filling in the experience level for the remaining players:

- Jason Umosella (La Salle University): Freshman (as of 2024-25) ([link](https://goexplorers.com/sports/mens-basketball/roster/jason-umosella/12302))
- Fedor Zugic (Creighton University): Sophomore (as of 2024-25) ([link](https://gocreighton.com/sports/mens-basketball/roster/fedor-zcaronugi/8569))

```{r}
df_enhanced |>
  filter(!(exp %in% c("Fr", "So", "Jr", "Sr")))
```

```{r}
df_enhanced <- df_enhanced |>
  mutate(exp = case_when(
    player == "Jason Umosella" ~ "Fr",
    player == "Fedor Zugic" ~ "So",
    TRUE ~ exp
  ))

df_enhanced
```

```{r}
# Checking that the data preparation works
df_enhanced |>
  filter(player %in% c("Jason Umosella", "Fedor Zugic"))
```

```{r}
df_enhanced
```

Kendall helped me filter for super-seniors in the dataset, which is also an essential aspect to note related to the loopholes of NCAA transfer rules. The new indicator for super-seniors is denoted as `SSr`.

Note that we do not want to take the data for players in the 2020-21 season and prior because eligibility is heavily affected by the Covid-19 pandemic. Therefore, even if a player is a super-senior in the 2021-22 season, the dataset will still mark them as senior, or `Sr`.

I also created another indicator variable, `no_change_exp`, that yields 1 if the player does not change their experience (`Fr/So/Jr/Sr`) compared to the previous season they played in NCAA Division I, and 0 otherwise. This variable is expected to effectively spot both super-seniors and double freshmen/ sophomores/ juniors due to redshirting or transferring out of Division I.

```{r}
df_enhanced <- df_enhanced |>
  mutate(no_change_exp = case_when(
    exp == prev_exp ~ 1,
    is.na(prev_exp) ~ 0,
    TRUE ~ 0
  )) |>
  mutate(exp = ifelse(prev_exp == "Sr" & no_change_exp == 1, "SSr", exp))

df_enhanced
```

Side note: How many players are registered as double freshmen, sophomores, or juniors? (In the following code block, duplicated players are counted as multiple records.)

```{r}
df_enhanced |>
  select(id, year, exp, prev_exp, no_change_exp) |>
  filter(no_change_exp == 1 & prev_exp != "Sr") |>
  count(prev_exp)
```

Next, I will factor team ratings into the player statistics. In the future, I might also factor conference ratings in my model.

Since I have had some issues with accessing updated data from the `cbbdata` package (no longer actively maintained as of June 2026), I will use the `hoopR` package to access additional data, including team data.

I currently use Bart Torvik's T-Rank, which fits with the Bart Torvik API used in the `cbbdata` package, particularly in the team names. Using data from the same API source is very convenient in merging data where we do not have to deal with many differences in data entry.

```{r}
# The data is obtained from the hoopR package
# Note that the Torvik ratings for 2025 was unavailable

df_team_ratings_2022 <- torvik_ratings(year = 2022)
df_team_ratings_2023 <- torvik_ratings(year = 2023)
df_team_ratings_2024 <- torvik_ratings(year = 2024)
df_team_ratings_2025 <- torvik_ratings(year = 2025)

# Concatenate all datasets together and filter for only necessary rows

# Note that some team names in `df_team_ratings` are different from our previous
# dataset, `df_enhanced`; so we need to change their names to merge these two
# datasets better

df_team_ratings <- bind_rows(df_team_ratings_2022, df_team_ratings_2023,
                             df_team_ratings_2024, df_team_ratings_2025) |>
  select(year, team, barthag) |>
  rename(team_hoopR = team) |>
  mutate(team_cbbdata = case_when(
    team_hoopR == "Charleston" ~ "College of Charleston",
    team_hoopR == "LIU" ~ "LIU Brooklyn",
    team_hoopR == "North Carolina" ~ "North Carolina St.",
    team_hoopR == "Purdue Fort Wayne" ~ "Fort Wayne",
    team_hoopR == "East Texas A&M" ~ "Texas A&M Commerce",
    team_hoopR == "Detroit Mercy" ~ "Detroit",
    team_hoopR == "Saint Francis" ~ "St. Francis PA",
    team_hoopR == "Louisiana" ~ "Louisiana Lafayette",
    team_hoopR == "IU Indy" ~ "IUPUI",
    TRUE ~ team_hoopR
  )) |>
  select(-team_hoopR)

df_team_ratings
```

We also need to standardize the team names in the original dataset, `df_enhanced` as well:

```{r}
df_enhanced <- df_enhanced |>
  mutate(team = case_when(
    team == "Charleston" ~ "College of Charleston",
    team == "LIU" ~ "LIU Brooklyn",
    team == "North Carolina" ~ "North Carolina St.",
    team == "Purdue Fort Wayne" ~ "Fort Wayne",
    team == "East Texas A&M" ~ "Texas A&M Commerce",
    team == "Detroit Mercy" ~ "Detroit",
    team == "Saint Francis" ~ "St. Francis PA",
    team == "Louisiana" ~ "Louisiana Lafayette",
    team == "IU Indy" ~ "IUPUI",
    TRUE ~ team
  ))
```

I will create some more variables in the working dataframe, `df_enhanced`. This applies to all players including those who did not transfer. They should be blank for the 2022 season because we were not so sure about the player's previous team:

- `prev_year`: The previous season for the player, which can be represented as a formula: `prev_year = year - 1`. \[adjust later: add for redshirting or temporarily away from NCAA DI\]
- `prev_barthag_current`: The Barthag rating of the current team in the previous season (the "destination team" for player transfers).
- `current_barthag`: The Barthag rating of the current team in the current season.

```{r}
df_enhanced <- df_enhanced |>
  mutate(prev_year = year - 1) |>
  # Add the team rating in the current season
  left_join(df_team_ratings, join_by(year, team == team_cbbdata)) |>
  rename(current_barthag = barthag) |>
  # Add the team rating of the current team in the previous season
  left_join(df_team_ratings, join_by(prev_year == year, team == team_cbbdata)) |>
  rename(prev_barthag_current = barthag)

df_enhanced
```

We will check for null values in the team ranks as an indicator of whether our merging is valid, as spelling and styling variations create different value of team names that refer to the same institution:

```{r}
# https://www.datanovia.com/en/lessons/identify-and-remove-duplicate-data-in-r/
df_enhanced |>
  filter((is.na(current_barthag) | is.na(prev_barthag_current)) & (year != 2022)) |>
  select(team, year, current_barthag, prev_barthag_current) |>
  distinct(team, year, .keep_all = TRUE) |>
  arrange(team)
```

All observations in this output does not have previous Barthag score, because they did not compete in NCAA Division I men's basketball in the previous year (they were in Division II). Therefore, we can temporarily safely declare that the merge on standardized team names is valid.

For better modeling, we can also fill in these `NA` values with a zero Barthag score. According to [this blog post](https://adamcwisports.blogspot.com/p/every-possession-counts.html), the Barthag score represents a team's chance of winning against an *average* DI team, and it seems that (and AI said that) there is no team with a Barthag rating of exactly zero. The gap of athletic ability between Division I and lower-division teams is usually large, so we can safely impute zero for all the `NA` values that we have seen in the previous output.

```{r}
# https://tidyr.tidyverse.org/reference/replace_na.html
df_enhanced <- df_enhanced |>
  mutate(prev_barthag_current = case_when(
    year == 2022 ~ NA, # We do not have 2021 Barthag ratings for the teams yet
    year >= 2023 ~ replace_na(prev_barthag_current, 0)
  ))

df_enhanced
```

Finally, I will export the dataframe to a CSV file for later use (This should not be pushed into a public GitHub repo):

```{r}
write.csv(df_enhanced, "player_stats.csv", row.names = FALSE)
```

#### Appendix 1: Kendall's original code for the `changed_team` variable

```{r, eval = FALSE}
# Kendall's original code:

# trying to make transfer out year/transfer in year
# changed_team is the first year after transfer
#df_enhanced <- df_enhanced |>
  #arrange(id, year) |>
  #group_by(id) |>
  #mutate(prev_team = lag(team),
         #prev_year = lag(year),
         #if they have a previous team AND 
         #the year is the one right after the previous year AND 
         #the team they're on is not the same as the previous team THEN
         #changed_team is through IF NOT THEN
         #changed_team is false 
         #changed_team = ifelse(!is.na(prev_team) &
                                 #year == (prev_year + 1) &
                                 #team != prev_team,
                               #1,
                               #0))

#df_enhanced
```

#### Appendix 2: Kendall's original code for team conference classification variables

```{r, eval = FALSE}
#df_enhanced <- df_enhanced |>
  #mutate(conference_classification = case_when(
    #conf %in% c("B10", "SEC", "ACC", "B12", "BE") ~ "high-major",
    #conf %in% c("Amer", "A10", "CAA", "MVC", "MWC", "P12", "WCC") ~ "mid-major",
    #conf %in% c("AE", "ASun", "BSky", "BSth", "BW", "CUSA", "Horz",
              #"Ivy", "MAAC", "MAC", "MEAC", "NEC", "OVC", "Pat",
              #"SC", "Slnd", "SWAC", "Sum", "SB", "WAC", "ind") ~ "low-major"
  #))
```

```{r, eval = FALSE}
# bar graph for transfer players by conference
#df_enhanced <- df_enhanced |>
  #mutate(prev_conf_class = lag(conference_classification))

#df_enhanced <- df_enhanced |>
  #relocate(prev_conf_class)

#df_enhanced <- df_enhanced |>
  #mutate(prev_conf = lag(conf))

#df_enhanced <- df_enhanced |>
  #relocate(prev_conf)
```

#### Appendix 3: The interaction between transfer and year, built for certain types of models

Create new "transfer incoming year" variables to better fit the modeling process:

```{r, eval = FALSE}
#incoming_freshman_ids <- df_enhanced |>
  #filter(changed_team == 1 & exp == "Fr") |>
  #distinct(id) |>
  #mutate(transfer_incoming_fr = 1)

#incoming_sophomore_ids <- df_enhanced |>
  #filter(changed_team == 1 & exp == "So") |>
  #distinct(id) |>
  #mutate(transfer_incoming_so = 1)

#incoming_junior_ids <- df_enhanced |>
  #filter(changed_team == 1 & exp == "Jr") |>
  #distinct(id) |>
  #mutate(transfer_incoming_jr = 1)

#incoming_senior_ids <- df_enhanced |>
  #filter(changed_team == 1 & exp == "Sr") |>
  #distinct(id) |>
  #mutate(transfer_incoming_sr = 1)

#incoming_freshman_ids
#incoming_sophomore_ids
#incoming_junior_ids
#incoming_senior_ids
```

Now rejoin these columns back into the dataset and fill in the NA values:

```{r, eval = FALSE}
# https://tidyr.tidyverse.org/reference/replace_na.html

#df_enhanced <- df_enhanced |>
  #left_join(incoming_freshman_ids, join_by(id)) |>
  #left_join(incoming_sophomore_ids, join_by(id)) |>
  #left_join(incoming_junior_ids, join_by(id)) |>
  #left_join(incoming_senior_ids, join_by(id)) |>
  # Replace the NA values (for columns not present in the "right" table) with 0
  #mutate(transfer_incoming_fr = replace_na(transfer_incoming_fr, 0),
         #transfer_incoming_so = replace_na(transfer_incoming_so, 0),
         #transfer_incoming_jr = replace_na(transfer_incoming_jr, 0),
         #transfer_incoming_sr = replace_na(transfer_incoming_sr, 0))
  
```

#### Appendix 4 \[WORKING CODE\]: Mismatches between `hoopR` and `cbbdata` packages about team names

I also need to inspect the quality of the join later (such as imperfect matches between school names). However, the join looks quite accurate for now; I will fix these imperfect matches in the near future.

```{r, eval = FALSE}
# Look at NA values in the current T-Rank ratings:

# Filtered by team and year
df_enhanced |>
  ungroup() |>
  filter(is.na(current_t_rank_all)) |>
  distinct(year, team)

# Filtered by team names only
df_enhanced |>
  ungroup() |>
  filter(is.na(current_t_rank_all)) |>
  distinct(team)
```

Note on the `df_team_ratings` dataset for more accurate matching:

| Team name in `df_enhanced` | Team name in `df_team_ratings` |
|:--------------------------:|:------------------------------:|
|   College of Charleston    |           Charleston           |
|        LIU Brooklyn        |              LIU               |
|     North Carolina St.     |         North Carolina         |
|         Fort Wayne         |       Purdue Fort Wayne        |
|     Texas A&M Commerce     |         East Texas A&M         |
|          Detroit           |         Detroit Mercy          |
|       St. Francis PA       |         Saint Francis          |
|    Louisiana Lafayette     |           Louisiana            |
|           IUPUI            |            IU Indy             |

Some special notes about matching (for anyone who wants to ask):

- There is also an University of Charleston, but the institution does not participate in NCAA Division I men's basketball.

- In the `df_team_ratings` dataframe, the University of North Carolina has already been marked with the UNC initials.

- The Texas A&M University Commerce, as mentioned in the `df_enhanced` dataset, has been renamed to East Texas A&M University in 2024.

- The `df_team_ratings` dataframe contains both "Saint Francis" and "St. Francis NY" as two different schools. Using the process of elimination, we can see that the remaining Saint Francis school is located in Pennsylvania.

- Both Lafayette College and University of Louisiana at Lafayette compete in NCAA Division I men's basketball, so the matching is less certain in this case!

#### Appendix 5: Testing some datasets of team statistics

```{r, eval = FALSE}
# Team statistics by month
df2 <- try(cbd_torvik_team_split(year = 2022, split = "game_result"))
df2
table(df2$type)
```

```{r, eval = FALSE}
df3 <- try(cbd_torvik_ncaa_results(min_year=2022, max_year=2025, type='team'))
df3
```

```{r, eval = FALSE}
# Note that the Barthag ratings are only available before 2025...
df4 <- try(cbd_torvik_ratings(year = 2024))
df4
```

```{r, eval = FALSE}
# Confirmation that year = 2023 represents the 2022-23 season
# Other years also apply the similar rule
df_games <- try(cbd_torvik_game_box(year = 2023))
df_games
min(df_games$date)
max(df_games$date)
```

```{r, eval = FALSE}
# https://stackoverflow.com/questions/18931006/how-to-suppress-warning-messages-when-loading-a-library

# Note that you should download the package using the install.github command
# to get the most functionalities (according to my experience)
library(hoopR)

# Note that the API calls are expected to fail multiple times before retrieving
# the actual data. Use suppressMessages() to avoid printing unnecessary outputs

# The hoopR package also contains a "crosswalk" component, which helps users
# merge data from different sources with different team naming conventions together.
df_team_hoopr <- suppressMessages(
    mbb_team_crosswalk(season = 2022)
)

df_team_hoopr
```
