```{r}
# Read the dataset as usual
rm(list = ls()) # Clear the environment
library(tidyverse)
library(cbbdata)

df_2022 <- cbd_torvik_player_season(year = 2022)
df_2023 <- cbd_torvik_player_season(year = 2023)
df_2024 <- cbd_torvik_player_season(year = 2024)
df_2025 <- cbd_torvik_player_season(year = 2025)

df <- bind_rows(df_2022, df_2023, df_2024, df_2025)
df
```

Kendall helped me with the code for the `has_transferred` and `changed_team` variables!

```{r}
# Exploring Kendall's code (adjusted to group by player IDs instead of player names)

transfer_df <- df |>
  group_by(id) |>
  filter(n_distinct(team) > 1) |>
  ungroup()

transfer_df
```

```{r}
# Whether a player has transferred - make it a binary variable for easier coding
df_enhanced <- df |>
  mutate(has_transferred = ifelse(id %in% transfer_df$id, 1, 0))

df_enhanced
```


```{r}
# trying to make transfer out year/transfer in year
# changed_team is the first year after transfer
df_enhanced <- df_enhanced |>
  arrange(id, year) |>
  group_by(id) |>
  mutate(prev_team = lag(team),
         prev_year = lag(year),
         #if they have a previous team AND 
         #the year is the one right after the previous year AND 
         #the team they're on is not the same as the previous team THEN
         #changed_team is through IF NOT THEN
         #changed_team is false 
         changed_team = ifelse(!is.na(prev_team) &
                                 year == (prev_year + 1) &
                                 team != prev_team,
                               1,
                               0))

df_enhanced
```

Filling in the experience level for the remaining players:
- Jason Umosella (La Salle University): Freshman (as of 2024-25) ([link](https://goexplorers.com/sports/mens-basketball/roster/jason-umosella/12302))
- Fedor Zugic (Creighton University): Sophomore (as of 2024-25) ([link](https://gocreighton.com/sports/mens-basketball/roster/fedor-zcaronugi/8569))

```{r}
df_enhanced |>
  filter(!(exp %in% c("Fr", "So", "Jr", "Sr")))
```

```{r}
df_enhanced <- df_enhanced |>
  mutate(exp = case_when(
    player == "Jason Umosella" ~ "Fr",
    player == "Fedor Zugic" ~ "So",
    TRUE ~ exp
  ))

df_enhanced
```

```{r}
# Checking that the data preparation works
df_enhanced |>
  filter(player %in% c("Jason Umosella", "Fedor Zugic"))
```

Finally, I will export the dataframe to a CSV file for later use (This should not be pushed into a public GitHub repo):

```{r}
write.csv(df_enhanced, "player_stats.csv", row.names = FALSE)
```
