```{r}
# Read the dataset as usual
rm(list = ls()) # Clear the environment
library(tidyverse)
library(cbbdata)

df_2022 <- cbd_torvik_player_season(year = 2022)
df_2023 <- cbd_torvik_player_season(year = 2023)
df_2024 <- cbd_torvik_player_season(year = 2024)
df_2025 <- cbd_torvik_player_season(year = 2025)

df <- bind_rows(df_2022, df_2023, df_2024, df_2025)
df
```

Kendall helped me with the code for the `has_transferred` and `changed_team` variables!

```{r}
# Exploring Kendall's code (adjusted to group by player IDs instead of player names)

transfer_df <- df |>
  group_by(id) |>
  filter(n_distinct(team) > 1) |>
  ungroup()

transfer_df
```

```{r}
# Whether a player has transferred - make it a binary variable for easier coding
df_enhanced <- df |>
  mutate(has_transferred = ifelse(id %in% transfer_df$id, 1, 0))

df_enhanced
```


```{r}
# trying to make transfer out year/transfer in year
# changed_team is the first year after transfer
df_enhanced <- df_enhanced |>
  arrange(id, year) |>
  group_by(id) |>
  mutate(prev_team = lag(team),
         prev_year = lag(year),
         #if they have a previous team AND 
         #the year is the one right after the previous year AND 
         #the team they're on is not the same as the previous team THEN
         #changed_team is through IF NOT THEN
         #changed_team is false 
         changed_team = ifelse(!is.na(prev_team) &
                                 year == (prev_year + 1) &
                                 team != prev_team,
                               1,
                               0))

df_enhanced
```

Filling in the experience level for the remaining players:
- Jason Umosella (La Salle University): Freshman (as of 2024-25) ([link](https://goexplorers.com/sports/mens-basketball/roster/jason-umosella/12302))
- Fedor Zugic (Creighton University): Sophomore (as of 2024-25) ([link](https://gocreighton.com/sports/mens-basketball/roster/fedor-zcaronugi/8569))

```{r}
df_enhanced |>
  filter(!(exp %in% c("Fr", "So", "Jr", "Sr")))
```

```{r}
df_enhanced <- df_enhanced |>
  mutate(exp = case_when(
    player == "Jason Umosella" ~ "Fr",
    player == "Fedor Zugic" ~ "So",
    TRUE ~ exp
  ))

df_enhanced
```

```{r}
# Checking that the data preparation works
df_enhanced |>
  filter(player %in% c("Jason Umosella", "Fedor Zugic"))
```

Update the conference classification and previous conference to the dataset (code by Kendall):

```{r}
df_enhanced <- df_enhanced |>
  mutate(conference_classification = case_when(
    conf %in% c("B10", "SEC", "ACC", "B12", "BE") ~ "high-major",
    conf %in% c("Amer", "A10", "CAA", "MVC", "MWC", "P12", "WCC") ~ "mid-major",
    conf %in% c("AE", "ASun", "BSky", "BSth", "BW", "CUSA", "Horz",
              "Ivy", "MAAC", "MAC", "MEAC", "NEC", "OVC", "Pat",
              "SC", "Slnd", "SWAC", "Sum", "SB", "WAC", "ind") ~ "low-major"
  ))
```

```{r}
# bar graph for transfer players by conference
df_enhanced <- df_enhanced |>
  mutate(prev_conf_class = lag(conference_classification))

df_enhanced <- df_enhanced |>
  relocate(prev_conf_class)

df_enhanced <- df_enhanced |>
  mutate(prev_conf = lag(conf))

df_enhanced <- df_enhanced |>
  relocate(prev_conf)
```

Create new "transfer incoming year" variables to better fit the diff-in-diff modeling:

```{r}
incoming_freshman_ids <- df_enhanced |>
  filter(changed_team == 1 & exp == "Fr") |>
  distinct(id) |>
  mutate(transfer_incoming_fr = 1)

incoming_sophomore_ids <- df_enhanced |>
  filter(changed_team == 1 & exp == "So") |>
  distinct(id) |>
  mutate(transfer_incoming_so = 1)

incoming_junior_ids <- df_enhanced |>
  filter(changed_team == 1 & exp == "Jr") |>
  distinct(id) |>
  mutate(transfer_incoming_jr = 1)

incoming_senior_ids <- df_enhanced |>
  filter(changed_team == 1 & exp == "Sr") |>
  distinct(id) |>
  mutate(transfer_incoming_sr = 1)

incoming_freshman_ids
incoming_sophomore_ids
incoming_junior_ids
incoming_senior_ids
```

Now rejoin these columns back into the dataset and fill in the NA values:

```{r}
# https://tidyr.tidyverse.org/reference/replace_na.html

df_enhanced <- df_enhanced |>
  left_join(incoming_freshman_ids, join_by(id)) |>
  left_join(incoming_sophomore_ids, join_by(id)) |>
  left_join(incoming_junior_ids, join_by(id)) |>
  left_join(incoming_senior_ids, join_by(id)) |>
  # Replace the NA values (for columns not present in the "right" table) with 0
  mutate(transfer_incoming_fr = replace_na(transfer_incoming_fr, 0),
         transfer_incoming_so = replace_na(transfer_incoming_so, 0),
         transfer_incoming_jr = replace_na(transfer_incoming_jr, 0),
         transfer_incoming_sr = replace_na(transfer_incoming_sr, 0))
  
```

```{r}
df_enhanced
```


Finally, I will export the dataframe to a CSV file for later use (This should not be pushed into a public GitHub repo):

```{r}
write.csv(df_enhanced, "player_stats.csv", row.names = FALSE)
```
