```{r}
# Read the dataset as usual
rm(list = ls()) # Clear the environment
library(tidyverse)
library(cbbdata)

df_2022 <- cbd_torvik_player_season(year = 2022)
df_2023 <- cbd_torvik_player_season(year = 2023)
df_2024 <- cbd_torvik_player_season(year = 2024)
df_2025 <- cbd_torvik_player_season(year = 2025)

df <- bind_rows(df_2022, df_2023, df_2024, df_2025)
df
```

How many players have actually made a transfer?

```{r}
# Count the number of unique players by their ID
df |> count(id)
```

```{r}
df |>
  count(id) |>
  filter(n > 1)
```

Among 10435 unique players mentioned in the dataset, there are 5757 players who transferred from the 2021-2025 timeframe. Note that we have not yet covered players who actually had a transfer but outside of the time frame.

For now, we will add the new `transferred` variable to our dataframe, which returns 0 if player has not made a transfer in their entire career in that particular time frame, and 1 otherwise.

```{r}
# Filter for player IDs who transferred
# Note that player IDs stay constant throughout the years

transferred_ids <- df |>
  count(id) |>
  filter(n > 1) |>
  select(-n) |>
  mutate(transferred = 1)

transferred_ids
```

```{r}
# Add this into the concatenated dataframe
# https://www.r-bloggers.com/2022/06/replace-na-with-zero-in-r/

df_enhanced <- df |>
  left_join(transferred_ids, join_by(id)) |>
  mutate(transferred = ifelse(is.na(transferred), 0, transferred))

df_enhanced
```

```{r}
# Additional sanity check for "transferred IDs" and not transferred
df_enhanced |>
  distinct(id, .keep_all = TRUE) |>
  count(transferred)
```

#### 1. Plot the 1D and 2D distributions of variables

Summarize all variables in the dataset:

```{r}
summary(df)
```

Many quantitative variables look very left-skewed! How about examining the skewness of each variable?

```{r}
# I will compute them quantitatively later...
```


I will plot the histogram of most quantitative variable indicators of player performance in this dataset, as some inferece models ([like hypothesis testing in linear regression models](https://library.virginia.edu/data/articles/normality-assumption)) prefer normally distributed variables.

```{r}
# https://stackoverflow.com/questions/63334436/pivot-longer-on-all-columns
df_enhanced |>
  select(-(player:conf), -year, -id, -pick, -transferred) |>
  select(1:20) |>
  pivot_longer(cols = everything(), names_to = "metric",
               values_to = "value") |>
  ggplot(aes(x = value)) +
  geom_density() +
  facet_wrap(~ metric, scales = "free")
```

```{r}
# Continue plotting for other relevant quantitative variables
df_enhanced |>
  select(-(player:conf), -year, -id, -pick, -transferred) |>
  select(21:40) |>
  pivot_longer(cols = everything(), names_to = "metric",
               values_to = "value") |>
  ggplot(aes(x = value)) +
  geom_density() +
  facet_wrap(~ metric, scales = "free")
```

```{r}
# Continue plotting for other relevant quantitative variables
# There are 55 quantitative variables in total
df_enhanced |>
  select(-(player:conf), -year, -id, -pick, -transferred) |>
  select(41:55) |>
  pivot_longer(cols = everything(), names_to = "metric",
               values_to = "value") |>
  ggplot(aes(x = value)) +
  geom_density() +
  facet_wrap(~ metric, scales = "free")
```

Future work:

- Check the correlation between variables with a heat map