```{r}
# Clear the environment and read the prepared dataset
rm(list = ls()) # Clear the environment
library(tidyverse)
library(cbbdata)

df <- read.csv("player_stats.csv")
df
```

#### 1. Plot the 1D and 2D distributions of variables

Summarize all variables in the dataset:

```{r}
summary(df)
```

Many quantitative variables look very left-skewed! How about examining the skewness of each variable?

```{r}
# I will compute them quantitatively later...
```


I will plot the histogram of most quantitative variable indicators of player performance in this dataset, as some inferece models ([like hypothesis testing in linear regression models](https://library.virginia.edu/data/articles/normality-assumption)) prefer normally distributed variables.

```{r}
quantitative_df <- df |>
  select(-(player:conf), -year, -id, -pick, -has_transferred, -prev_team,
         -prev_year, -changed_team)

quantitative_df
```


```{r}
# https://stackoverflow.com/questions/63334436/pivot-longer-on-all-columns
quantitative_df |>
  select(1:20) |>
  pivot_longer(cols = everything(), names_to = "metric",
               values_to = "value") |>
  ggplot(aes(x = value)) +
  geom_density() +
  facet_wrap(~ metric, scales = "free")
```

```{r}
# Continue plotting for other relevant quantitative variables
quantitative_df |>
  select(21:40) |>
  pivot_longer(cols = everything(), names_to = "metric",
               values_to = "value") |>
  ggplot(aes(x = value)) +
  geom_density() +
  facet_wrap(~ metric, scales = "free")
```

```{r}
# Continue plotting for other relevant quantitative variables
# There are 55 quantitative variables in total
quantitative_df |>
  select(41:55) |>
  pivot_longer(cols = everything(), names_to = "metric",
               values_to = "value") |>
  ggplot(aes(x = value)) +
  geom_density() +
  facet_wrap(~ metric, scales = "free")
```

Check the correlation between relevant quantitative variables with a heat map:

```{r}
# https://www.sthda.com/english/wiki/ggplot2-quick-correlation-matrix-heatmap-r-software-and-data-visualization

# install.packages(reshape2)
library(reshape2)

corr_matrix <- quantitative_df |>
  cor(use = "pairwise.complete.obs") |>
  round(2) |>
  melt() |>
  # Highlight the correlations that are greater than 0.85 in absolute value
  # (both positive and negative)
  mutate(plot_value = ifelse(value > 0.85 | value < -0.85, value, NA))
  
corr_matrix
```

```{r}
corr_matrix |>
  ggplot(aes(x = Var1, y = Var2, fill = plot_value)) +
  geom_tile()
```

This one looks messy, but things will clear up when we know which variables to include in our model. For now, there are some pairs of variables with high correlation.

#### 2. Explore differences in player metrics based on their transfer status

As a refresher, the documentation for the original dataset, `torvik_player_season`, can be found [here](https://github.com/andreweatherman/cbbdata/blob/main/data/dictionary_torvik_player_season.csv).

Comparing players who have transferred in their careers and players who do not:

```{r}
# Note that we have removed some NA values to properly display the results
df |>
  select(g, min, mpg, ppg, has_transferred) |>
  group_by(has_transferred) |>
  summarize(median_g = median(g, na.rm = TRUE),
            median_min = median(min, na.rm = TRUE),
            median_mpg = median(mpg, na.rm = TRUE),
            median_ppg = median(ppg, na.rm = TRUE)) |>
  pivot_longer(median_g:median_ppg, names_to = "metric", values_to = "value") |>
  ggplot(aes(x = has_transferred, y = value)) +
  geom_col() + coord_flip() +
  scale_x_continuous(breaks = seq(0, 1, 1)) +
  facet_wrap(~ metric, scales = "free")
```

Are these differences statistically significant? Perhaps we can use a statistical test later...

```{r}
# Add code to perform the statistical test here
```

[MORE COMING SOON]

#### 3. Explore the transferring data for teams and conferences

Which specific team has the most number of transfer-ins and transfer-outs during the 2021-25 periods and for each season?

(Note that we do not have data outside of the 2021-25 range. The transfer-ins data starts from the 2022-23 season and the transfer-outs data ends at the 2023-24 season.)

```{r}
# Filter the dataset to only obtain athletes' first year after transfer
transfer_ins <- df |>
  filter(changed_team == 1) |>
  select(team, year)

transfer_ins
```

```{r}
# Teams with the most transfer-ins each year
transfer_ins |>
  group_by(team, year) |>
  count(team) |>
  ungroup() |>
  group_by(year) |>
  slice_max(n, n = 10) |>
  ggplot(aes(x = n, y = fct_reorder(team, n))) +
  geom_col() + facet_wrap(~ year, scales = "free_x")
```

What about the teams which have the most players transferring out?

```{r}
df |>
  filter(changed_team == 1) |>
  select(prev_team, prev_year) |>
  group_by(prev_team, prev_year) |>
  count(prev_team) |>
  ungroup() |>
  group_by(prev_year) |>
  slice_max(n, n = 5) |>
  ggplot(aes(x = n, y = fct_reorder(prev_team, n))) +
  geom_col() + facet_wrap(~ prev_year, scales = "free_x")
```

In the presentation, we might link these EDA graph's result to management or coaching changes. Additional research is needed to confirm if that is the case.

[MORE COMING SOON]

#### 4. Explore the general developments of player statistics over time ("time trends")

[COMING SOON]