Note that the `has_transferred` and `changed_team` variable in this file is obtained from Kendall's code.

```{r}
# Read the dataset as usual
rm(list = ls()) # Clear the environment
library(tidyverse)
library(cbbdata)

df_2022 <- cbd_torvik_player_season(year = 2022)
df_2023 <- cbd_torvik_player_season(year = 2023)
df_2024 <- cbd_torvik_player_season(year = 2024)
df_2025 <- cbd_torvik_player_season(year = 2025)

df <- bind_rows(df_2022, df_2023, df_2024, df_2025)
df
```

```{r}
# Exploring Kendall's code (adjusted to group by player IDs instead of player names)

transfer_df <- df |>
  group_by(id) |>
  filter(n_distinct(team) > 1) |>
  ungroup()

transfer_df
```

```{r}
# Whether a player has transferred - make it a binary variable for easier coding
df_enhanced <- df |>
  mutate(has_transferred = ifelse(id %in% transfer_df$id, 1, 0))

df_enhanced
```


```{r}
# trying to make transfer out year/transfer in year
# changed_team is the first year after transfer
df_enhanced <- df_enhanced |>
  arrange(id, year) |>
  group_by(id) |>
  mutate(prev_team = lag(team),
         prev_year = lag(year),
         #if they have a previous team AND 
         #the year is the one right after the previous year AND 
         #the team they're on is not the same as the previous team THEN
         #changed_team is through IF NOT THEN
         #changed_team is false 
         changed_team = ifelse(!is.na(prev_team) &
                                 year == (prev_year + 1) &
                                 team != prev_team,
                               1,
                               0))

df_enhanced
```

```{r}
df_enhanced |>
  select(id, team, year, has_transferred, changed_team)
```

My modeling approach is inspired from [this website](https://www.tilburgsciencehub.com/topics/analyze/causal-inference/did/canonical-did-regression/), which contains a brief description of how to implement difference-in-differences (DiD) in R.

```{r}
# Most simple baseline model, using points per game (ppg) as an example metric
baseline_lm <- lm(ppg ~ changed_team, data = df_enhanced)
summary(baseline_lm)
```

```{r}
# Controlling for player and year fixed effects
df_enhanced <- df_enhanced |>
  mutate(id = as.factor(id),
         year = as.factor(year))

df_enhanced
```

```{r}
# Fixed effects model
# install.packages("fixest")
library(fixest)
fe_model <- feols(ppg ~ changed_team | year, data = df_enhanced)
fe_model
```


```{r}
# Now let's add the near-DiD set up without the 2-way fixed effects
near_did_lm <- lm(ppg ~ has_transferred + changed_team
                  + has_transferred*changed_team,
                  data = df_enhanced)
summary(near_did_lm)
```

```{r}
cor(df_enhanced$has_transferred*df_enhanced$changed_team, df_enhanced$changed_team)
```

Note the error "1 not defined because of singularities". There is currently a problem with implementing DiD in this case. We have created two indicator variables:

- The group indicator variable (has not transferred in career vs. has transferred)
- The treatment indicator variable (before transfer vs. first year after transfer)

This might contribute to the perfect collinearity issue! (once we see `has_transferred == 0`, we can also conclude that `changed_team == 0`.)

[Future work] Another path to address the perfect collinearity: Set the `has_transferred` variable to be `1` only when the player has made their first transfer, and `0` otherwise.

#### Additional control variables to consider in the model

Some confounding variables (the best that we can find) in the dataset to model the relationship between transferring (`changed_team`) and athletic performance:

- Experience level (freshman, sophomore, junior, senior)
- Differences in ratings or results between destination team and original team
- The same differences but between destination and original conferences
- Postion (well there are many crazy values here)
- Whether they are good or bad within the current team

Other control variables that we might include:

- NIL (the one we all know but currently do not have sufficient data)

```{r}
# Note that the effect of `changed_team` has changed dramatically when we added the position into the model

improved_fe_model <- feols(ppg ~ changed_team | id + year,
                           data = df_enhanced)
improved_fe_model
```

(Currently having a singleton issue associated with the model if we use the student fixed effect `id`)

Fill in the position for the remaining players:
- Jason Umosella (La Salle University): Freshman (as of 2024-25) ([link](https://goexplorers.com/sports/mens-basketball/roster/jason-umosella/12302))
- Fedor Zugic (Creighton University): Sophomore (as of 2024-25) ([link](https://gocreighton.com/sports/mens-basketball/roster/fedor-zcaronugi/8569))