```{r}
# Clear the environment and read the prepared dataset
rm(list = ls()) # Clear the environment
library(tidyverse)
library(cbbdata)
library(broom)
library(hoopR)

df <- read.csv("player_stats.csv")
df
```

#### 1. EDA before modeling

Distribution of the outcome variable (currently points per game):

```{r}
df |>
  mutate(transfer_type = case_when(
    changed_team == 1 ~ "transfer",
    TRUE ~ "non_transfer"
  )) |>
  ggplot(aes(x = ppg)) +
  geom_density() +
  facet_wrap(~ transfer_type, ncol = 1)
```

```{r}
df |>
  mutate(transfer_type = case_when(
    changed_team == 1 ~ "transfer",
    TRUE ~ "non_transfer"
  )) |>
  filter(ppg == 0) |>
  group_by(transfer_type) |>
  count()
```

The proportion of zero observations (zero points per game) in the dataset is: `(971+45)/20117` = approx. 5%.

Since the number of points per game of the players are strictly non-negative and right skewed, we can also use generalized linear models such as gamma regression and tweedie regression (as 5% of zeroes might be a substantial part of the data, depending on the use case).

#### 2. Data preprocessing

Splitting the dataset into athletes who have just transferred (`changed_team = 1`) and the rest:

```{r}
transfers <- df |>
  filter(changed_team == 1) |>
  filter(year > 2022)

non_transfers <- df |>
  filter(changed_team != 1 | is.na(changed_team)) |>
  filter(year > 2022)

nrow(transfers)
nrow(non_transfers)
```

We will build a regression model with cross-validation. Our goal is to closely predict performance as much as possible, and no inference is needed at the current stage.

As a reminder, the previous points per game takes the points per game from the last season of the player in NCAA Division I. We also temporarily filter out players in the 2021-22 season (`year == 2022`) because we do not currently have their previous season's points per game.

#### 1. Linear regression

The most basic baseline regression to answer the question:

```{r}
linreg_1 <- lm(ppg ~ prev_ppg, data = non_transfers)
tidy(linreg_1, conf.int = TRUE)
```