```{r}
# Clear the environment and read the prepared dataset
rm(list = ls()) # Clear the environment

library(tidyverse)
library(broom)
library(cbbdata)
library(hoopR)
library(lme4) # For mixed effects models
library(gtsummary) # Display the results of mixed effects models (tidy doesn't work)

df <- read.csv("player_stats.csv")
df
```

```{r}
# Change necessary variable data types for the models to work properly
df <- df |>
  mutate(id = as.factor(id),
         player = as.factor(player),
         pos = as.factor(pos),
         exp = factor(exp, levels = c("Fr", "So", "Jr", "Sr", "SSr")))
```

```{r}
# Split the data to transfers and non-transfers based on the variable `changed_team`

transfers <- df |>
  filter(changed_team == 1) |>
  filter(year > 2022)

non_transfers <- df |>
  filter(changed_team != 1 | is.na(changed_team)) |>
  filter(year > 2022)

nrow(transfers)
nrow(non_transfers)
```

First, I will fit some classic linear regression models. The covariates that I will consider in my model, thanks to the advisor's help, are:

- Team ratings
- Player ratings *(need work)*
- Position *(wait for Kendall)*
- Conference
- Experience level (may be duplicated)
- Year/Season

```{r}
# The simplest linear regression model
linreg_1 <- lm(ppg ~ prev_ppg, data = non_transfers)
tidy(linreg_1, conf.int = TRUE)
```

```{r}
# Multiple linear regression model (note the assumptions)
# This will take a lot of time because there are too many parameters to fit
linreg_2 <- lm(ppg ~ prev_ppg + id + team + pos + exp + conf +
                 year + current_barthag,
               data = non_transfers)

# Filter for the necessary terms to avoid showing too many parameters
tidy(linreg_2, conf.int = TRUE) |>
  filter(term %in% c("(Intercept)", "prev_ppg", "year", "current_barthag"))
```

```{r}
# Check for heteroskedasticity; also take a while to run because of many parameters
# Code taken from Lecture 23 slides
augment(linreg_2) |>
  ggplot(aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_smooth() +
  labs(x = "Fitted value", y = "Residual")+
  geom_hline(linetype = "dashed", color = "red", yintercept = 0)
```

```{r}
# Is the slope in the residual plot significantly different from zero?
augmented_linreg <- augment(linreg_2)
linreg_check <- lm(.resid ~ .fitted, data = augmented_linreg)
tidy(linreg_check, conf.int = TRUE)
```

The p-value is very high (approximately equal to 1), so the residual trends in the dataset does not show significant deviation from zero.

```{r}
# ADDITIONAL DIAGNOSTICS WITH QQ PLOTS
```

#### 1. Mixed effects model with random intercepts

Random intercepts (random effects) in player and team level, fixed effects in conference level.

```{r}
# https://stackoverflow.com/questions/67123529/how-to-include-nested-effects-in-mixed-effect-model-with-lme4-package
mixed_effects_1 <- lmer(ppg ~ prev_ppg + pos + exp + conf +
                 year + current_barthag + (1|id/team),
               data = non_transfers)

tbl_regression(mixed_effects_1)
```

Is the random effects even necessary at all, using an ANOVA test?

```{r}
# ANOVA TEST CODE HERE
```

#### 2. Mixed effects model with random slopes

For the initial set-up, I will add random slopes to player and team level, and still fixed effects in conference level.

[COMING SOON]

#### Appendix: GAMs to estimate the direction of relationship between current points per game with other useful predictors

[COMING SOON]