```{r}
# Clear the environment and read the prepared dataset
rm(list = ls()) # Clear the environment
library(tidyverse)
library(cbbdata)

df <- read.csv("player_stats.csv")
df
```

```{r}
# Cast the "numeric" variables (id and year) into factor for more convenient modeling
df <- df |>
  mutate(id = as.factor(id),
         year = as.factor(year))

df
```

My modeling approach is inspired from [this website](https://www.tilburgsciencehub.com/topics/analyze/causal-inference/did/canonical-did-regression/), which contains a brief description of how to implement difference-in-differences (DiD) in R.

```{r}
# Most simple baseline model, using points per game (ppg) as an example metric
baseline_lm <- lm(ppg ~ changed_team, data = df)
summary(baseline_lm)
```

```{r}
# Fixed effects model
# install.packages("fixest")
library(fixest)
fe_model <- feols(ppg ~ changed_team | year, data = df)
fe_model
```


```{r}
# Now let's add the near-DiD set up without the 2-way fixed effects
near_did_lm <- lm(ppg ~ has_transferred + changed_team
                  + has_transferred*changed_team,
                  data = df)
summary(near_did_lm)
```

```{r}
cor(df$has_transferred*df$changed_team, df$changed_team)
```

Note the error "1 not defined because of singularities". There is currently a problem with implementing DiD in this case. We have created two indicator variables:

- The group indicator variable (has not transferred in career vs. has transferred)
- The treatment indicator variable (before transfer vs. first year after transfer)

This might contribute to the perfect collinearity issue! (once we see `has_transferred == 0`, we can also conclude that `changed_team == 0`.)

[Future work] Another path to address the perfect collinearity: Set the `has_transferred` variable to be `1` only when the player has made their first transfer, and `0` otherwise.

#### Additional control variables to consider in the model

Some confounding variables (the best that we can find) in the dataset to model the relationship between transferring (`changed_team`) and athletic performance:

- Experience level (freshman, sophomore, junior, senior)
- Differences in ratings or results between destination team and original team
- The same differences but between destination and original conferences
- Postion (well there are many crazy values here)
- Whether they are good or bad within the current team

Other control variables that we might include:

- NIL (the one we all know but currently do not have sufficient data)

```{r}
# Note that the effect of `changed_team` has changed dramatically when we added the position into the model

improved_fe_model <- feols(ppg ~ changed_team | id + year,
                           data = df)
improved_fe_model
```

(Currently having a singleton issue associated with the model if we use the student fixed effect `id`)