```{r}
# Clear the environment and read the prepared dataset
rm(list = ls()) # Clear the environment
library(tidyverse)
library(cbbdata)
library(broom)

df <- read.csv("player_stats.csv")
df
```

```{r}
# Cast the "numeric" variables (id and year) into factor for more convenient modeling
df <- df |>
  mutate(id = as.factor(id),
         year = as.factor(year))

df
```

#### 1. Basic linear regression and diff-in-diff set-up

My modeling approach is inspired from [this website](https://www.tilburgsciencehub.com/topics/analyze/causal-inference/did/canonical-did-regression/), which contains a brief description of how to implement difference-in-differences (DiD) in R.

```{r}
# Most simple baseline model, using points per game (ppg) as an example metric
baseline_lm <- lm(ppg ~ changed_team, data = df)
summary(baseline_lm)
```

```{r}
# Fixed effects model
# install.packages("fixest")
library(fixest)
fe_model <- feols(ppg ~ changed_team | year, data = df)
fe_model
```

```{r}
# Now let's add the near-DiD set up without the 2-way fixed effects
near_did_lm <- lm(ppg ~ has_transferred + changed_team
                  + has_transferred*changed_team,
                  data = df)
summary(near_did_lm)
```

```{r}
cor(df$has_transferred*df$changed_team, df$changed_team)
```

Note the error "1 not defined because of singularities". There is currently a problem with implementing DiD in this case. We have created two indicator variables:

- The group indicator variable (has not transferred in career vs. has transferred)
- The treatment indicator variable (before transfer vs. first year after transfer)

This might contribute to the perfect collinearity issue! (once we see `has_transferred == 0`, we can also conclude that `changed_team == 0`.)

[Future work] Another path to address the perfect collinearity: Set the `has_transferred` variable to be `1` only when the player has made their first transfer, and `0` otherwise.

#### 2. Isolating the diff-in-diff by transferring times (eg: junior/senior year)

I will run a DiD model filtered by the incoming transferring experience level, without any additional fixed effects or control variables.

```{r}
# Sample: DiD with players transferring in their senior year
near_did_sr <- lm(ppg ~ transfer_incoming_sr + changed_team
                  + transfer_incoming_sr*changed_team,
                  data = df)
tidy(near_did_sr, conf.int = TRUE)
```

```{r}
near_did_so <- lm(ppg ~ transfer_incoming_so + changed_team
                  + transfer_incoming_so*changed_team,
                  data = df)
tidy(near_did_so, conf.int = TRUE)
```

```{r}
# Report multiple regression model results at a time in a concise table
# https://libguides.princeton.edu/R-stargazer
# install.packages("stargazer")
library(stargazer)

near_did_fr <- lm(ppg ~ transfer_incoming_fr + changed_team
                  + transfer_incoming_fr*changed_team,
                  data = df)

near_did_so <- lm(ppg ~ transfer_incoming_so + changed_team
                  + transfer_incoming_so*changed_team,
                  data = df)

near_did_jr <- lm(ppg ~ transfer_incoming_jr + changed_team
                  + transfer_incoming_jr*changed_team,
                  data = df)

near_did_sr <- lm(ppg ~ transfer_incoming_sr + changed_team
                  + transfer_incoming_sr*changed_team,
                  data = df)

stargazer(near_did_fr, near_did_so, near_did_jr, near_did_sr, type = "text")
```

What about using another package, `modelsummary`? ([link](https://www.tilburgsciencehub.com/topics/visualization/data-visualization/regression-results/model-summary/))

```{r}
# install.packages("modelsummary")
library(modelsummary)

models <- list(
  "(1)" = near_did_fr,
  "(2)" = near_did_so,
  "(3)" = near_did_jr,
  "(4)" = near_did_sr
)

# stars = TRUE: add stars (***) to denote statistical significance
msummary(models, stars = TRUE)
```

Adding fixed effects with regard to team and season (year), along with control variables:

```{r}
library(fixest)

near_did_fr_2 <- feols(ppg ~ changed_team + pos
                       + transfer_incoming_fr
                       + transfer_incoming_fr*changed_team
                       | team + year,
                  data = df)

near_did_so_2 <- feols(ppg ~ changed_team + pos
                       + transfer_incoming_so
                       + transfer_incoming_so*changed_team
                       | team + year,
                  data = df)

near_did_jr_2 <- feols(ppg ~ changed_team + pos
                       + transfer_incoming_jr
                       + transfer_incoming_jr*changed_team
                       | team + year,
                  data = df)

near_did_sr_2 <- feols(ppg ~ changed_team + pos
                       + transfer_incoming_sr
                       + transfer_incoming_sr*changed_team
                       | team + year,
                  data = df)

models_2 <- list(
  "(1)" = near_did_fr_2,
  "(2)" = near_did_so_2,
  "(3)" = near_did_jr_2,
  "(4)" = near_did_sr_2
)

msummary(models_2, stars = TRUE)
```

#### Additional control variables to consider in the model

Some confounding variables (the best that we can find) in the dataset to model the relationship between transferring (`changed_team`) and athletic performance:

- Experience level (freshman, sophomore, junior, senior)
- Differences in ratings or results between destination team and original team
- The same differences but between destination and original conferences
- Postion (well there are many crazy values here)
- Whether they are good or bad within the current team

Other control variables that we might include:

- NIL (the one we all know but currently do not have sufficient data)

```{r}
# Note that the effect of `changed_team` has changed dramatically when we added the position into the model

improved_fe_model <- feols(ppg ~ changed_team | id + year,
                           data = df)
improved_fe_model
```

(Currently having a singleton issue associated with the model if we use the student fixed effect `id`)