Some useful resources:

- Paper mentioning the package: [(link)](https://imai.fas.harvard.edu/research/files/matchit.pdf)

- More on ICC and a section that mentions singular model fit, where the between-group variance is estimated as 0: [(link)](https://library.virginia.edu/data/articles/understanding-icc)

- Generate publication-like regression tables in R: [(link)](https://r-statistics.co/Regression-Tables-in-R.html)

- Sample economics paper on how to assemble a mixed effects model: [(link)](https://link.springer.com/article/10.1007/s10834-026-10076-w/tables/4)

```{r}
# Clear the environment and read the prepared dataset
rm(list = ls()) # Clear the environment

library(tidyverse)
library(broom)
library(cbbdata)
library(hoopR)
library(lme4) # For mixed effects models
library(mgcv) # For GAMs in R
library(MatchIt) # For propensity matching
library(modelsummary) # Display the results of regression models
library(tinytable) # More table customizations with modelsummary package
library(easystats) # For more advanced statistical metrics like the ICC

df <- read.csv("player_stats_hoopR.csv")
df
```

```{r}
# Change necessary variable data types for the models to work properly
df <- df |>
  mutate(id = as.factor(id),
         player = as.factor(player),
         pos = as.factor(pos),
         team = as.factor(team),
         conf = as.factor(conf)) |>
  filter(!is.na(changed_team) & !is.na(prev_ppg))
  # We don't use data influenced by the pandemic period
```

More on arguments of the `matchit` command: [(link)](https://cran.r-project.org/web/packages/MatchIt/vignettes/matching-methods.html)

Selecting variables for propensity score matching: [(link)](https://link.springer.com/article/10.1007/s10654-019-00494-6)

Simplest propensity matching as a baseline:

```{r}
match_simplest <- matchit(changed_team ~ prev_ppg,
                          data = df, method = "nearest", distance = "glm",
                          replace = TRUE, ratio = 1)
match_simplest_summary <- summary(match_simplest)

match_simplest_summary
```

```{r}
plot(match_simplest, type = "density")
```

```{r}
# Extract the largest standardized mean differences in a matchit summary dataset
max_std_mean_diff <- function(matchit_summary) {
  
  # Extract the standardized mean differences of covariates among all data and matched data
  sum_all_mean_diff <- as.data.frame(match_simplest_summary$sum.all) |>
    select(`Std. Mean Diff.`)
  sum_matched_mean_diff <- as.data.frame(match_simplest_summary$sum.matched) |>
    select(`Std. Mean Diff.`)
  
  # Combine two datasets together and return the maximum value of the column
  result_df <- bind_rows(sum_all_mean_diff, sum_matched_mean_diff)
  result <- max(result_df$`Std. Mean Diff.`)
  
  return(result)
}

max_std_mean_diff(match_simplest_summary)
```

Checking the balance in covariates between matched and unmatched data:

```{r}
plot(match_simplest_summary, var.order = "unmatched",
     xlim = c(0, max_std_mean_diff(match_simplest_summary)))
```

```{r}
# match.data(<matchit object>) represents the dataframe that has been matched
match.data(match_simplest)
```

```{r}
# Use the propensity matching weights in downstream regression
mixed_effects_simplest <- lmer(ppg ~ prev_ppg + pos + exp + year +
                                 current_barthag + (1|id/team/conf),
                           data = match.data(match_simplest), weights = weights)

# https://modelsummary.com/vignettes/modelsummary_extension.html
modelsummary(mixed_effects_simplest, stars = TRUE)
```

```{r, eval = FALSE}
match_object <- matchit(changed_team ~ id + player + pos + exp + team + conf +
                          prev_ppg + current_barthag,
                        data = df, method = "nearest", distance = "glm",
                        replace = TRUE, ratio = 1)
saveRDS(match_object, file = "match_object.rds")
```

```{r, eval = FALSE}
# match_object <- readRDS("match_object.rds")
```

Coming soon: May consider inverse probability score weighting (not omitting so many observations).