This is a supplementary file where I compared between the data obtained from `cbbdata` package and the `hoopR` package. Both of them extracts data from the Bart Torvik API, but `hoopR` is more updated and is currently maintained compared to the other package. There are actually naming differences that impact some crucial analysis results.

```{r}
# Clear the environment and read the prepared dataset
rm(list = ls()) # Clear the environment

library(tidyverse)
library(broom)
library(cbbdata)
library(hoopR)
library(lme4) # For mixed effects models
library(gtsummary) # Display the results of mixed effects models (tidy doesn't work)
library(mgcv) # For GAMs in R

df_cbb <- read.csv("player_stats.csv")
df_hoopR <- read.csv("player_stats_hoopR.csv")
```

```{r}
df_cbb
df_hoopR
```

Sort and standardize the two datasets:

```{r}
df_cbb2 <- df_cbb |>
  select(id, player, year, pos, exp, team, conf, ppg) |>
  arrange(id, year) |>
  # Adjust for naming differences between two datasets
  mutate(team = case_when(
    team ==  "College of Charleston" ~ "Charleston",
    team == "LIU Brooklyn" ~ "LIU",
    team == "North Carolina St." ~ "North Carolina",
    team == "Fort Wayne" ~ "Purdue Fort Wayne",
    team == "Texas A&M Commerce" ~ "East Texas A&M",
    team == "Detroit" ~ "Detroit Mercy",
    team == "St. Francis PA" ~ "Saint Francis",
    team == "Louisiana Lafayette" ~ "Louisiana",
    team == "IUPUI" ~ "IU Indy",
    TRUE ~ team
  ))

df_cbb2
```

```{r}
df_hoopR2 <- df_hoopR |>
  select(id, player, year, pos, exp, team, conf, ppg) |>
  arrange(id, year) |>
  filter(year != 2026)

df_hoopR2
```

```{r}
setdiff(df_cbb2, df_hoopR2) |>
  select(id, player, year, pos, exp, team) |>
  arrange(id, year)
```

```{r}
setdiff(df_hoopR2, df_cbb2) |>
  select(id, player, year, pos, exp, team) |>
  arrange(id, year)
```

```{r}
df_hoopR2 |>
  filter(team == "N.C. State" | team == "North Carolina")
```

Note that I might have made an error when filtering for the `cbbdata` dataset: I confused University of North Carolina and North Carolina State University.