Note that the `player_stats.csv` dataset is created when we run all code chunks, in their specified order, of the `cp1_raw_data_prep.qmd` script.

```{r}
# Clear the environment and read the prepared dataset
rm(list = ls()) # Clear the environment
library(tidyverse)
library(cbbdata)

df <- read.csv("player_stats.csv")
df
```

```{r}
# Redshirted players who were listed as freshman twice
temp <- df |>
  
  select(id, exp, has_transferred, changed_team, cumulative_career_transfers, ppg) |>
    mutate(first_transfer_made = ifelse(cumulative_career_transfers > 0, 1, 0)) |>
  
  filter(has_transferred == 1 & changed_team == 0
         & first_transfer_made == 1 & exp == "So")

temp
```

The players who are "double freshmen" have the IDs: 76132, 76580, 76910.

Some players who are "double sophomores" include the following player IDs: 77600, 77535, 77390.

```{r}
temp2 <- df |>
  filter(id %in% temp$id) |>
  mutate(first_transfer_made = ifelse(cumulative_career_transfers > 0, 1, 0)) |>
  select(id, player, exp, year, team, has_transferred, first_transfer_made,
         changed_team, cumulative_career_transfers)
```


Transfer age curve:

```{r}
age_curves_dataset <- df |>
  
  # Add another column of whether player has made their first transfer in their career (time-sensitive)
  select(exp, has_transferred, changed_team, cumulative_career_transfers, ppg) |>
  mutate(first_transfer_made = ifelse(cumulative_career_transfers > 0, 1, 0)) |>
  select(-cumulative_career_transfers) |>
  
  # Get the mean points per game of players of different types
  group_by(exp, has_transferred, changed_team, first_transfer_made) |>
  summarize(mean_ppg = mean(ppg, na.rm = TRUE)) |>
  
  # Group players based on their transfer history
  ungroup() |>
  mutate(transfer_type = case_when(
    has_transferred == 0 & changed_team == 0 ~ "Never transferred",
    has_transferred == 1 & changed_team == 1 ~ "Year immediately after transfer",
    has_transferred == 1 & changed_team == 0 & first_transfer_made == 0 ~ "Has not transferred yet",
    has_transferred == 1 & changed_team == 0 & first_transfer_made == 1 ~ "2+ years after last transfer*"
    ),
    exp = factor(exp, levels = c("Fr", "So", "Jr", "Sr")))

age_curves_dataset
```


```{r}
# https://www.geeksforgeeks.org/r-language/how-to-fold-legend-into-two-rows-in-ggplot2-in-r/
# https://stackoverflow.com/questions/8454243/move-axis-labels-ggplot
# https://stackoverflow.com/questions/20407773/increase-legend-font-size-ggplot2

# Set the theme of the plot
theme_set(theme_bw())

age_curves_dataset |>
  # The base plot
  ggplot(aes(x = exp, y = mean_ppg, color = transfer_type)) +
  geom_point() + geom_line(aes(group = transfer_type)) +
  # Set the colors for each line
  scale_color_brewer(palette = "Dark2") +
  #theme(legend.position = "bottom") +
  #guides(color = guide_legend(nrow = 2)) +
  # Refine axis, tick marks, and legend titles
  labs(caption = "Data courtesy of the 'cbbdata' package by Andrew Weatherman (2023).\nSuper-seniors are also counted as 'seniors' in this dataset.",
       x = "Player's year in college",
       y = "Average points per game",
       color = "Type of player") +
  scale_x_discrete(labels = c("Freshman", "Sophomore", "Junior", "Senior")) +
  # Slightly adjust the position of axis titles and captions
  # Increase the size of axis labels and tick marks for readability
  theme(axis.title.x = element_text(size = 12, vjust = -1.5, face = "bold"),
        axis.title.y = element_text(size = 12, vjust = 3.5, face = "bold"),
        legend.title = element_text(size = 12, hjust = 0.5, face = "bold"),
        plot.caption = element_text(size = 9, vjust = -2.5),
        axis.text = element_text(size = 11),
        legend.text = element_text(size = 12))

ggsave("cp1_age_curves.png")
```

Additional issues to consider: the 4-2-4 transfer pathways (temporarily transfer to a junior college) and redshirting in determining whether a player has transferred.