Some useful resources that I happened to find:

- Saving data in R (eg: variable values, fitted models, nicely formatted tables, etc.) to a file in your local computer that you can access later: <https://www.sthda.com/english/wiki/saving-data-into-r-data-format-rds-and-rdata>

- Crossed and nested random effects (with sports examples): <https://optimumsportsperformance.com/blog/crossed-vs-nested-random-effects>

For my methods and results section, I will load the data as RData files instead of copying and pasting all the code to the final report. This also helps the report become much cleaner.

```{r}
# Clear the environment and read the prepared dataset
rm(list = ls()) # Clear the environment

library(tidyverse)
library(lme4) # For mixed effects models
library(modelsummary) # Display the results of regression models like in publications
library(MatchIt) # For propensity score matching models

#library(performance)
#library(broom)

df <- read.csv("player_stats_hoopR.csv")
df
```

```{r}
# Change necessary variable data types for the models to work properly
df <- df |>
  mutate(id = as.factor(id),
         player = as.factor(player),
         pos = as.factor(pos),
         team = as.factor(team),
         conf = as.factor(conf)) |>
  
  # Filter out players who have changed team and skipped the year before
  # (they are likely to transfer out of DI to lower divisions or JUCOs)
  filter(!(skipped_year_before == 1 & changed_team == 1)) |>
  
  # Filter out for NA values in the dataset
  drop_na(changed_team, prev_ppg, pos, exp, year, current_barthag, current_barthag_conf,
          id, team, conf)

  # We don't use data influenced by the pandemic period

df
```

Simple linear regression vs. multiple linear regression without team and player fixed effects vs. crossed and nested random effects and partially crossed and nested random effects:

```{r}

# Simple linear regression (age curve and transfer effects)
slr_1 <- lm(ppg ~ prev_ppg, data = df)
slr_2 <- lm(ppg ~ changed_team, data = df)

# Multiple linear regression without player and team effects
linreg <- lm(ppg ~ changed_team + prev_ppg + pos + exp + year + conf +
               current_barthag + current_barthag_conf,
             data = df)

# I wanted to add team and player fixed effects, but this might take ages to run!
#linreg_2 <- feols(ppg ~ changed_team + prev_ppg + pos + exp + year + conf +
                    #current_barthag + current_barthag_conf + id + team,
                  #data = df)
# https://stackoverflow.com/questions/8389843/double-clustered-standard-errors-for-panel-data

# Mixed effects: Crossed, nested, and partially crossed and nested (additions to linreg)
mixed_effects_1 <- lmer(ppg ~ changed_team + prev_ppg + pos + exp + year + conf +
                          current_barthag + current_barthag_conf +
                          (1|team) + (1|id),
                        data = df)

mixed_effects_2 <- lmer(ppg ~ changed_team + prev_ppg + pos + exp + year + conf +
                          current_barthag + current_barthag_conf + (1|team/id),
                        data = df)

mixed_effects_3 <- lmer(ppg ~ changed_team + prev_ppg + pos + exp + year + conf +
                          current_barthag + current_barthag_conf + (1|team:id) +
                          (1|team) + (1|id),
                        data = df)
```

```{r}
# Helper function: Goodness-of-fit metric names of a fixed effects regression model
# and mixed effects model
get_gof(linreg)
get_gof(mixed_effects_3)
get_estimates(mixed_effects_3)
```

```{r}
models_fe <- list(
  "(1)" = slr_1,
  "(2)" = slr_2,
  "(3)" = linreg
)

# Adjust the coefficients and goodness-of-fit display results for readability

models_fe_coef_map <- c(
  "changed_team" = "Transferred",
  "prev_ppg" = "Points per game in previous season",
  "exp" = "Experience level"
  #"posforward" = "Position is Forward",
  #"posguard" = "Position is Guard",
  #"year" = "Season"
  #"current_barthag" = "Team rating",
  #"current_barthag_conf" = "Conference rating"
)

models_fe_gof_map <- tibble(
  raw = c("nobs", "adj.r.squared"),
  clean = c("Number of observations", "Adjusted R<sup>2</sup>"),
  fmt = c("%.0f", "%.3f")
)

models_fe_summary <- modelsummary(models_fe, stars = TRUE,
                                  coef_map = models_fe_coef_map,
                                  gof_map = models_fe_gof_map)
models_fe_summary
```

```{r}

models_mixed <- list(
  "(1)" = mixed_effects_1,
  "(2)" = mixed_effects_2,
  "(3)" = mixed_effects_3
)

models_mixed_coef_map <- c(
  "changed_team" = "Transferred",
  "prev_ppg" = "Points per game in previous season",
  "exp" = "Experience level",
  #"posforward" = "Position is Forward",
  #"posguard" = "Position is Guard",
  #"year" = "Season",
  #"conf" = "Conference",
  #"current_barthag" = "Team rating",
  #"current_barthag_conf" = "Conference rating",
  "SD (Intercept teamid)" = "SD (Team × Player)",
  "SD (Intercept id)" = "SD (Player)",
  "SD (Intercept team)" = "SD (Team)",
  "SD (Observations)" = "SD (Observations)"
)

# Note that conditional R^2 (a.k.a. R^2 that accounts for both fixed and random effects)
# are not available because there are some random effects with zero intraclass correlation

models_mixed_gof_map <- tibble(
  raw = c("nobs", "r2.marginal"),
  clean = c("Number of observations", "Marginal R<sup>2</sup>"),
  fmt = c("%.0f", "%.3f")
)

models_mixed_summary <- modelsummary(models_mixed, stars = TRUE,
                                     coef_map = models_mixed_coef_map,
                                     gof_map = models_mixed_gof_map)
models_mixed_summary
```

```{r}
# Save the nicely-formatted summary tables to the local computer
# Note that you need to create a folder in your working directory called "saved"
saveRDS(models_fe_summary, "saved/models_fe_summary.rds")
saveRDS(models_mixed_summary, "saved/models_mixed_summary.rds")
```

Performing propensity score matching and fitting multiple models together:

```{r}
# Propensity score matching in normal conditions
match_object <- matchit(changed_team ~ pos + exp + current_barthag + current_barthag_conf,
                        data = df, method = "nearest", distance = "glm",
                        replace = FALSE, ratio = 2)

match_object_data <- match.data(match_object)

mixed_effects_matched <- lmer(ppg ~ changed_team + prev_ppg + pos + exp + year + conf +
                                current_barthag + current_barthag_conf + (1|team:id),
                              data = match_object_data, weights = weights)
```

```{r}
# Changed the definition of transferred variable: only including underperformer transfers
df_underperf <- df |>
  mutate(underperf_transfer = changed_team * is_prev_underperformer) |>
  select(-changed_team) |>
  rename(changed_team = underperf_transfer)

# Propensity score matching with underperformer transfers
match_object_underperf <- matchit(changed_team ~ pos + exp + current_barthag +
                                    current_barthag_conf,
                                  data = df_underperf, method = "nearest", distance = "glm",
                                  replace = FALSE, ratio = 2)

match_object_underperf_data <- match.data(match_object_underperf)

mixed_effects_underperf <- lmer(ppg ~ changed_team + prev_ppg + pos + exp + year + conf +
                                  current_barthag + current_barthag_conf + (1|team:id),
                                data = match_object_underperf_data, weights = weights)
```

```{r}
# Changed the definition of transferred variable: only including overperformer transfers
df_overperf <- df |>
  mutate(overperf_transfer = changed_team * is_prev_overperformer) |>
  select(-changed_team) |>
  rename(changed_team = overperf_transfer)

# Propensity score matching with overperformer transfers
match_object_overperf <- matchit(changed_team ~ pos + exp + current_barthag +
                                    current_barthag_conf,
                                  data = df_overperf, method = "nearest", distance = "glm",
                                  replace = FALSE, ratio = 2)

match_object_overperf_data <- match.data(match_object_overperf)

mixed_effects_overperf <- lmer(ppg ~ changed_team + prev_ppg + pos + exp + year + conf +
                                  current_barthag + current_barthag_conf + (1|team:id),
                                data = match_object_overperf_data, weights = weights)
```

```{r}
# Visualize these three models in a table
models_matching <- list(
  "(1)" = mixed_effects_matched,
  "(2)" = mixed_effects_underperf,
  "(3)" = mixed_effects_overperf
)

models_matching_coef_map <- c(
  "changed_team" = "Transferred",
  "prev_ppg" = "Points per game in previous season",
  "exp" = "Experience level",
  #"posforward" = "Position is Forward",
  #"posguard" = "Position is Guard",
  #"year" = "Season",
  #"conf" = "Conference",
  #"current_barthag" = "Team rating",
  #"current_barthag_conf" = "Conference rating",
  "SD (Intercept teamid)" = "SD (Team × Player)",
  "SD (Observations)" = "SD (Observations)"
)

# Note that conditional R^2 (a.k.a. R^2 that accounts for both fixed and random effects)
# are not available because there are some random effects with zero intraclass correlation

models_matching_gof_map <- tibble(
  raw = c("nobs", "r2.marginal", "r2.conditional"),
  clean = c("Number of observations", "Marginal R<sup>2</sup>", "Conditional R<sup>2</sup>"),
  fmt = c("%.0f", "%.3f", "%.3f")
)

models_matching_summary <- modelsummary(models_matching, stars = TRUE,
                                        coef_map = models_matching_coef_map,
                                        gof_map = models_matching_gof_map)
models_matching_summary
```

Plotting a graph of model point estimates and their confidence intervals (95%):

```{r}
# Fixed effects with SLR and MLR, Random effects without & with matching

# Helper function: Extract the point estimates of a named variable and their confidence interval
# as a dataframe for plotting

compare_causal_models <- function(var_name, model_list) {
  # var_name: Variable name of interest (need to be the same across all models).
  # var_name must be in quotes
  
  # model_list: List of models (not named) to compare the effects of a variable.
  # model_list should contain an user-specified name and the model variable itself.
  
  # Create a blank dataframe to store the results
  result <- tibble(
    model_name = character(),
    estimate = numeric(),
    std.error = numeric(),
    statistic = numeric(),
    p.value = numeric(),
    conf.low = numeric(),
    conf.high = numeric(),
  )
  
  # Extract the estimates and append to the result dataframe
  for (model_name in names(model_list)) {
    
    # Extract the model's name as a value:
    # https://stackoverflow.com/questions/24309910/how-to-get-name-of-variable-in-r-substitute
    model <- model_list[[model_name]]
    
    model_name_df <- tibble(
      model_name = c(model_name)
    )
    
    model_df <- get_estimates(model) |>
      filter(term == var_name) |>
      select(estimate, std.error, statistic, p.value, conf.low, conf.high)
    
    # Combine the model names with model statistics together, then append the result to the
    # returning dataframe
    model_df <- bind_cols(model_name_df, model_df)
    result <- bind_rows(result, model_df)
  }
  
  # Return the resulting dataframe
  return(result)
}

my_model_list <- list("Simple Linear Regression\n(unadjusted standard errors)" = slr_2,
                      "Multiple Linear Regression" = linreg,
                      "Mixed Effects Without Matching" = mixed_effects_3,
                      "Mixed Effects With Matching" = mixed_effects_matched)

compare_causal_models_df <- compare_causal_models("changed_team", my_model_list)
compare_causal_models_df
```

```{r}
# Set the themes for each plot
theme_set(theme_bw())

compare_causal_plot <- compare_causal_models_df |>
  # Generate a blank canvas and add a zero baseline for reference
  ggplot() +
  geom_vline(xintercept = 0, color = "black", linetype = "dashed",
             linewidth = 0.7, alpha = 0.5) +
  
  # Supply the plot with data
  geom_point(aes(x = estimate, y = model_name)) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high, y = model_name,
                     width = 0.3)) +
  
  # Modify axis labels for readability
  labs(x = "Coefficient value of the Transferred variable with confidence interval",
       y = "") +
  
  theme(
    # Remove unnecessary vertical gridlines from the graph
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  )

compare_causal_plot
```


```{r}
# Model estimates in general, overperformer, and underperformer
```


Plotting balance diagnostics of the matched models (Love plots):

```{r}
# Helper function to assist with the Love plots:

# Extract the largest standardized mean differences in a matchit summary dataset
max_std_mean_diff <- function(matchit_summary) {
  
  # Extract the standardized mean differences of covariates among all data and matched data
  sum_all_mean_diff <- as.data.frame(matchit_summary$sum.all) |>
    select(`Std. Mean Diff.`)
  sum_matched_mean_diff <- as.data.frame(matchit_summary$sum.matched) |>
    select(`Std. Mean Diff.`)
  
  # Combine two datasets together and return the maximum value of the column
  result_df <- bind_rows(sum_all_mean_diff, sum_matched_mean_diff)
  result <- max(result_df$`Std. Mean Diff.`)
  
  return(result)
}
```

```{r}
library(cobalt) # Generate a ggplot representation of the Love plot

# drop.distance = TRUE: Drop the distance metric when visualizing the Love plot

match_object_love_plot <- love.plot(match_object, stat = "mean.diffs",
                                    var.order = "unadjusted", threshold = 0.1,
                                    drop.distance = TRUE, binary = "std") +
  
  # Edit the labels and colors to be colorblind-friendly
  labs(title = "",
       color = "Group") +
  scale_y_discrete(labels = c("exp" = "Experience level",
                              "current_barthag_conf" = "Conference rating",
                              "current_barthag" = "Team rating",
                              "pos_center" = "Position is Center",
                              "pos_guard" = "Position is Guard",
                              "pos_forward" = "Position is Forward")) +
  scale_color_manual(values = c("#fc8d62", "#66c2a5"),
                     labels = c("All data", "Matched data"))

match_object_love_plot
```

```{r}
# Save the propensity score matching model results as a checkpoint
saveRDS(models_matching_summary, "saved/models_matching_summary.rds")
saveRDS(compare_causal_plot, "saved/compare_causal_plot.rds")
saveRDS(match_object_love_plot, "saved/match_object_love_plot.rds")
```