#### 0. Data preprocessing

```{r}
# Clear the environment and read the prepared dataset
rm(list = ls()) # Clear the environment

library(tidyverse) # For data wrangling
library(lme4) # For mixed effects models
library(modelsummary) # Display the results of regression models like in publications
library(MatchIt) # For propensity score matching models
library(broom) # For more information on each regression model (not mixed effects)
library(broom.mixed) # Special broom library for mixed effects models

df <- read.csv("player_stats_hoopR.csv")
df
```

```{r}
# Change necessary variable data types for the models to work properly
df <- df |>
  mutate(id = as.factor(id),
         player = as.factor(player),
         pos = as.factor(pos),
         team = as.factor(team),
         conf = as.factor(conf)) |>
  
  # Filter out players who have changed team and skipped the year before
  # (they are likely to transfer out of DI to lower divisions or JUCOs)
  filter(!(skipped_year_before == 1 & changed_team == 1)) |>
  
  # Filter out for NA values in the dataset
  drop_na(changed_team, prev_ppg, pos, exp, year, current_barthag, current_barthag_conf,
          id, team, conf)

  # We don't use data influenced by the pandemic period

df
```

Generate separate datasets for overperformer and underperformer transfers (changing the definition of the variable `changed_team`):

```{r}
# Changed the definition of transferred variable: only including underperformer transfers
df_underperf <- df |>
  mutate(underperf_transfer = changed_team * is_prev_underperformer) |>
  select(-changed_team) |>
  rename(changed_team = underperf_transfer)
```

```{r}
# Changed the definition of transferred variable: only including overperformer transfers
df_overperf <- df |>
  mutate(overperf_transfer = changed_team * is_prev_overperformer) |>
  select(-changed_team) |>
  rename(changed_team = overperf_transfer)
```

Baseline multiple linear regression model and mixed effects model (note that this chunk will not run automatically):

```{r, eval = FALSE}
# Note that we do not include team and player fixed effects because of the singleton issue
# and high runtime associated with creating thousands of binary variables

# Normal conditions:
linreg_normal <- lm(ppg ~ changed_team + prev_ppg + pos + exp + year + conf +
               current_barthag + current_barthag_conf,
             data = df)

no_matching_normal <- lmer(ppg ~ changed_team + prev_ppg + pos + exp + year + conf +
                             current_barthag + current_barthag_conf + (1|team:id),
                           data = df)

# Underperformers dataset:
linreg_underperf <- lm(ppg ~ changed_team + prev_ppg + pos + exp + year + conf +
                         current_barthag + current_barthag_conf,
                       data = df_underperf)

no_matching_underperf <- lmer(ppg ~ changed_team + prev_ppg + pos + exp + year + conf +
                             current_barthag + current_barthag_conf + (1|team:id),
                           data = df_underperf)

# Overperformers dataset:
linreg_overperf <- lm(ppg ~ changed_team + prev_ppg + pos + exp + year + conf +
                        current_barthag + current_barthag_conf,
                       data = df_overperf)

no_matching_overperf <- lmer(ppg ~ changed_team + prev_ppg + pos + exp + year + conf +
                             current_barthag + current_barthag_conf + (1|team:id),
                           data = df_overperf)
```

#### 1. Initial modeling

Propensity score matching:

```{r}
# Functions to automate the propensity score matching workflow without copying and pasting code

propensity_score_matching <- function(my_data, match_ratio = 2) {
  
  # my_data: The dataset in question

  # match_ratio: The ratio to match between treatment and control.
  # This depends on the ratio between transfers and non-transfers for each group.
  
  # Function returns the match_object, match_object_data, and mixed_effects_matched for the
  # dataset. We can consider the result as a "propensity_score_matching" object
  
  match_object <- matchit(changed_team ~ pos + exp + current_barthag + current_barthag_conf,
                          data = my_data, method = "nearest", distance = "glm",
                          link = "linear.logit", caliper = 0.2, std.caliper = TRUE,
                          replace = FALSE, ratio = match_ratio)

  match_object_data <- match.data(match_object)

  mixed_effects_matched <- lmer(ppg ~ changed_team + prev_ppg + pos + exp + year + conf +
                                  current_barthag + current_barthag_conf + (1|team:id),
                                data = match_object_data)
  
  results <- list("match_object" = match_object, "match_object_data" = match_object_data,
                  "mixed_effects_matched" = mixed_effects_matched)
  
  return(results)
  
}
```

```{r}
# Implement the function with normal conditions, underperformers, and overperformers
matching_normal <- propensity_score_matching(df, match_ratio = 2)
matching_underperf <- propensity_score_matching(df_underperf, match_ratio = 7)
matching_overperf <- propensity_score_matching(df_overperf, match_ratio = 4)
```

```{r}
# Obtain individual objects of each propensity score matching specification

match_object_normal <- matching_normal$match_object
match_object_underperf <- matching_underperf$match_object
match_object_overperf <- matching_overperf$match_object

match_object_data_normal <- matching_normal$match_object_data
match_object_data_underperf <- matching_underperf$match_object_data
match_object_data_overperf <- matching_overperf$match_object_data

mixed_effects_normal <- matching_normal$mixed_effects_matched
mixed_effects_underperf <- matching_underperf$mixed_effects_matched
mixed_effects_overperf <- matching_overperf$mixed_effects_matched
```

#### 2. Mixed effects model diagnostics

Note that because our sample size is very large (sometimes up to over 10000 observations!), the normality assumption does not matter so much.

```{r}
# Function to automate the plotting of a mixed effects regression model's diagnostics

# 1. Plotting fitted value versus actual value
fitted_vs_actual_mixed_effects <- function(model, data) {

  # Retrieve the data to plot model diagnostics plot
  augmented_dataset <- model |>
    augment(data)
  
  # Return the plot itself
  result_plot <- augmented_dataset |>
    ggplot(aes(x = .fitted, y = ppg)) +
    geom_point(alpha = 0.2) +
    geom_abline(slope = 1, intercept = 0, color = "red",
                linetype = "dashed", linewidth = 1.5)
  
  return(result_plot)
  
}
```

```{r}
fitted_vs_actual_mixed_effects(mixed_effects_normal, match_object_data_normal)
fitted_vs_actual_mixed_effects(mixed_effects_underperf, match_object_data_underperf)
fitted_vs_actual_mixed_effects(mixed_effects_overperf, match_object_data_overperf)
```

```{r}
# Function to automate the plotting of a mixed effects regression model's diagnostics

# 2. Plotting fitted value versus residuals
fitted_vs_resid_mixed_effects <- function(model, data) {

  # Retrieve the data to plot model diagnostics plot
  augmented_dataset <- model |>
    augment(data)
  
  # Return the plot itself
  result_plot <- augmented_dataset |>
    ggplot(aes(x = .fitted, y = .resid)) +
    geom_point(alpha = 0.2) +
    geom_hline(yintercept = 0, color = "red", linetype = "dashed", linewidth = 1.5) +
    geom_smooth()
  
  return(result_plot)
  
}
```

```{r}
fitted_vs_resid_mixed_effects(mixed_effects_normal, match_object_data_normal)
fitted_vs_resid_mixed_effects(mixed_effects_underperf, match_object_data_underperf)
fitted_vs_resid_mixed_effects(mixed_effects_overperf, match_object_data_overperf)
```

#### 3. Propensity score matching model diagnostics

```{r}
# Load additional libraries for logistic regression diagnostics
library(regressinator)
library(DHARMa)
```

The diagnostics plots are directly used from the CMSACamp lecture. For normal conditions first: Note that the empirical link plot shows a roughly linear correlation when we considered the data as a whole, but not so linear when we only considered the matched data.

```{r}
# Helper function to create empirical link plots (specified equation)
empirical_link_plot <- function(dataset, varname) {
  
  # Note that varname (variable of interest) should come with quotes

  # Fit a logit model on the data
  logit_matched_normal <- glm(changed_team ~ pos + exp + current_barthag +
                                current_barthag_conf,
                              data = dataset,
                              family = binomial)
  
  # Create data for the empirical link plot (aggregate by quantiles)
  plot_data_by_variable <- dataset |>
    bin_by_quantile({{ varname }}, breaks = 20) |>
    summarize(
      mean_varname = mean({{ varname }}),
      prob = mean(changed_team),
      log_odds = empirical_link(changed_team, family = binomial(link = "logit")),
      n = n()
    )
  
  plot_data_by_variable |>
    ggplot(aes(x = mean_varname, y = log_odds)) +
    geom_point(aes(size = n)) +
    geom_smooth()

}
```

```{r}
empirical_link_plot(df, current_barthag)
empirical_link_plot(df, current_barthag_conf)
empirical_link_plot(df_underperf, current_barthag)
empirical_link_plot(df_underperf, current_barthag_conf)
empirical_link_plot(df_overperf, current_barthag)
empirical_link_plot(df_overperf, current_barthag_conf)
```

Randomized quantile residuals:

```{r}
# Helper function to create randomized quantile residuals for a dataset (specified equation)

randomized_quantile_residuals <- function(dataset) {
  
  # Setting up a logistic regression model with the data
  logit_model <- glm(changed_team ~ pos + exp + current_barthag + current_barthag_conf,
                     data = dataset, family = binomial)
  
  # Simulate residuals for binary variables and add them to the dataset
  dh_object <- simulateResiduals(logit_model)
  
  augmented_df <- augment(logit_model) |>
    mutate(.quantile.resid = residuals(dh_object))
  
  # Plot the fitted value versus quantile residuals
  augmented_df |>
    ggplot(aes(x = .fitted, y = .quantile.resid)) +
    geom_point(alpha = 0.2) +
    geom_hline(yintercept = 0.5, color = "red",
               linetype = "dashed", linewidth = 0.8) +
    geom_smooth()
}
```

```{r}
randomized_quantile_residuals(df)
randomized_quantile_residuals(df_underperf)
randomized_quantile_residuals(df_overperf)
```
