Some useful resources that I happened to find:

- Saving data in R (eg: variable values, fitted models, nicely formatted tables, etc.) to a file in your local computer that you can access later: <https://www.sthda.com/english/wiki/saving-data-into-r-data-format-rds-and-rdata>

- Crossed and nested random effects (with sports examples): <https://optimumsportsperformance.com/blog/crossed-vs-nested-random-effects>

For my methods and results section, I will load the data as RData files instead of copying and pasting all the code to the final report. This also helps the report become much cleaner.

#### 1. Data preprocessing for modeling

```{r}
# Clear the environment and read the prepared dataset
rm(list = ls()) # Clear the environment

library(tidyverse)
library(lme4) # For mixed effects models
library(modelsummary) # Display the results of regression models like in publications
library(MatchIt) # For propensity score matching models

#library(performance)
#library(broom)

df <- read.csv("player_stats_hoopR.csv")
df
```

```{r}
# Change necessary variable data types for the models to work properly
df <- df |>
  mutate(id = as.factor(id),
         player = as.factor(player),
         pos = as.factor(pos),
         team = as.factor(team),
         conf = as.factor(conf)) |>
  
  # Filter out players who have changed team and skipped the year before
  # (they are likely to transfer out of DI to lower divisions or JUCOs)
  filter(!(skipped_year_before == 1 & changed_team == 1)) |>
  
  # Filter out for NA values in the dataset
  drop_na(changed_team, prev_ppg, pos, exp, year, current_barthag, current_barthag_conf,
          id, team, conf)

  # We don't use data influenced by the pandemic period

df
```

Simple linear regression vs. multiple linear regression without team and player fixed effects vs. crossed and nested random effects and partially crossed and nested random effects:

```{r}

# Simple linear regression (age curve and transfer effects)
slr_1 <- lm(ppg ~ prev_ppg, data = df)
slr_2 <- lm(ppg ~ changed_team, data = df)

# Multiple linear regression without player and team effects
linreg <- lm(ppg ~ changed_team + prev_ppg + pos + exp + year + conf +
               current_barthag + current_barthag_conf,
             data = df)

# I wanted to add team and player fixed effects, but this might take ages to run!
#linreg_2 <- feols(ppg ~ changed_team + prev_ppg + pos + exp + year + conf +
                    #current_barthag + current_barthag_conf + id + team,
                  #data = df)
# https://stackoverflow.com/questions/8389843/double-clustered-standard-errors-for-panel-data

# Mixed effects: Crossed, nested, and partially crossed and nested (additions to linreg)
mixed_effects_1 <- lmer(ppg ~ changed_team + prev_ppg + pos + exp + year + conf +
                          current_barthag + current_barthag_conf +
                          (1|team) + (1|id),
                        data = df)

mixed_effects_2 <- lmer(ppg ~ changed_team + prev_ppg + pos + exp + year + conf +
                          current_barthag + current_barthag_conf + (1|team/id),
                        data = df)

mixed_effects_3 <- lmer(ppg ~ changed_team + prev_ppg + pos + exp + year + conf +
                          current_barthag + current_barthag_conf + (1|team:id) +
                          (1|team) + (1|id),
                        data = df)
```

```{r}
# Helper function: Goodness-of-fit metric names of a fixed effects regression model
# and mixed effects model
get_gof(linreg)
get_gof(mixed_effects_3)
get_estimates(mixed_effects_3)
```

```{r}
models_fe <- list(
  "(1)" = slr_1,
  "(2)" = slr_2,
  "(3)" = linreg
)

# Adjust the coefficients and goodness-of-fit display results for readability

models_fe_coef_map <- c(
  "changed_team" = "Transferred",
  "prev_ppg" = "Points per game in previous season",
  "exp" = "Experience level"
  #"posforward" = "Position is Forward",
  #"posguard" = "Position is Guard",
  #"year" = "Season"
  #"current_barthag" = "Team rating",
  #"current_barthag_conf" = "Conference rating"
)

models_fe_gof_map <- tibble(
  raw = c("nobs", "adj.r.squared"),
  clean = c("Number of observations", "Adjusted R<sup>2</sup>"),
  fmt = c("%.0f", "%.3f")
)

models_fe_summary <- modelsummary(models_fe, stars = TRUE,
                                  coef_map = models_fe_coef_map,
                                  gof_map = models_fe_gof_map)
models_fe_summary
```

```{r}

models_mixed <- list(
  "(1)" = mixed_effects_1,
  "(2)" = mixed_effects_2,
  "(3)" = mixed_effects_3
)

models_mixed_coef_map <- c(
  "changed_team" = "Transferred",
  "prev_ppg" = "Points per game in previous season",
  "exp" = "Experience level",
  #"posforward" = "Position is Forward",
  #"posguard" = "Position is Guard",
  #"year" = "Season",
  #"conf" = "Conference",
  #"current_barthag" = "Team rating",
  #"current_barthag_conf" = "Conference rating",
  "SD (Intercept teamid)" = "SD (Team × Player)",
  "SD (Intercept id)" = "SD (Player)",
  "SD (Intercept team)" = "SD (Team)",
  "SD (Observations)" = "SD (Observations)"
)

# Note that conditional R^2 (a.k.a. R^2 that accounts for both fixed and random effects)
# are not available because there are some random effects with zero intraclass correlation

models_mixed_gof_map <- tibble(
  raw = c("nobs", "r2.marginal"),
  clean = c("Number of observations", "Marginal R<sup>2</sup>"),
  fmt = c("%.0f", "%.3f")
)

models_mixed_summary <- modelsummary(models_mixed, stars = TRUE,
                                     coef_map = models_mixed_coef_map,
                                     gof_map = models_mixed_gof_map)
models_mixed_summary
```

#### Final ANOVA: are the random effects valid to be included in our model?

```{r}
test_lm <- lm(ppg ~ changed_team + prev_ppg + pos + exp + year + conf +
                current_barthag + current_barthag_conf,
              data = df)

test_lmer <- lmer(ppg ~ changed_team + prev_ppg + pos + exp + year + conf +
                    current_barthag + current_barthag_conf + (1|team:id),
                  data = df, REML = FALSE)

anova(test_lmer, test_lm)
```


#### Saving models and plots to make the final report or presentation cleaner

Note that you need to create a folder in your working directory called `saved` for the code to be reproducible.

```{r}
# Save the nicely-formatted summary tables to the local computer
saveRDS(models_fe_summary, "saved/models_fe_summary.rds")
saveRDS(models_mixed_summary, "saved/models_mixed_summary.rds")
```