```{r}
rm(list = ls()) # Clear the environment
library(tidyverse)
library(cbbdata)
library(hoopR)
```

```{r}
# Read the player stats by season dataset
df_2022 <- torvik_player_stats(year = 2022)
df_2023 <- torvik_player_stats(year = 2023)
df_2024 <- torvik_player_stats(year = 2024)
df_2025 <- torvik_player_stats(year = 2025)
df_2026 <- torvik_player_stats(year = 2026)

df <- bind_rows(df_2022, df_2023, df_2024, df_2025, df_2026)
df
```

```{r}
# Select the necessary columns and rename it to fit with the cbbdata package
df <- df |>
  select(id = player_id, player = player_name, year, pos = role, exp = class,
         team, conf, ppg = pts, min_pct)

df
```

#### 1. Recoding values

In this data preparation, we will do the following:

- Update the conference classification and previous conference to the dataset.

- For simplicity in models, we will recode the experience level of each college athletes as a numeric value from 1 to 4, where 1 corresponds to Freshman (`Fr`) and 4 corresponds to Senior (`Sr`). Later in the code, we will add level 5 called Super-senior (`SSr`).

```{r}
df_enhanced <- df |>
  # Move the unique player ID to the front for readability first
  relocate(id) |>
  mutate(
    # Update the conference classification (code by Kendall)
    conf_class = case_when(
      conf %in% c("B10", "SEC", "ACC", "B12", "BE") ~ "high-major",
      conf %in% c("Amer", "A10", "CAA", "MVC", "MWC", "P12", "WCC") ~ "mid-major",
      conf %in% c("AE", "ASun", "BSky", "BSth", "BW", "CUSA", "Horz",
                  "Ivy", "MAAC", "MAC", "MEAC", "NEC", "OVC", "Pat",
                  "SC", "Slnd", "SWAC", "Sum", "SB", "WAC", "ind") ~ "low-major"
    ),
    # Recode the experience level (note that we are wrapping the result as strings
    # so that case_when can work properly)
    exp = case_when(
      exp == "Fr" ~ "1", exp == "So" ~ "2", exp == "Jr" ~ "3",
      exp == "Sr" ~ "4", TRUE ~ exp
      ),
    # Recode the position names for easier analysis
    pos = case_when(
      pos %in% c("Wing F", "Stretch 4", "PF/C") ~ "forward",
      pos == "C" ~ "center",
      pos %in% c("Combo G", "Scoring PG", "Wing G", "Pure PG") ~ "guard"
      )
    ) |>
  # Convert experience level to numeric
  mutate(exp = as.numeric(exp))

df_enhanced
```

#### 2. Indicators for players' transfer history

Kendall helped me with the code for the `has_transferred` and `changed_team` variables!

```{r}
# Exploring Kendall's code (adjusted to group by player IDs instead of player names)

transfer_df <- df_enhanced |>
  group_by(id) |>
  filter(n_distinct(team) > 1) |>
  ungroup()

transfer_df
```

```{r}
# Whether a player has transferred - make it a binary variable for easier coding
df_enhanced <- df_enhanced |>
  mutate(has_transferred = ifelse(id %in% transfer_df$id, 1, 0))

df_enhanced
```

I have refined the code for the `changed_team` variable originally proposed by Kendall to account for players who transferred out of a NCAA Division I institution, or redshirted due to certain factors.

I also created new columns to better identify the transfer patterns for each student-athlete:

- `cumulative_career_transfers`: Cumulative number of college career transfers for each player. Each time a player transfers to a new team, the number increases by 1.

- `num_transfers`: The total number of transfers that a particular player has made throughout their entire college career (some people can have more than one transfers).

```{r}
career_transfers_by_player <- transfer_df |>
  select(id, year, team, exp, conf, conf_class, ppg, min_pct) |>
  arrange(id, year) |>
  group_by(id) |>
  mutate(prev_team = lag(team),
         prev_exp = lag(exp),
         prev_conf = lag(conf),
         prev_conf_class = lag(conf_class),
         prev_ppg = lag(ppg),
         prev_min_pct = lag(min_pct),
         # Player's transfer history
         changed_team = ifelse(!(is.na(prev_team)) & (team != prev_team),
                                1, 0),
         cumulative_career_transfers = cumsum(changed_team),
         num_transfers = max(cumulative_career_transfers)) |>
  
  select(-team, -exp, -conf, -conf_class, -ppg, -min_pct)

career_transfers_by_player
```

Note that the previous points per game takes the points per game from the last season of the player in NCAA Division I.

Now rejoin these columns back into the dataset:

```{r}
df_enhanced <- df_enhanced |>
  left_join(career_transfers_by_player, join_by(id, year))

df_enhanced
```

I have also created another indicator variable, `skipped_year_before`, that denotes players who skipped one or more years in the Division I based on the lack of data. Some possible reasons for this are redshirting or transferring to non-DI institutions.

```{r}
year_presence_d1 <- df_enhanced |>
  select(id, year) |>
  arrange(id, year) |>
  group_by(id) |>
  mutate(
    
    # Comparing the player's previous year in NCAA DI and the year immediately
    # before to understand which player skipped one or more years in NCAA
         prev_d1_year = lag(year),
         year_before = year - 1,
         skipped_year_before = case_when(
           prev_d1_year == year_before ~ 0,
           is.na(prev_d1_year) ~ 0,
           prev_d1_year != year_before ~ 1
         )) |>
  
  # Separate mutate command to specify that we don't know whether a player has
  # skipped NCAA DI the year before if the year before is 2021
  
  mutate(skipped_year_before = case_when(
    year_before == 2021 ~ NA,
    TRUE ~ skipped_year_before
  ))

year_presence_d1
```

Again, join the new columns that I have created above back into the full dataset:

```{r}
df_enhanced <- df_enhanced |>
  left_join(year_presence_d1, join_by(id, year))

df_enhanced
```

We can check the code's accuracy by filtering the current `df_enhanced` dataset for a particular player ID that transfers the most, such as `65627` and `70342`.

#### 3. Experience level (class year) of the college athletes

Filling in the experience level for the remaining players:

- Jason Umosella (La Salle University): Freshman (as of 2024-25) ([link](https://goexplorers.com/sports/mens-basketball/roster/jason-umosella/12302))
- Alvin Stredic (Misissippi Valley State University): Junior (as of 2022-23) ([link](https://mvsusports.com/sports/mens-basketball/roster/alvin-stredic/3290))

```{r}
df_enhanced |>
  filter(!(exp %in% c(1, 2, 3, 4)))
```

```{r}
df_enhanced <- df_enhanced |>
  mutate(exp = case_when(
    player == "Jason Umosella" & year == 2025 ~ 1,
    player == "Alvin Stredic" & year == 2023 ~ 3,
    TRUE ~ exp
  ))

df_enhanced
```

```{r}
# Checking that the data preparation works
df_enhanced |>
  filter(player %in% c("Jason Umosella", "Alvin Stredic"))
```

```{r}
df_enhanced
```

Kendall helped me filter for super-seniors in the dataset, which is also an essential aspect to note related to the loopholes of NCAA transfer rules. The new numeric indicator for super-seniors is denoted as `5`.

Note that we do not want to take the data for players in the 2020-21 season and prior because eligibility is heavily affected by the Covid-19 pandemic. Therefore, even if a player is a super-senior in the 2021-22 season, the dataset will still mark them as senior, or `4`.

I also created another indicator variable, `no_change_exp`, that yields 1 if the player does not change their experience level compared to the previous season they played in NCAA Division I, and 0 otherwise. This variable is expected to effectively spot both super-seniors and double freshmen/ sophomores/ juniors due to redshirting or transferring out of Division I.

```{r}
df_enhanced <- df_enhanced |>
  mutate(no_change_exp = case_when(
    exp == prev_exp ~ 1,
    is.na(prev_exp) ~ 0,
    TRUE ~ 0
  )) |>
  mutate(exp = ifelse(prev_exp == 4 & no_change_exp == 1, 5, exp))

df_enhanced
```

Side note: How many players are registered as double freshmen, sophomores, or juniors? (In the following code block, duplicated players are counted as multiple records.)

```{r}
df_enhanced |>
  select(id, year, exp, prev_exp, no_change_exp) |>
  filter(no_change_exp == 1 & prev_exp != 4) |>
  count(prev_exp)
```

#### 4. Add team ratings

Next, I will factor team ratings into the player statistics. In the future, I might also factor conference ratings in my model.

Since I have had some issues with accessing updated data from the `cbbdata` package (no longer actively maintained as of June 2026), I will use the `hoopR` package to access additional data, including team data.

I currently use Bart Torvik's T-Rank, which fits with the Bart Torvik API used in the `cbbdata` package, particularly in the team names. Using data from the same API source is very convenient in merging data where we do not have to deal with many differences in data entry.

```{r}
# The data is obtained from the hoopR package
# Note that the Torvik ratings for 2025 was unavailable

df_team_ratings_2022 <- torvik_ratings(year = 2022)
df_team_ratings_2023 <- torvik_ratings(year = 2023)
df_team_ratings_2024 <- torvik_ratings(year = 2024)
df_team_ratings_2025 <- torvik_ratings(year = 2025)
df_team_ratings_2026 <- torvik_ratings(year = 2026)

# Concatenate all datasets together and filter for only necessary rows

# Note that some team names in `df_team_ratings` are different from our previous
# dataset, `df_enhanced`; so we need to change their names to merge these two
# datasets better

df_team_ratings_full <- bind_rows(df_team_ratings_2022, df_team_ratings_2023,
                             df_team_ratings_2024, df_team_ratings_2025,
                             df_team_ratings_2026)

df_team_ratings <- df_team_ratings_full |>
  select(year, team, barthag)

df_team_ratings
```

I realized that I need to also account for the fact that not every player is transferring and some new players join the NCAA Division I in each year (like freshmen graduating from high school), so I will fill in certain `NA`s in the dataset for players who have not yet transferred.

```{r}
df_enhanced <- df_enhanced |>
  
  # Add information for players who have not yet transferred and who have just joined the team
  mutate(prev_team = ifelse(year != 2022 & changed_team == 0, team, prev_team)) |>
  mutate(changed_team = ifelse(year != 2022 & is.na(prev_team), 0, changed_team)) |>
  
  # Some teams have not joined Division I until recently
  # Therefore, we also need to replace the prev_team and changed_team as NA because we have
  # no other information
  mutate(
    prev_team = case_when(
      team == "East Texas A&M" & year <= 2023	~ NA,
      team == "Le Moyne" & year <= 2024	~ NA,
      team == "Lindenwood" & year <= 2023	~ NA,
      team == "Mercyhurst" & year <= 2025 ~ NA,
      team == "New Haven" & year <= 2026 ~ NA,
      team == "Queens" & year <= 2023 ~ NA,
      team == "Southern Indiana" & year <= 2023 ~ NA,
      team == "Stonehill" & year <= 2023 ~ NA,
      team == "West Georgia" & year <= 2025 ~ NA,
      TRUE ~ prev_team
      ),
    changed_team = case_when(
      team == "East Texas A&M" & year <= 2023	~ NA,
      team == "Le Moyne" & year <= 2024	~ NA,
      team == "Lindenwood" & year <= 2023	~ NA,
      team == "Mercyhurst" & year <= 2025 ~ NA,
      team == "New Haven" & year <= 2026 ~ NA,
      team == "Queens" & year <= 2023 ~ NA,
      team == "Southern Indiana" & year <= 2023 ~ NA,
      team == "Stonehill" & year <= 2023 ~ NA,
      team == "West Georgia" & year <= 2025 ~ NA,
      TRUE ~ changed_team
      )
  )
```

I will create some more variables in the working dataframe, `df_enhanced`. This applies to all players including those who did not transfer. They should be blank for the 2022 season because we were not so sure about the player's previous team:

- `prev_year`: The previous season for the player, which can be represented as a formula: `prev_year = year - 1`.
- `current_barthag`: The Barthag rating of the current team in the current season.
- `prev_barthag_current`: The Barthag rating of the current team in the previous season (the "destination team" for players who transferred).
- `prev_barthag_prev`: The Barthag rating of the previous team in the previous season (the "departure team" for players who transferred).

I also want to take the differences in the ratings as whether the player transfers from a lower-ranked school to a higher-ranked school and vice versa, and add whether. Thus, I will create a new variable called `team_rating_change`, which takes the difference in team's Barthag rating from previous year to current year.

```{r}
df_enhanced <- df_enhanced |>
  mutate(prev_year = year - 1) |>
  # Add the team rating in the current season
  left_join(df_team_ratings, join_by(year, team)) |>
  rename(current_barthag = barthag) |>
  # Add the team rating of the current team in the previous season
  left_join(df_team_ratings, join_by(prev_year == year, team)) |>
  rename(prev_barthag_current = barthag) |>
  # Add the team rating of the previous team in the previous season
  left_join(df_team_ratings, join_by(prev_year == year, prev_team == team)) |>
  rename(prev_barthag_prev = barthag) |>
  
  # Create the team rating change of the player from this year to last year,
  # and create indicators that signal moving from small to big teams and vice versa
  
  # The threshold of 0.1 for better/worse team is actually suggested by AI (eg: 0.6 vs 0.5)
  # I need someone with more domain knowledge here
  mutate(team_rating_change = current_barthag - prev_barthag_prev,
         to_better_team = ifelse(
           changed_team == 1 & (!is.na(team_rating_change)) & team_rating_change >= 0.1,
           1, 0
         ),
         to_worse_team = ifelse(
           changed_team == 1 & (!is.na(team_rating_change)) & team_rating_change <= -0.1,
           1, 0
         ),
         to_same_tier_team = ifelse(
           changed_team == 1 & (!is.na(team_rating_change)) &
             team_rating_change > -0.1 & team_rating_change < 0.1,
           1, 0
         )) |>
  
  # Note that if changed_team is currently NA (not sufficient information),
  # each transfer indicators should also return NA
  mutate(to_better_team = ifelse(is.na(changed_team), NA, to_better_team),
         to_worse_team = ifelse(is.na(changed_team), NA, to_worse_team),
         to_same_tier_team = ifelse(is.na(changed_team), NA, to_same_tier_team))

df_enhanced
```

We will check for null values in the team ranks as an indicator of whether our merging is valid, as spelling and styling variations create different value of team names that refer to the same institution:

```{r}
# https://www.datanovia.com/en/lessons/identify-and-remove-duplicate-data-in-r/
df_enhanced |>
  filter((is.na(current_barthag) | is.na(prev_barthag_current)) & (year != 2022)) |>
  select(team, year, current_barthag, prev_barthag_current) |>
  distinct(team, year, .keep_all = TRUE) |>
  arrange(team)
```

All observations in this output does not have previous Barthag score, because they did not compete in NCAA Division I men's basketball in the previous year (they were in Division II). Therefore, we can temporarily safely declare that the merge on standardized team names is valid.

Next, I will determine whether a player is transferring to a higher-tier, lower-tier, or same-tier conference using the same programming approach that I did above.

The reason why I chose this over pure numerical ratings is that the high-major to low-major classification is also heavily tied to television rights and thus in turn affects the player's NIL, which is a crucial factor in a transferring decision that we have almost no data to control. In addition, this choice also attempts to replicate the insights from the original paper on "big fish, little pond".

```{r}
df_enhanced <- df_enhanced |>
  # Add information for players who have not yet transferred and who have just joined the team
  mutate(
    prev_conf_class = ifelse(year != 2022 & changed_team == 0, conf_class, prev_conf_class)
    ) |>
  mutate(
    prev_conf_class = case_when(
      team == "East Texas A&M" & year <= 2023	~ NA,
      team == "Le Moyne" & year <= 2024 ~ NA,
      team == "Lindenwood" & year <= 2023 ~ NA,
      team == "Mercyhurst" & year <= 2025 ~ NA,
      team == "New Haven" & year <= 2026 ~ NA,
      team == "Queens" & year <= 2023 ~ NA,
      team == "Southern Indiana" & year <= 2023 ~ NA,
      team == "Stonehill" & year <= 2023 ~ NA,
      team == "West Georgia" & year <= 2025 ~ NA,
      TRUE ~ prev_conf_class
      )
  )
```

```{r}
df_enhanced <- df_enhanced |>
  
  # This is just an intermediate step. I will also have another mutate to make the syntax cleaner
  mutate(to_better_conf = ifelse(changed_team == 0, 0, NA),
         to_worse_conf = ifelse(changed_team == 0, 0, NA),
         to_same_tier_conf = ifelse(changed_team == 0, 0, NA)) |>
  
  # Now do the actual filtering with the values that I have set to NA
  mutate(
    
    to_better_conf = case_when(
      (!is.na(conf_class)) & (!is.na(prev_conf_class)) & is.na(to_better_conf) &
        conf_class %in% c("high-major", "mid-major") & prev_conf_class == "low-major" ~ 1,
      (!is.na(conf_class)) & (!is.na(prev_conf_class)) & is.na(to_better_conf) &
        conf_class == "high-major" & prev_conf_class %in% c("mid-major", "low-major") ~ 1,
      year == 2022 ~ NA, # we don't have information
      TRUE ~ 0
      ),
    
    to_worse_conf = case_when(
      (!is.na(conf_class)) & (!is.na(prev_conf_class)) & is.na(to_worse_conf) &
        conf_class %in% c("low-major", "mid-major") & prev_conf_class == "high-major" ~ 1,
      (!is.na(conf_class)) & (!is.na(prev_conf_class)) & is.na(to_worse_conf) &
        conf_class == "low-major" & prev_conf_class %in% c("mid-major", "high-major") ~ 1,
      year == 2022 ~ NA,
      TRUE ~ 0
      ),
    
    to_same_tier_conf = case_when(
      (!is.na(conf_class)) & (!is.na(prev_conf_class)) & is.na(to_same_tier_conf) &
        conf_class == prev_conf_class ~ 1,
      year == 2022 ~ NA,
      TRUE ~ 0
      )
    
  )
```

I will also add the conference's average Barthag rating in the current year (`current_barthag_conf`) and the same conference's average Barthag rating in the previous year (`prev_barthag_current_conf`) as an indicator of the conference's overall competitiveness.

```{r}
conference_ratings <- df_team_ratings_full |>
  select(year, conf, barthag) |>
  group_by(year, conf) |>
  summarize(barthag_conf = mean(barthag)) |>
  ungroup() |>
  # I will also arrange conferences based on descending ratings for each year
  arrange(year, desc(barthag_conf))

conference_ratings
```

```{r}
df_enhanced <- df_enhanced |>
  # Add the conference rating in the current season
  left_join(conference_ratings, join_by(year, conf)) |>
  rename(current_barthag_conf = barthag_conf) |>
  # Add the conference rating of the current conference in the previous season
  left_join(conference_ratings, join_by(prev_year == year, conf)) |>
  rename(prev_barthag_current_conf = barthag_conf)

df_enhanced
```

Side note: How does the Barthag rating correlate with the traditional classification of NCAA Division I conferences into high-major, mid-major, and low-major?

```{r}
df_enhanced |>
  select(year, conf, conf_class, current_barthag_conf) |>
  distinct(year, conf, .keep_all = TRUE) |>
  mutate(conf_class = factor(conf_class, levels = c("high-major", "mid-major", "low-major"))) |>
  ggplot(aes(x = conf_class, y = current_barthag_conf)) +
  geom_boxplot() +
  facet_wrap(~ year)
```

#### 5. Whether the player is an overperformer or underperformer

The code is somewhat inspired by Kendall's implementation. Using the "big fish, little pond" paper, if a player performs greater than or equal to the team median in the previous year, they will be coded as an overperformer (`is_prev_overperformer = 1`) and vice versa. I use the median because player performance across a team or across the entire league is usually very heavily skewed.

In a similar way, I will create another indicator variable, `is_prev_underperformer`, to spot players who perform lower than the team average in the previous year. Note that `is_prev_overperformer` and `is_prev_underperformer` is not perfectly correlated, because for new players, `is_prev_overperformer = is_prev_underperformer = 0`.

```{r}
# But first, we need to fill in the previous points per game for players who do not transfer

# Note that because of my data wrangling, new players have a NA for prev_d1_year but still
# have prev_team (? I might address that later, but this might take a long time)

prev_team_fill <- df_enhanced |>
  select(id, year, changed_team, team, ppg, prev_d1_year, prev_team, prev_ppg) |>
  filter(!is.na(prev_d1_year)) |>
  # The "coalesce" command replace any NA of prev_team with the corresponding value in team
  mutate(prev_team = coalesce(prev_team, team))

prev_team_fill

# Select and rename necessary columns for a merge-friendly version with the original dataset
prev_team_fill_merge <- prev_team_fill |>
  select(id, year, prev_team_adjusted = prev_team)

prev_team_fill_merge
```

```{r}
# Get the overall PPG for (ID, year) player as a single dataset to better track data
player_ppg_only <- df_enhanced |>
  select(id, year, ppg)

player_ppg_only

# Get the (ID, year) observations where prev_ppg is still NA
na_prev_ppg <- prev_team_fill |>
  filter(is.na(prev_ppg)) |>
  select(id, prev_d1_year)

na_prev_ppg

# Then, join both datasets together to fill these NA values
prev_ppg_fill <- na_prev_ppg |>
  left_join(player_ppg_only, join_by(id, prev_d1_year == year)) |>
  rename(prev_ppg = ppg)

prev_ppg_fill

full_prev_ppg_fill <- prev_team_fill |>
  select(id, prev_d1_year, prev_ppg) |>
  rows_patch(prev_ppg_fill, by = c("id", "prev_d1_year")) |>
  # Rename columns to better merge with the original dataset
  rename(prev_ppg_adjusted = prev_ppg)

full_prev_ppg_fill
```

```{r}
# Finally, we can work on the merging into the original dataset
df_enhanced <- df_enhanced |>
  left_join(prev_team_fill_merge, join_by(id, year)) |>
  left_join(full_prev_ppg_fill, join_by(id, prev_d1_year)) |>
  # I can also use "coalesce" again to fill in the NA positions
  mutate(prev_team = coalesce(prev_team, prev_team_adjusted),
         prev_ppg = coalesce(prev_ppg, prev_ppg_adjusted)) |>
  # Finally, I will drop the intermediary variables created for the fill-in purposes
  select(-prev_team_adjusted, -prev_ppg_adjusted)

df_enhanced
```

```{r}
# Dataset of the median points per game for each team in each year
median_ppg_team <- df_enhanced |>
  select(year, team, ppg) |>
  group_by(year, team) |>
  summarize(prev_median_ppg_team = median(ppg)) |>
  ungroup()

median_ppg_team
```

```{r}
# Join the median points per game of team in the previous year in D1
# We use a comparison to see if the player was previously an overperformer or not

df_enhanced <- df_enhanced |>
  left_join(median_ppg_team, join_by(prev_d1_year == year, prev_team == team)) |>
  
  # We also want to remove the previous team for players who have not yet joined
  # Division I before
  
  mutate(prev_team = ifelse(year != 2022 & is.na(prev_d1_year), NA, prev_team),
         is_prev_overperformer = ifelse(
           prev_ppg >= prev_median_ppg_team & (!is.na(prev_ppg)) &
             (!is.na(prev_median_ppg_team)), 1, 0
           ),
         is_prev_underperformer = ifelse(
           prev_ppg < prev_median_ppg_team & (!is.na(prev_ppg)) &
             (!is.na(prev_median_ppg_team)), 1, 0)
         )
```

```{r}
df_enhanced
```

#### Final step

Finally, I will export the dataframe to a CSV file for later use (This should not be pushed directly into a public GitHub repo):

```{r}
write.csv(df_enhanced, "player_stats_hoopR.csv", row.names = FALSE)
```