Some useful resources that I happened to find:

- Propensity score matching - Caliper width on logit of the propensity score: <https://stackoverflow.com/questions/72223928/how-to-use-matchit-r-package-to-use-caliper-width-on-logit-of-the-propensity-sco>

#### 1. Data preprocessing for modeling

```{r}
# Clear the environment and read the prepared dataset
rm(list = ls()) # Clear the environment

library(tidyverse)
library(lme4) # For mixed effects models
library(modelsummary) # Display the results of regression models like in publications
library(MatchIt) # For propensity score matching models
library(cobalt) # Generate a ggplot representation of the Love plot

df <- read.csv("player_stats_hoopR.csv")
df
```

```{r}
# Change necessary variable data types for the models to work properly
df <- df |>
  mutate(id = as.factor(id),
         player = as.factor(player),
         pos = as.factor(pos),
         team = as.factor(team),
         conf = as.factor(conf)) |>
  
  # Filter out players who have changed team and skipped the year before
  # (they are likely to transfer out of DI to lower divisions or JUCOs)
  filter(!(skipped_year_before == 1 & changed_team == 1)) |>
  
  # Filter out for NA values in the dataset
  drop_na(changed_team, prev_ppg, pos, exp, year, current_barthag, current_barthag_conf,
          id, team, conf)

  # We don't use data influenced by the pandemic period

df
```

Generate separate datasets for overperformer and underperformer transfers (changing the definition of the variable `changed_team`):

```{r}
# Changed the definition of transferred variable: only including underperformer transfers
df_underperf <- df |>
  mutate(underperf_transfer = changed_team * is_prev_underperformer) |>
  select(-changed_team) |>
  rename(changed_team = underperf_transfer)
```

```{r}
# Changed the definition of transferred variable: only including overperformer transfers
df_overperf <- df |>
  mutate(overperf_transfer = changed_team * is_prev_overperformer) |>
  select(-changed_team) |>
  rename(changed_team = overperf_transfer)
```

Keep the mixed effects model and multiple linear regression model as a baseline for comparison:

```{r}
# Note that we do not include team and player fixed effects because of the singleton issue
# and high runtime associated with creating thousands of binary variables

# Normal conditions:
linreg_normal <- lm(ppg ~ changed_team + prev_ppg + pos + exp + year + conf +
               current_barthag + current_barthag_conf,
             data = df)

no_matching_normal <- lmer(ppg ~ changed_team + prev_ppg + pos + exp + year + conf +
                             current_barthag + current_barthag_conf + (1|team:id),
                           data = df)

# Underperformers dataset:
linreg_underperf <- lm(ppg ~ changed_team + prev_ppg + pos + exp + year + conf +
                         current_barthag + current_barthag_conf,
                       data = df_underperf)

no_matching_underperf <- lmer(ppg ~ changed_team + prev_ppg + pos + exp + year + conf +
                             current_barthag + current_barthag_conf + (1|team:id),
                           data = df_underperf)

# Overperformers dataset:
linreg_overperf <- lm(ppg ~ changed_team + prev_ppg + pos + exp + year + conf +
                        current_barthag + current_barthag_conf,
                       data = df_overperf)

no_matching_overperf <- lmer(ppg ~ changed_team + prev_ppg + pos + exp + year + conf +
                             current_barthag + current_barthag_conf + (1|team:id),
                           data = df_overperf)
```

Preliminary diagnostics - Approximating ratio to match transfers with non-transfers in all three transfer definitions:

```{r}
df |>
  count(changed_team)
# Approximated ratio: 8335 non-transfers / 4278 transfers = 1.95 ~ 2

df_underperf |>
  count(changed_team)
# Approximated ratio: 11058 non-transfers / 1555 transfers = 7.11 ~ 7

df_overperf |>
  count(changed_team)
# Approximated ratio: 9898 non-transfers / 2715 transfers = 3.64 ~ 4
```

### 1. Propensity score matching object implementation

```{r}
# Functions to automate the propensity score matching workflow without copying and pasting code

propensity_score_matching <- function(my_data, match_ratio = 2) {
  
  # my_data: The dataset in question

  # match_ratio: The ratio to match between treatment and control.
  # This depends on the ratio between transfers and non-transfers for each group.
  
  # Function returns the match_object, match_object_data, and mixed_effects_matched for the
  # dataset. We can consider the result as a "propensity_score_matching" object
  
  match_object <- matchit(changed_team ~ pos + exp + current_barthag + current_barthag_conf,
                          data = my_data, method = "nearest", distance = "glm",
                          link = "linear.logit", caliper = 0.2, std.caliper = TRUE,
                          replace = FALSE, ratio = match_ratio)

  match_object_data <- match.data(match_object)

  mixed_effects_matched <- lmer(ppg ~ changed_team + prev_ppg + pos + exp + year + conf +
                                  current_barthag + current_barthag_conf + (1|team:id),
                                data = match_object_data)
  
  results <- list("match_object" = match_object, "match_object_data" = match_object_data,
                  "mixed_effects_matched" = mixed_effects_matched)
  
  return(results)
  
}

```

```{r}
# Implement the function with normal conditions, underperformers, and overperformers
matching_normal <- propensity_score_matching(df, match_ratio = 2)
matching_underperf <- propensity_score_matching(df_underperf, match_ratio = 7)
matching_overperf <- propensity_score_matching(df_overperf, match_ratio = 4)
```

```{r}
# Obtain individual objects of each propensity score matching specification

match_object_normal <- matching_normal$match_object
match_object_underperf <- matching_underperf$match_object
match_object_overperf <- matching_overperf$match_object

match_object_data_normal <- matching_normal$match_object_data
match_object_data_underperf <- matching_underperf$match_object_data
match_object_data_overperf <- matching_overperf$match_object_data

mixed_effects_normal <- matching_normal$mixed_effects_matched
mixed_effects_underperf <- matching_underperf$mixed_effects_matched
mixed_effects_overperf <- matching_overperf$mixed_effects_matched
```

### 2. Balance diagnostics for propensity score matching models

Plotting balance diagnostics of the matched models (Love plots) with a large function:

```{r}
# Function to obtain a nicely formatted Love plot for any propensity score matching data

ps_matching_love_plot <- function(my_data) {
  
  # Get the matchit object to create the Love plot
  love_plot_object <- propensity_score_matching(my_data)$match_object
  
  # drop.distance = TRUE: Drop the distance metric when visualizing the Love plot
  match_object_love_plot <- love.plot(love_plot_object, stat = "mean.diffs",
                                      var.order = "unadjusted", threshold = 0.1,
                                      drop.distance = TRUE, binary = "std", abs = TRUE) +
    
    # Edit the labels and colors to be colorblind-friendly
    labs(title = "",
         color = "Group",
         x = "Absolute Standardized Mean Differences") +
    scale_y_discrete(labels = c("exp" = "Experience level",
                                "current_barthag_conf" = "Conference rating",
                                "current_barthag" = "Team rating",
                                "pos_center" = "Position is Center",
                                "pos_guard" = "Position is Guard",
                                "pos_forward" = "Position is Forward")) +
    scale_color_manual(values = c("darkred", "darkblue"),
                       labels = c("All data", "Matched data")) +
    
    # Scale up the zero line and the 0.1 threshold line
    geom_vline(xintercept = 0, color = "black") +
    geom_vline(xintercept = 0.1, color = "darkorange",
               linewidth = 0.7, alpha = 0.5, linetype = "dashed") +
    
    # Set the theme and other edits to the plot for aesthetic purposes
    theme_bw() +
    
    theme(
      # Remove unnecessary vertical gridlines from the graph
      panel.grid.major.y = element_blank(),
      panel.grid.minor.y = element_blank(),
      
      # Scale up text to show in presentation and report
      axis.text.x = element_text(size = 16),
      axis.text.y = element_text(size = 16, face = "bold", hjust = 0.5),
      axis.title.x = element_text(size = 16, face = "bold", margin = margin(t = 15)),
      axis.title.y = element_blank(),
      
      # We also need to scale up the legends
      legend.title = element_text(size = 16, face = "bold", hjust = 0.5),
      legend.text = element_text(size = 16)
    )

  # Return the final plot for later use
  return(match_object_love_plot)
  
}
```

```{r}
balance_diagnostics_normal <- ps_matching_love_plot(df)
balance_diagnostics_underperf <- ps_matching_love_plot(df_underperf)
balance_diagnostics_overperf <- ps_matching_love_plot(df_overperf)
```

```{r}
balance_diagnostics_normal
balance_diagnostics_underperf
balance_diagnostics_overperf
```

```{r}
# Scale down graph labels to be appropriate for a report

report_balance_diagnostics_normal <- balance_diagnostics_normal +
    theme(
      axis.text.x = element_text(size = 13),
      axis.text.y = element_text(size = 13, face = "bold", hjust = 0.5),
      axis.title.x = element_text(size = 13, face = "bold", margin = margin(t = 15)),
      axis.title.y = element_blank(),
      
      legend.title = element_text(size = 13, face = "bold", hjust = 0.5),
      legend.text = element_text(size = 13)
    )
#balance_diagnostics_underperf
#balance_diagnostics_overperf

report_balance_diagnostics_normal

ggsave("saved/report_love_plot_all.png", report_balance_diagnostics_normal, width = 8, height = 6)
```


### 3. Summarize the effects in tabular and graphical format

```{r}
# Summarize the results in a nicely formatted table

models_matching <- list(
  "(1)" = mixed_effects_normal,
  "(2)" = mixed_effects_underperf,
  "(3)" = mixed_effects_overperf
)

models_matching_coef_map <- c(
  "changed_team" = "Transferred",
  "prev_ppg" = "Points per game in previous season",
  "exp" = "Experience level",
  "SD (Intercept teamid)" = "SD (Team × Player)",
  "SD (Observations)" = "SD (Observations)"
)

# Note that conditional R^2 (a.k.a. R^2 that accounts for both fixed and random effects)
# may not be available because there are some random effects with zero intraclass correlation

models_matching_gof_map <- tibble(
  raw = c("nobs", "r2.marginal", "r2.conditional"),
  clean = c("Number of observations", "Marginal R<sup>2</sup>", "Conditional R<sup>2</sup>"),
  fmt = c("%.0f", "%.3f", "%.3f")
)

models_matching_summary <- modelsummary(models_matching, stars = TRUE,
                                        coef_map = models_matching_coef_map,
                                        gof_map = models_matching_gof_map)
models_matching_summary
```

Plotting a graph of model point estimates and their confidence intervals (95%):

```{r}
# Helper function: Extract the point estimates of a named variable and their confidence interval
# as a dataframe for plotting

compare_causal_models <- function(var_name, model_list) {
  # var_name: Variable name of interest (need to be the same across all models).
  # var_name must be in quotes
  
  # model_list: List of models (not named) to compare the effects of a variable.
  # model_list should contain an user-specified name and the model variable itself.
  
  # Create a blank dataframe to store the results
  result <- tibble(
    model_name = character(),
    estimate = numeric(),
    std.error = numeric(),
    statistic = numeric(),
    p.value = numeric(),
    conf.low = numeric(),
    conf.high = numeric(),
  )
  
  # Extract the estimates and append to the result dataframe
  for (model_name in names(model_list)) {
    
    # Extract the model's name as a value:
    # https://stackoverflow.com/questions/24309910/how-to-get-name-of-variable-in-r-substitute
    model <- model_list[[model_name]]
    
    model_name_df <- tibble(
      model_name = c(model_name)
    )
    
    model_df <- get_estimates(model) |>
      filter(term == var_name) |>
      select(estimate, std.error, statistic, p.value, conf.low, conf.high)
    
    # Combine the model names with model statistics together, then append the result to the
    # returning dataframe
    model_df <- bind_cols(model_name_df, model_df)
    result <- bind_rows(result, model_df)
  }
  
  # Return the resulting dataframe
  return(result)
}
```

```{r}
# Function to obtain a nicely formatted comparing effects for multiple model configurations

compare_causal_plot <- function(compare_causal_models_df) {
  # compare_causal_models_df is a dataframe generated from the previous function
  
  # Generate a blank canvas and add a zero baseline for reference
  resulting_plot <- compare_causal_models_df |>
    ggplot() +
    geom_vline(xintercept = 0, color = "darkred", linetype = "dashed",
               linewidth = 0.7, alpha = 0.5) +
    
    # Supply the plot with data
    geom_point(aes(x = estimate, y = model_name), color = "blue", size = 3) +
    geom_errorbar(aes(xmin = conf.low, xmax = conf.high, y = model_name),
                  linewidth = 1.1, width = 0.25, orientation = "y") +
    # Use vjust with geom_text to move labels of point estimates up a little
    geom_text(aes(label = round(estimate, 3), x = estimate, y = model_name),
              vjust = -0.9, fontface = "bold", size = 6, color = "black") +
    
    # Modify axis labels for readability
    labs(x = "Expected change in points per game following a transfer",
         y = "") +
    
    # Set the theme and other edits to the plot for aesthetic purposes
    theme_bw() +
    
    theme(
      # Remove unnecessary vertical gridlines from the graph
      panel.grid.major.y = element_blank(),
      panel.grid.minor.y = element_blank()
    )
  
  # Return the resulting plot
  return(resulting_plot)
}
```

```{r}
my_model_list <- list("Multiple Linear\nRegression" = linreg_normal,
                      "Mixed Effects\nWithout Matching" = no_matching_normal,
                      "Mixed Effects\nWith Matching" = mixed_effects_normal)

model_comparison_df <- compare_causal_models("changed_team", my_model_list)
compare_causal_plot_all <- compare_causal_plot(model_comparison_df) +
  theme(
    axis.text.x = element_text(size = 16),
    axis.text.y = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.title.x = element_text(size = 16, face = "bold", margin = margin(t = 15)),
    axis.title.y = element_blank()
  )

compare_causal_plot_all
```

```{r}
full_model_list <- list("Multiple Linear Regression" = linreg_normal,
                        "Mixed Effects Without Matching" = no_matching_normal,
                        "Mixed Effects With Matching" = mixed_effects_normal,
                        
                        "Multiple Linear Regression 2" = linreg_underperf,
                        "Mixed Effects Without Matching 2" = no_matching_underperf,
                        "Mixed Effects With Matching 2" = mixed_effects_underperf,
                        
                        "Multiple Linear Regression 3" = linreg_overperf,
                        "Mixed Effects Without Matching 3" = no_matching_overperf,
                        "Mixed Effects With Matching 3" = mixed_effects_overperf)

full_model_comparison_df <- compare_causal_models("changed_team", full_model_list) |>
  cbind(condition = c("All transfers", "All transfers", "All transfers",
                      "Underperformer transfers only", "Underperformer transfers only",
                      "Underperformer transfers only", "Overperformer transfers only",
                      "Overperformer transfers only", "Overperformer transfers only")) |>
  
  # Adjust the model names to be similar to each other
  # Note that we cannot pass similar model names to the list for compare_causal_models function
  # because of constraints in naming list items in R itself
  mutate(model_name = rep(c("Multiple Linear\nRegression", "Mixed Effects\nWithout Matching",
                            "Mixed Effects\nWith Matching"), times = 3)) |>
  mutate(model_name = factor(model_name, levels = c("Multiple Linear\nRegression",
                                                    "Mixed Effects\nWithout Matching",
                                                    "Mixed Effects\nWith Matching")))

full_model_comparison_df
```

```{r}
# library(ggrepel)
# Ensure that the text does not touch the data points or error bars with geom_text_repel

compare_plot_robust <- full_model_comparison_df |>
  ggplot(aes(y = model_name, x = estimate)) +
  geom_vline(xintercept = 0, color = "darkred", linetype = "dashed",
               linewidth = 0.7, alpha = 0.5) +
  
  # Supply the plot with data
  geom_point(color = "blue") +
  geom_errorbar(aes(xmin = conf.low, xmax = conf.high),
                  width = 0.3, orientation = "y") +
  geom_text(aes(label = round(estimate, 3), x = estimate, y = model_name),
            nudge_y = 0.3, fontface = "bold", size = 6, color = "black") +
  facet_wrap(~ condition, scales = "free_x") +
  
  # Modify axis labels for readability
  labs(x = "Expected change in points per game following a transfer",
       y = "") +
  
  # Plot with custom order
  # Note that the plot shows Mixed Effects with Matching first, then Mixed Effects Without
  # Matching, then Multiple Linear Regression by default
  scale_y_discrete(limits = rev) +
  
  # Set the theme and other edits to the plot for aesthetic purposes
  theme_bw() +
  
  theme(
    
    # Remove unnecessary vertical gridlines from the graph
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    
    # Adjust textual elements including axis labels and facet plot text
    axis.title.x = element_text(size = 16, face = "bold", margin = margin(t = 15)),
    axis.text.x = element_text(size = 16),
    axis.title.y = element_blank(),
    axis.text.y = element_text(size = 16, face = "bold", hjust = 0.5),
    strip.text = element_text(size = 17, face = "bold")
    
    )
  
compare_plot_robust
```

#### Additional table visualization to be informative

```{r}
# Visualize models: linreg_normal, no_matching_normal, and mixed_effects_normal with modelsummary

# Summarize the results in a nicely formatted table

models_matching_add <- list(
  "(1)" = linreg_normal,
  "(2)" = no_matching_normal,
  "(3)" = mixed_effects_normal
)

models_matching_add_coef_map <- c(
  "changed_team" = "Transferred",
  "prev_ppg" = "Points per game in previous season",
  "exp" = "Experience level",
  "SD (Intercept teamid)" = "SD (Team × Player)",
  "SD (Observations)" = "SD (Observations)"
)

# Note that conditional R^2 (a.k.a. R^2 that accounts for both fixed and random effects)
# may not be available because there are some random effects with zero intraclass correlation

models_matching_add_gof_map <- tibble(
  raw = c("nobs", "r2.adjusted", "r2.marginal", "r2.conditional"),
  clean = c("Number of observations", "Adjusted R<sup>2</sup>",
            "Marginal R<sup>2</sup>", "Conditional R<sup>2</sup>"),
  fmt = c("%.0f", "%.3f", "%.3f", "%.3f")
)

additional_summary <- modelsummary(models_matching_add, stars = TRUE,
                                   coef_map = models_matching_add_coef_map,
                                   gof_map = models_matching_add_gof_map,
                                   output = "saved/model_selection_summary.png")
additional_summary
```


#### Saving models and plots to make the final report or presentation cleaner

Note that you need to create a folder in your working directory called `saved` for the code to be reproducible.

```{r}
# Saving for display on a screen (i.e. presentations)
ggsave("saved/love_plot_all.png", balance_diagnostics_normal, width = 8, height = 6)
ggsave("saved/love_plot_underperf.png", balance_diagnostics_underperf, width = 8, height = 6)
ggsave("saved/love_plot_overperf.png", balance_diagnostics_overperf, width = 8, height = 6)

ggsave("saved/causal_plot_baseline.png", compare_causal_plot_all, width = 8, height = 6)
ggsave("saved/causal_plot_robust.png", compare_plot_robust, width = 13, height = 6)
```

For posters, we need to scale up the graphs and axis labels for people to read at a certain distance.

```{r}
# Comparing effects if we consider all types of transfers as a whole

compare_plot_robust_poster <- full_model_comparison_df |>
  ggplot(aes(y = model_name, x = estimate)) +
  geom_vline(xintercept = 0, color = "darkred", linetype = "dashed",
               linewidth = 2, alpha = 0.5) +
  
  # Supply the plot with data
  geom_point(color = "blue", size = 8) +
  geom_errorbar(aes(xmin = conf.low, xmax = conf.high),
                  width = 0.3, orientation = "y", linewidth = 2) +
  geom_text(aes(label = round(estimate, 3), x = estimate, y = model_name),
            nudge_y = 0.3, fontface = "bold", size = 13, color = "black") +
  facet_wrap(~ condition, scales = "free_x") +
  
  # Modify axis labels for readability
  labs(x = "Expected change in points per game following a transfer",
       y = "") +
  
  # Plot with custom order
  # Note that the plot shows Mixed Effects with Matching first, then Mixed Effects Without
  # Matching, then Multiple Linear Regression by default
  scale_y_discrete(limits = rev) +
  
  # Set the theme and other edits to the plot for aesthetic purposes
  theme_bw() +
  
  theme(
    
    # Remove unnecessary vertical gridlines from the graph
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    
    # Adjust textual elements including axis labels and facet plot text
    axis.title.x = element_text(size = 40, face = "bold", margin = margin(t = 15)),
    axis.text.x = element_text(size = 40),
    axis.title.y = element_blank(),
    axis.text.y = element_text(size = 40, face = "bold", hjust = 0.5),
    strip.text = element_text(size = 40, face = "bold")
    
    )

ggsave("saved/poster_causal_plot_robust.png", compare_plot_robust_poster,
       width = 35, height = 10)
```


```{r, eval = FALSE}
# Save the propensity score matching model results and diagnostics plots as a checkpoint
saveRDS(models_matching_summary, "saved/models_matching_summary.rds")
saveRDS(compare_causal_plot, "saved/compare_causal_plot.rds")
saveRDS(match_object_love_plot, "saved/match_object_love_plot.rds")
```

#### Appendices and Drafts

Helper function to assist with the x-axis range of Love plots for balance diagnostics in propensity score matching models created by `MatchIt`:

```{r, eval = FALSE}
# Helper function to assist with the Love plots:

# Extract the largest standardized mean differences in a matchit summary dataset
max_std_mean_diff <- function(matchit_summary) {
  
  # Extract the standardized mean differences of covariates among all data and matched data
  sum_all_mean_diff <- as.data.frame(matchit_summary$sum.all) |>
    select(`Std. Mean Diff.`)
  sum_matched_mean_diff <- as.data.frame(matchit_summary$sum.matched) |>
    select(`Std. Mean Diff.`)
  
  # Combine two datasets together and return the maximum value of the column
  result_df <- bind_rows(sum_all_mean_diff, sum_matched_mean_diff)
  result <- max(result_df$`Std. Mean Diff.`)
  
  return(result)
}
```