---
title: "Effects of Transferring on Player Performance"
subtitle: "An Analysis on Division I Men's Basketball"
author:
  - An Le, Bates College
  - Kendall Lucas, Duquesne University
date: "July 24, 2026"
toc: true
format:
  html:
    theme: cosmo
    html-math-method: katex
    self-contained: true
    anchor-sections: false
execute:
  echo: false
  warning: false
  message: false
editor: source
---

------------------------------------------------------------------------

```{r}
# Just a code to clear the variables environment
rm(list = ls())
```

## Methods

We estimated the effect of transferring on player performance using a mixed effects regression model following propensity score matching. The results that informed our model selection process is mentioned in detail in [Appendix 2](#appendix-2).

We also attempted to validate the findings of [Cole et al. (2026)](https://doi.org/10.51221/sc.jiia.2026.19.1.15) about the differential effects of transferring on player performance related to whether the player was overperforming or underperforming in their previous season at the NCAA Division I men's basketball. By changing the definition of the $\text{Transferred}$ variable to only include players who transferred as overperformers (above or equal to the median performance among all players in the team) and underperformers (below the median performance among all players in the team), we expected to see a more pronounced effect of transferring on player performance.

### 1. Propensity score matching

First, we performed propensity score matching using the MatchIt package in R [(Ho et al., 2011)](https://doi.org/10.18637/jss.v042.i08). As certain groups of players are more likely to transfer than others, the matching ruled out possibilities of a selection bias when making inferences about the effect of transferring on player performance.

The process involved matching a transferred player to $n$ non-transferred players without replacement. Here, $n$ is an integer representing the approximate ratio of the number of transfers to the number of non-transfers. Using our exploratory data analysis, $n$ equals $2$ if we considered all types of transfers, $7$ for overperformer transfers, and $4$ for underperformer transfers. Our flexible choice of the matching ratio allowed us to obtain a representative sample of NCAA players after the matching process.

We specified the following logistic regression model to determine the propensity scores associated with how likely each player is an incoming transfer for each player and year combination:

$$
\widehat{\text{Transferred}_{ijkt}} = \beta_1 \text{Position}_i + \beta_2 \text{Experience}_{it} + \beta_3 \text{TeamRating}_{jt} + \beta_4 \text{ConfRating}_{kt}
$$

Regarding the matching method, we used MatchIt's default nearest neighbor matching and set the caliper width to "$0.2$ the standard deviation of the logit of the propensity score" using the recommendation from [Austin (2011)](https://doi.org/10.1002/pst.433), which prevents players with vastly different statistics to be matched with each other. For the variable selection, we only included covariates that could affect a player's performance, but do not necessarily affect their choice of staying in the team or transferring. According to [Brookhart et al. (2006)](https://doi.org/10.1093/aje/kwj149), this choice of variables would improve the accuracy of our inferences because it would decrease the variance of our estimated effects of a transfer without increasing bias.

Before we proceeded to fitting a regression model on the matched data, we constructed Love plots to check for balance between the treatment group and control group, which is outlined in [Appendix 3](#appendix-3). In the matched data, all covariates had an absolute standardized mean difference of less than or equal to $0.1$, which implies that the treatment and control groups were similar and we had properly controlled for confounding variables related to players' choices to transfer.

### 2. Mixed effects regression model

After we controlled for selection biases in players' transfer decisions using propensity score matching, we fit a mixed effects regression model on the matched data to precisely determine the effect of transferring on player performance, controlling for more factors such as individual athletic ability and team context. Here, we quantified a player's performance by their points per game (PPG), since this metric is easily accessible and interpretable for non-statistical audiences.

The full equation for our mixed effects regression model is:

$$
\widehat{\text{PPG}_{ijkt}} = a_{ij} + \beta_1 \text{Transferred}_{ij} + \beta_2 t + \beta_3 \text{PPG}_{i,t-1} + \beta_4 \text{Position}_{i} + \beta_5 \text{Experience}_{it} +\\ \beta_6 \text{Conference}_{kt} + \beta_7 \text{TeamRating}_{jt} + \beta_8 \text{ConfRating}_{kt} + \epsilon_{ijkt}
$$

In this model, the coefficients numbered from $\beta_1$ to $\beta_8$, represents the fixed effects, and $\epsilon_{ijkt}$ represents the error term.

Using the model selection results from [Appendix 2](#appendix-2) and domain knowledge, we specify an intercept $a_{ij}$ specific to each player and team combination:

$$
a_{ij} = \alpha_0 + u_{ij}, \quad \quad u_{ij} \sim N(0,\tau^2).
$$

The coefficient $\beta_1$ for the $\text{Transferred}$ variable would give the magnitude of the effect of transferring on a player's points per game in the current year, while the standard error associated with that coefficient would determine the statistical significance of our estimates.

## Results

(1) Fixed effects or mixed effects model? If mixed effects, which effects?

```{r}
models_fe_summary <- readRDS("saved/models_fe_summary.rds")
models_mixed_summary <- readRDS("saved/models_mixed_summary.rds")
```

The fixed effects model:

```{r}
models_fe_summary
```

The mixed effects model:

```{r}
models_mixed_summary
```

(2) With and without matching (optional? if so, include balance diagnostics)

```{r}
models_matching_summary <- readRDS("saved/models_matching_summary.rds")
compare_causal_plot <- readRDS("saved/compare_causal_plot.rds")
match_object_love_plot <- readRDS("saved/match_object_love_plot.rds")
```

```{r}
models_matching_summary
```

```{r}
compare_causal_plot
```

\[Note that the `theme_set` command does not save the theme in a `.rds` file.\]

```{r}
match_object_love_plot
```

(3) Attempt to replicate the big fish little pond paper [(Cole et al., 2026)](https://doi.org/10.51221/sc.jiia.2026.19.1.15)

## Discussion

\[Add some connection between results and domain knowledge\]

### 2. Limitations and next steps:

- Weirdness in eligibility data (changes in eligibility rules, redshirting, one-and-done, transferring involving outside DI institutions, etc.)

- Transfers for non-athletic-related reasons (eg: academic or proximity to home, NIL, injury, coaching changes) -\> More data collection is needed

- Points per game might not be an accurate indicator of performance -\> use publicly available datasets for points per possession, RAPM, etc.

- Explore more ways to fit the model, such as using random slopes, GLMs and GAMs to capture non-linear relationships between variables to the outcome (points per game).

## Acknowledgments

We would like to give special thanks to:

- **Professors Ron Yurko and Zach Branson**, who had offered us great guidance on the methodology for causal inference in sports,
- **Katerina Wu and Lucca Ferraz** from the University of Pittsburgh Athletics, who provided invaluable data and domain knowledge,
- **Sara Colando** and **Erin Franke**, the program instructors for their lectures,
- **Princess Allotey**, **JungHo Lee**, and **Yuchen Chen**, the program's Teaching Assistants for their statistical expertise, their presence and willingness to help in the daily labs.

## References

Austin, P. C. (2011). Optimal caliper widths for propensity‐score matching when estimating differences in means and differences in proportions in observational studies. *Pharmaceutical Statistics, 10*(2), 150–161. <https://doi.org/10.1002/pst.433>

Bates, D., Maechler, M., Bolker, B., Walker, S. (2015). Fitting Linear Mixed-Effects Models Using lme4. *Journal of Statistical Software, 67*(1), 1-48. <https://doi.org/10.18637/jss.v067.i01>

Brookhart, M. A., Schneeweiss, S., Rothman, K. J., Glynn, R. J., Avorn, J., & Stürmer, T. (2006). Variable selection for propensity score models. *American Journal of Epidemiology, 163*(12), 1149–1156. <https://doi.org/10.1093/aje/kwj149>

Cole, K., Pearman, F., & Khanna, S. (2026). Big Fish or Little Pond? National Evidence on the Relationship Between Transferring and Athletic Performance Within Division I Men's College Basketball. *Journal of Issues in Intercollegiate Athletics, 19*(1). <https://doi.org/10.51221/sc.jiia.2026.19.1.15>

Gilani, S. (2026). *hoopR: Access Men's Basketball Play by Play Data*. (Version 3.1.0) \[Computer software\]. <https://github.com/sportsdataverse/hoopR>

Gonzalez Sanchez, A., Martinez, S., Yurko, R., Elmore, R., & Macdonald, B. (2024). Beyond the Box Score: Does Icing the Field Goal Kicker Work in the NFL? *CHANCE, 37*(3), 41–48. <https://doi.org/10.1080/09332480.2024.2415841>

Ho, D., Imai, K., King, G., Stuart, E. (2011). MatchIt: Nonparametric Preprocessing for Parametric Causal Inference. *Journal of Statistical Software, 42*(8), 1-28. <https://doi.org/10.18637/jss.v042.i08>.

## Contact Information and Code Availability

We are happy to answer questions and receive feedback on the presentation of our capstone project. Here are our email information:

- **An Le** (Bates College): [thaianle1102\@gmail.com](mailto:thaianle1102@gmail.com)
- **Kendall Lucas** (Duquesne University): [kendalllucas724\@gmail.com](mailto:kendalllucas724@gmail.com)

All of our code are available on a public GitHub repository named [`ncaa-capstone-cmsacamp-2026`](https://github.com/thaianle/ncaa-capstone-cmsacamp-2026/tree/main).

## Appendix 1: Variable naming in equations

\[Create a table with variable in mathematical equations (in the Methods section) and their detailed description, some data wrangling steps are important such as changing `exp` to a continuous scale of 1-5\]

Notation: $i$ is the individual athlete, $j$ is the team, $k$ is the conference, and $t$ is the year.

## Appendix 2: Model selection process {#appendix-2}

\[Fixed vs random effects model\]

## Appendix 3: Balance diagnostics for propensity score matching {#appendix-3}

\[Show a table or Love plot: t-test - are the means for the control and treatment groups actually similar?\]
