**NBA Age Curve**

Load data and packages

```{r}
library(cbbdata)
library(tidyverse)
library(broom)
library(hoopR)
library(gt)
#install.packages("hoopR", repos = c("https://sportsdataverse.r-universe.dev", "https#://cloud.r-project.org"))
df_2022 <- cbd_torvik_player_season(year = 2022)

df_2023 <- cbd_torvik_player_season(year = 2023)

df_2024 <- cbd_torvik_player_season(year = 2024)

df_2025 <- cbd_torvik_player_season(year = 2025)

df_2026 <- cbd_torvik_player_season(year = 2026)

df <- bind_rows(df_2022, df_2023, df_2024, df_2025)


```

ratings table for 2021-2026

```{r}

#df_team_ratings_2022 <- torvik_ratings(year = 2022)
#df_team_ratings_2023 <- torvik_ratings(year = 2023)
#df_team_ratings_2024 <- torvik_ratings(year = 2024)
#df_team_ratings_2025 <- torvik_ratings(year = 2025)
#df_team_ratings_2026 <- torvik_ratings(year = 2026)

#ratings_df <- bind_rows(df_team_ratings_2022, df_team_ratings_2023, 
#                        df_team_ratings_2024, df_team_ratings_2025)

torvik_ratings_all <- bind_rows(
  torvik_ratings(year = 2021) |> mutate(season = 2021),
  torvik_ratings(year = 2022) |> mutate(season = 2022),
  torvik_ratings(year = 2023) |> mutate(season = 2023),
  torvik_ratings(year = 2024) |> mutate(season = 2024),
  torvik_ratings(year = 2025) |> mutate(season = 2025)
)

url_2021 <- "http://barttorvik.com/2021_team_results.csv"
url_2026 <- "http://barttorvik.com/2026_team_results.csv"
torvik_2021 <- read_csv(url_2021, col_names = FALSE)
torvik_2026 <- read_csv(url_2026, col_names = FALSE)


torvik_raw_2021 <- torvik_2021 |> 
  slice(-1) |>
  select(
    rank = X1, 
    team = X2) |>
  mutate(rank = as.integer(rank),
         year = 2021)

torvik_raw_2026 <- torvik_2026 |> 
  slice(-1) |>
  select(
    rank = X1, 
    team = X2) |>
  mutate(rank = as.integer(rank),
         year = 2026)

torvik_ratings_all <- bind_rows(
  torvik_ratings_all, torvik_raw_2021, torvik_raw_2026
)

torvik_ratings_all <- torvik_ratings_all |>
  select(team, rank, year)
```

Divide the dataset into transfers and non-transfers and just do some data cleaning

```{r}
df <- df |>
  arrange(id, year) |>
  group_by(id) |>
  mutate(prev_team = lag(team),
         prev_year = lag(year),
         #if they have a previous team AND 
         #the year is the one right after the previous year AND 
         #the team they're on is not the same as the previous team THEN
         #changed_team is through IF NOT THEN
         #changed_team is false 
         changed_team = ifelse(!is.na(prev_team) &
                                 year == (prev_year + 1) &
                                 team != prev_team,
                               TRUE,
                               FALSE)) |>
  ungroup()

transfer_df <- df |>
  group_by(player) |>
  filter(n_distinct(team) > 1) |>
  ungroup()

df <- df |>
  mutate(has_transferred = player %in% transfer_df$player)

df <- df |>
  filter((((prev_year + 1) == year)) | (year == 2022))

df <- df |>
  relocate(prev_team, prev_year, changed_team, year)

# this comes out to 0 meaning there are no year jumps in the data
#df |>
#  filter((prev_year + 1) != year)

# this fixes all the loophole issues we've discovered so far
# this is a dataset of all the players who are not new transfers but there are some where 
# they have transferred before. i should changed this 
non_transfers_df <- df |>
  filter((has_transferred == FALSE) & (changed_team == FALSE) | (exp == "Fr"))


#non_transfers_df|>
#  filter(prev_team != team)



```

get experience levels sorted out (add SSr)

```{r}
non_transfers_df <- non_transfers_df |>
  mutate(prev_exp = lag(exp)) |>
  filter((exp == "Fr") | (exp == "So" & prev_exp == "Fr") | (exp == "Jr" & prev_exp == "So") | (exp == "Sr" & prev_exp == "Jr") | (exp == "Sr" & prev_exp == "Sr"))
# there are 5257 players between the years 2021-2025 that have never transferred and have stayed in Division 1 their entire collegiate career

non_transfers_df <- non_transfers_df |>
  relocate(prev_exp, exp)

non_transfers_df <- non_transfers_df |>
  mutate(exp = ifelse((exp == "Sr" & prev_exp == "Sr"), "SSr", exp))

non_transfers_df <- non_transfers_df |>
  mutate(exp = as.factor(exp))

# there are 1571 super seniors who have never transferred 
#non_transfers_df |>
#  filter(exp == "SSr")

#non_transfers_df <- non_transfers_df |>
#  relocate(prev_exp, exp)
```

Add relevant variables

```{r}
# add team ranking
torvik_ratings_all <- torvik_ratings_all |>
  select(barthag_rk, year, team)


# add only 3 positions

#non_transfers_df |>
#  distinct(pos)

# forward = Wing F, Stretch 4, PF/C
# center = C
# guard = Combo G, Scoring PG, Wing G, Pure PG 

non_transfers_df <- non_transfers_df |>
  mutate(pos = case_when(
    pos %in% c("Wing F", "Stretch 4", "PF/C") ~ "forward",
    pos == "C" ~ "center",
    pos %in% c("Combo G", "Scoring PG", "Wing G", "Pure PG") ~ "guard"
  ))

# getting column with years of experience in the NCAA Division 1
# jk we don't need this
# non_transfers_df <- non_transfers_df |>
#   mutate(prev_prev_year = lag(prev_year),
#          prev_prev_prev_year = lag(prev_prev_year)) |>
#   mutate(years_in_ncaa = case_when(
#     is.na(prev_year) ~ 1,
#     is.na(prev_prev_year) == NA ~ 2,
#     is.na(prev_prev_prev_year) == NA ~ 3, 
#     TRUE ~ 4
#   ))
# 
# non_transfers_df <- non_transfers_df |>
#   relocate(years_in_ncaa)

# attempt to find 6th years



# there are 10 players within non_transfers_df that played a 6 year
# there are 409 players within non_transfers_df that played 5 years
# i will just lump them all in with super senior (aka SSr)

non_transfers_df <- non_transfers_df |>
  mutate(prev_ppg = lag(ppg))

non_transfers_df <- non_transfers_df |>
  relocate(prev_ppg)


non_transfers_df <- non_transfers_df |>
  mutate(prev_ppg = ifelse(exp == "Fr", NA, prev_ppg),
         prev_exp = ifelse(exp == "Fr", NA, prev_exp))

non_transfers_df |>
  filter(exp == "Fr")

# there are 13 double freshmen
# non_transfers_df |>
#  filter(exp == "Fr") |>
#  group_by(id) |>
#  summarise(freq = n()) |>
#  arrange(desc(freq)) |>
#  filter(freq > 1)
```

Attempt simple linear regression

```{r}


#non_transfers_lm <- lm(ppg ~ exp, data = non_transfers_df)
#tidy(non_transfers_lm, conf.int = TRUE)
# this compares the average of everyone

non_transfers_df <- non_transfers_df |>
  mutate(prev_ppg = lag(ppg))

non_transfers_correct_lm <- lm(ppg ~ prev_ppg + exp, data = non_transfers_df)
tidy(non_transfers_correct_lm, conf.int = TRUE)

# variables to use as predictors: team, team rank, position, conference, exp (of course), 
# previous points per game

# change correct variables to factor
non_transfers_df <- non_transfers_df |>
  mutate(across(c(team, pos, conf, exp), as.factor))

non_transfers_2_lm <- lm(ppg ~ prev_ppg + exp + barthag_rk + pos, data = non_transfers_df)

summary(non_transfers_2_lm)
```

Quick EDA

```{r}
non_transfers_df |>
  group_by(exp) |>
  summarise(median = median(ppg)) |>
  ggplot(aes(x = reorder(exp, median), y = median)) +
  geom_col() 
```

Predictors = ppg, prev_ppg, team, team_rank, pos, conference, exp

Level 1 variables = ppg, prev_ppg, team_rank, exp, position

Level 2 variables = team, conference

See if any players change position

```{r}
#non_transfers_df |>
#  mutate(prev_pos = lag(pos)) |>
#  relocate(prev_pos, pos)|>
#  filter(prev_pos != pos) |>
#  select()
```

Attempt at mixed effects model

```{r}
install.packages("Matrix")
#install.packages("lme4", type = "source")
library(Matrix)
library(lme4)
library(nlme)
mod_data <- non_transfers |>
  select(ppg, prev_ppg, barthag_rk, pos, conf, exp, id) |>
  filter(is.na(prev_ppg) != TRUE & is.na(barthag_rk) != TRUE) 
  
mod <- lme(ppg ~ prev_ppg + barthag_rk + pos + conf + exp, random = ~1 | id, data = mod_data)

summary(mod)

VarCorr(mod) 
```

new data

```{r}
data_2021 <- read_csv("../proprietary_data/mbb_202021.csv")
data_2122 <- read_csv("../proprietary_data/mbb_202122.csv")
data_2223 <- read_csv("../proprietary_data/mbb_202223.csv")
data_2324 <- read_csv("../proprietary_data/mbb_202324.csv")
data_2425 <- read_csv("../proprietary_data/mbb_202425.csv")
data_2526 <- read_csv("../proprietary_data/mbb_202526.csv")

data <- bind_rows(data_2021, data_2122, data_2223, data_2324, data_2425, data_2526)
```

ratings table for 2021-2026

```{r}

#df_team_ratings_2022 <- torvik_ratings(year = 2022)
#df_team_ratings_2023 <- torvik_ratings(year = 2023)
#df_team_ratings_2024 <- torvik_ratings(year = 2024)
#df_team_ratings_2025 <- torvik_ratings(year = 2025)
#df_team_ratings_2026 <- torvik_ratings(year = 2026)

#ratings_df <- bind_rows(df_team_ratings_2022, df_team_ratings_2023, 
#                        df_team_ratings_2024, df_team_ratings_2025)

torvik_ratings_all <- bind_rows(
  torvik_ratings(year = 2021) |> mutate(season = 2021),
  torvik_ratings(year = 2022) |> mutate(season = 2022),
  torvik_ratings(year = 2023) |> mutate(season = 2023),
  torvik_ratings(year = 2024) |> mutate(season = 2024),
  torvik_ratings(year = 2025) |> mutate(season = 2025)
)

url_2021 <- "http://barttorvik.com/2021_team_results.csv"
url_2026 <- "http://barttorvik.com/2026_team_results.csv"
torvik_2021 <- read_csv(url_2021, col_names = FALSE)
torvik_2026 <- read_csv(url_2026, col_names = FALSE)


torvik_raw_2021 <- torvik_2021 |> 
  slice(-1) |>
  select(
    rank = X1, 
    team = X2) |>
  mutate(rank = as.integer(rank),
         year = 2021)

torvik_raw_2026 <- torvik_2026 |> 
  slice(-1) |>
  select(
    rank = X1, 
    team = X2) |>
  mutate(rank = as.integer(rank),
         year = 2026)

torvik_ratings_all <- bind_rows(
  torvik_ratings_all, torvik_raw_2021, torvik_raw_2026
)

torvik_ratings_all <- torvik_ratings_all |>
  select(team, rank, year, barthag)
```

```{r}

```

split data into transfer vs non-transfer & some data cleaning

```{r}
# make year column neat using genius_season_name
# year column is the year that seasons championship was played
data <- data |>
  mutate(year = str_sub(GENIUS_SEASON_NAME, start = 1, end = 7)) |>
  relocate(year) |>
  mutate(year = paste0(str_sub(year, 1, 2), str_sub(year, 6, 7))) |>
  mutate(year = as.numeric(year))

data <- data |>
  rename(team = GENIUS_TEAM_NAME)


# id = GENIUS_PERSON_ID
# year = year
# team = team
# prev_team = prev_team
# prev_year = prev_year
# prev_ppg = prev_ppg
# 

data <- data |>
  arrange(GENIUS_PERSON_ID, year) |>
  group_by(GENIUS_PERSON_ID) |>
  mutate(prev_team = lag(team),
         prev_year = lag(year),
         #if they have a previous team AND 
         #the year is the one right after the previous year AND 
         #the team they're on is not the same as the previous team THEN
         #changed_team is through IF NOT THEN
         #changed_team is false 
         changed_team = ifelse(!is.na(prev_team) &
                                 year == (prev_year + 1) &
                                 team != prev_team,
                               TRUE,
                               FALSE)) |>
  ungroup() |>
  relocate(changed_team)

```

join the ratings on new data

```{r}
#data <- data |>
#  left_join(torvik_ratings_all, by = c("team", "year"))

#data |> 
#  summarize(na_count = sum(is.na(rank)))
```

address the 7778 na in the rank column

```{r}
unique(data$team)

unique(torvik_ratings_all$team)

data <- data  |>
  mutate(team = case_when(
    team == "St. John's (NY)"        ~ "St. John's",
    team == "UConn"                  ~ "Connecticut",
    team == "A&M-Corpus Christi"     ~ "Texas A&M Corpus Chris",
    team == "App State"              ~ "Appalachian St.",
    team == "Ark.-Pine Bluff"        ~ "Arkansas Pine Bluff",
    team == "Army West Point"        ~ "Army",
    team == "Bethune-Cookman"        ~ "Bethune Cookman",
    team == "Boston U."              ~ "Boston University",
    team == "California Baptist"     ~ "Cal Baptist",
    team == "Central Ark."           ~ "Central Arkansas",
    team == "Central Conn. St."      ~ "Central Connecticut",
    team == "Central Mich."          ~ "Central Michigan",
    team == "Charleston So."         ~ "Charleston Southern",
    team == "Col. of Charleston"     ~ "Charleston",
    team == "CSU Bakersfield"        ~ "Cal St. Bakersfield",
    team == "CSUN"                   ~ "Cal St. Northridge",
    team == "Eastern Ill."           ~ "Eastern Illinois",
    team == "Eastern Ky."            ~ "Eastern Kentucky",
    team == "Eastern Mich."          ~ "Eastern Michigan",
    team == "ETSU"                   ~ "East Tennessee St.",
    team == "FDU"                    ~ "Fairleigh Dickinson",
    team == "FGCU"                   ~ "Florida Gulf Coast",
    team == "Ga. Southern"           ~ "Georgia Southern",
    team == "Gardner-Webb"           ~ "Gardner Webb",
    team == "Grambling"              ~ "Grambling St.",
    team == "Kansas City"            ~ "UMKC",
    team == "Lamar University"       ~ "Lamar",
    team == "LMU (CA)"               ~ "Loyola Marymount",
    team == "Loyola Maryland"        ~ "Loyola MD",
    team == "ULM"                    ~ "Louisiana Monroe",
    team == "McNeese"                ~ "McNeese St.",
    team == "Miami (FL)"             ~ "Miami FL",
    team == "Miami (OH)"             ~ "Miami OH",
    team == "Middle Tenn."           ~ "Middle Tennessee",
    team == "Mississippi Val."       ~ "Mississippi Valley St.",
    team == "N.C. A&T"               ~ "North Carolina A&T",
    team == "N.C. Central"           ~ "North Carolina Central",
    team == "NC State"               ~ "N.C. State",
    team == "Nicholls"               ~ "Nicholls St.",
    team == "NIU"                    ~ "Northern Illinois",
    team == "Omaha"                  ~ "Nebraska Omaha",
    team == "Ole Miss"               ~ "Mississippi",
    team == "Prairie View"           ~ "Prairie View A&M",
    team == "Queens (NC)"            ~ "Queens",
    team == "Saint Mary's (CA)"      ~ "Saint Mary's",
    team == "Sam Houston"            ~ "Sam Houston St.",
    team == "Seattle U"              ~ "Seattle",
    team == "SFA"                    ~ "Stephen F. Austin",
    team == "SIUE"                   ~ "SIU Edwardsville",
    team == "Southeast Mo. St."      ~ "Southeast Missouri St.",
    team == "Southeastern La."       ~ "Southeastern Louisiana",
    team == "Southern Ind."          ~ "Southern Indiana",
    team == "Southern Miss."         ~ "Southern Miss",
    team == "Southern U."            ~ "Southern",
    team == "St. Francis Brooklyn"   ~ "St. Francis NY",
    team == "St. Thomas (MN)"        ~ "St. Thomas",
    team == "UT Martin"              ~ "Tennessee Martin",
    team == "UIC"                    ~ "Illinois Chicago",
    team == "UMES"                   ~ "Maryland Eastern Shore",
    team == "UNI"                    ~ "Northern Iowa",
    team == "UNCW"                   ~ "UNC Wilmington",
    team == "Southern California"    ~ "USC",
    team == "UTRGV"                  ~ "UT Rio Grande Valley",
    team == "Western Caro."          ~ "Western Carolina",
    team == "Western Ill."           ~ "Western Illinois",
    team == "Western Ky."            ~ "Western Kentucky",
    team == "Western Mich."          ~ "Western Michigan",
    team == "UAlbany"                ~ "Albany",
    team == "Alcorn"                 ~ "Alcorn St.",
    team == "UIW"         ~ "Incarnate Word",
    team == "Eastern Wash."  ~ "Eastern Washington",
    team == "Northern Colo." ~ "Northern Colorado",
    team == "Fla. Atlantic"  ~ "Florida Atlantic",
    team == "Northern Ariz." ~ "Northern Arizona",
    team == "Northern Ky."   ~ "Northern Kentucky",
    team == "South Fla."     ~ "South Florida",
    team == "North Ala."     ~ "North Alabama",
    team == "Southern Ill."  ~ "Southern Illinois",
    TRUE                             ~ team
  ))


data <- data |>
  left_join(torvik_ratings_all, by = c("team", "year"))

data |> 
  summarize(na_count = sum(is.na(rank)))

data <- data |>
  relocate(rank)

data |>
  filter(is.na(rank) == TRUE) |>
  select(team, year) |>
  group_by(team) |>
  summarise(freq = n()) |>
  arrange(desc(freq))

#data <- data |>
#  select(-c(rank.x, rank.x.x, rank.x.x.x, rank.x.x.x.x, rank.x.x.x.x.x, rank.y, rank.y.y, # rank.y.y.y, rank.y.y.y.y.y, rank.y.y.y.y))
```

ok now the na count is 418 which is pretty good

Predictors = ppg, prev_ppg, team, team_rank, pos, conference, exp

```{r}
data <- data |>
  mutate(prev_ppg = lag(POINTS_PG))

data |>
  select(POINTS_PG, prev_ppg, team, rank, pos, GENIUS_CONFERENCE_NAME, YEARS_PLAYED)
```

Add has transferred column

```{r}
transfer_data <- data |>
  group_by(GENIUS_PERSON_ID) |>
  filter(n_distinct(team) > 1) |>
  ungroup()

data <- data |>
  mutate(has_transferred = GENIUS_PERSON_ID %in% transfer_data$GENIUS_PERSON_ID)

```

Add cumulative career transfers

```{r}
career_transfers_by_player <- transfer_data |>
  select(GENIUS_PERSON_ID, year, team) |>
  arrange(GENIUS_PERSON_ID, year) |>
  group_by(GENIUS_PERSON_ID) |>
  mutate(prev_team = lag(team),
         transfer_made = ifelse(!(is.na(prev_team)) & (team != prev_team),
                                1, 0),
         cumulative_career_transfers = cumsum(transfer_made),
         num_transfers = max(cumulative_career_transfers)) |>
  select(GENIUS_PERSON_ID, year, cumulative_career_transfers, num_transfers)

career_transfers_by_player <- career_transfers_by_player |>
  select(cumulative_career_transfers, year, GENIUS_PERSON_ID)

data <- data |>
  left_join(career_transfers_by_player, by = c("GENIUS_PERSON_ID", "year"))
```

simple EDA using new data

#1 An's transfer status and trendline

```{r}
data |>
  select(YEARS_PLAYED, has_transferred, POINTS_PG) |>
  group_by(YEARS_PLAYED, has_transferred) |>
  summarize(mean_ppg = mean(POINTS_PG, na.rm = TRUE)) |>
  mutate(YEARS_PLAYED = factor(YEARS_PLAYED)) |>
  ggplot(aes(x = YEARS_PLAYED, y = mean_ppg, color = as.factor(has_transferred))) +
  geom_point() + geom_line(aes(group = has_transferred))

data |>
  select(YEARS_PLAYED, changed_team, POINTS_PG) |>
  group_by(YEARS_PLAYED, changed_team) |>
  summarize(mean_ppg = mean(POINTS_PG, na.rm = TRUE)) |>
  mutate(YEARS_PLAYED = factor(YEARS_PLAYED)) |>
  ggplot(aes(x = YEARS_PLAYED, y = mean_ppg, color = as.factor(changed_team))) +
  geom_point() + geom_line(aes(group = changed_team))


data |>
  select(YEARS_PLAYED, has_transferred, changed_team, cumulative_career_transfers, POINTS_PG) |>
  filter(YEARS_PLAYED < 7) |>
  mutate(first_transfer_made = ifelse(cumulative_career_transfers > 0, 1, 0)) |>
  select(-cumulative_career_transfers) |>
  
  group_by(YEARS_PLAYED, has_transferred, changed_team, first_transfer_made) |>
  summarize(mean_ppg = mean(POINTS_PG, na.rm = TRUE)) |>
  
  ungroup() |>
  mutate(transfer_type = case_when(
    has_transferred == 0 & changed_team == 0 ~ "Never transferred",
    has_transferred == 1 & changed_team == 1 ~ "Year immediately after transfer",
    has_transferred == 1 & changed_team == 0 & first_transfer_made == 0 ~ "Has not transferred yet",
    has_transferred == 1 & changed_team == 0 & first_transfer_made == 1 ~ "2+ years after last transfer"
    ),
    YEARS_PLAYED = factor(YEARS_PLAYED)) |>
  
  ggplot(aes(x = YEARS_PLAYED, y = mean_ppg, color = transfer_type)) +
  geom_point() + geom_line(aes(group = transfer_type)) +
  theme(legend.position = "bottom")


```

EDA #2

PPG and MPG side by side

```{r}
data |>
  filter(changed_team == TRUE) |>
  filter(POINTS_PG > 2 & prev_ppg > 2) |>
  ggplot(aes(x = prev_ppg, y = POINTS_PG)) +
  geom_point(alpha = 0.6) +
  geom_smooth() +
  geom_abline(intercept = 0, slope = 1, color = "red", linewidth = 1.5) + 
  labs(x = "before transfer",
       y = "after transfer",
       title = "Points per game") +
  theme(axis.title = element_text(size = 10),
        plot.title = element_text(size = 17)) 

data <- data |>
  mutate(prev_mpg = lag(MINUTES_PG))

data |>
  filter(changed_team == TRUE) |>
  filter(MINUTES_PG > 2 & prev_mpg > 2) |>
  ggplot(aes(x = prev_mpg, y = MINUTES_PG)) +
  geom_point(alpha = 0.6) +
  geom_smooth() +
  geom_abline(intercept = 0, slope = 1, color = "red", linewidth = 1.5) +
  labs(x = "before transfer",
       y = "after transfer",
       title = "Minutes per game") +
  theme(axis.title = element_text(size = 10),
        plot.title = element_text(size = 17))

```

EDA #3 Big fish little pond

```{r}
# median points per game before and after a player transfers from a small to big school
data |>
  filter(changed_team == TRUE) |>
  mutate(prev_major = lag(POWER_5_FLAG)) |>
  relocate(prev_major, POWER_5_FLAG) |>
  mutate(prev_major = as.numeric(prev_major),
         POWER_5_FLAG = as.numeric(POWER_5_FLAG)) |>
  filter(prev_major == 0 & POWER_5_FLAG == 1) |>
  summarize(median_prev_ppg = median(prev_ppg, na.rm = TRUE),
    median_ppg  = median(POINTS_PG, na.rm = TRUE)) |>
  pivot_longer(
    cols = everything(),
    names_to = "Season",
    values_to = "Median_PPG"
  ) |>
  ggplot(aes(x = Season, y = Median_PPG)) +
  geom_col(width = 0.4) +
  labs(
    title = "median points per game before and after a player transfers from a small to big school"
  )

#median points per game before and after a player transfers from a big to small school
data |>
  filter(changed_team == TRUE) |>
  mutate(prev_major = lag(POWER_5_FLAG)) |>
  relocate(prev_major, POWER_5_FLAG) |>
  mutate(prev_major = as.numeric(prev_major),
         POWER_5_FLAG = as.numeric(POWER_5_FLAG)) |>
  filter(prev_major == 1 & POWER_5_FLAG == 0) |>
  summarize(median_prev_ppg = median(prev_ppg, na.rm = TRUE),
    median_ppg  = median(POINTS_PG, na.rm = TRUE)) |>
  pivot_longer(
    cols = everything(),
    names_to = "Season",
    values_to = "Median_PPG"
  ) |>
  ggplot(aes(x = Season, y = Median_PPG)) +
  geom_col(width = 0.4) +
  labs(
    title = "median points per game before and after a player transfers from a big to small school"
  )

#median points per game before and after a player transfers from a big to another big school

data |>
  filter(changed_team == TRUE) |>
  mutate(prev_major = lag(POWER_5_FLAG)) |>
  relocate(prev_major, POWER_5_FLAG) |>
  mutate(prev_major = as.numeric(prev_major),
         POWER_5_FLAG = as.numeric(POWER_5_FLAG)) |>
  filter(prev_major == 1 & POWER_5_FLAG == 1) |>
  summarize(median_prev_ppg = median(prev_ppg, na.rm = TRUE),
    median_ppg  = median(POINTS_PG, na.rm = TRUE)) |>
  pivot_longer(
    cols = everything(),
    names_to = "Season",
    values_to = "Median_PPG"
  ) |>
  ggplot(aes(x = Season, y = Median_PPG)) +
  geom_col(width = 0.4) +
  labs(
    title = "median points per game before and after a player transfers from a big to another school"
  )

#median points per game before and after a player transfers from a small to another small school
data |>
  filter(changed_team == TRUE) |>
  mutate(prev_major = lag(POWER_5_FLAG)) |>
  relocate(prev_major, POWER_5_FLAG) |>
  mutate(prev_major = as.numeric(prev_major),
         POWER_5_FLAG = as.numeric(POWER_5_FLAG)) |>
  filter(prev_major == 0 & POWER_5_FLAG == 0) |>
  summarize(median_prev_ppg = median(prev_ppg, na.rm = TRUE),
    median_ppg  = median(POINTS_PG, na.rm = TRUE)) |>
  pivot_longer(
    cols = everything(),
    names_to = "Season",
    values_to = "Median_PPG"
  ) |>
  ggplot(aes(x = Season, y = Median_PPG)) +
  geom_col(width = 0.4) +
  labs(
    title = "median points per game before and after a player transfers from a small to another small school"
  )
```

EDA #4 transferring and looking at team rating

```{r}
# find average change in team ranking when it comes to transferring
data <- data |>
  relocate(barthag)

barthag_data <- data |>
  filter((is.na(rank) == FALSE) & (changed_team == TRUE)) |>
  mutate(prev_barthag = lag(barthag)) |>
  mutate(change_in_barthag = prev_barthag - barthag)
  

ggplot(barthag_data, aes(x = change_in_barthag)) +
  geom_histogram(aes(y = after_stat(density)), 
                 fill = "lightblue", 
                 color = "black", 
                 alpha = 0.7) +
  stat_function(fun = dnorm,
                args = list(mean = mean(barthag_data$change_in_barthag, na.rm = TRUE),
                            sd   = sd(barthag_data$change_in_barthag, na.rm = TRUE)),
                color = "darkred",
                linewidth = 1.5) +
  theme_minimal()

```

find overperformer and underperformer

```{r}
data <- data |>
  filter(is.na(POINTS_PG) != TRUE) |>
  group_by(team, year) |>
  mutate(team_avg_ppg = mean(POINTS_PG)) |>
  ungroup() |>
  relocate(team_avg_ppg) |>
  mutate(is_overperformer = ifelse(POINTS_PG >= team_avg_ppg, TRUE, FALSE)) |>
  relocate(is_overperformer)
```

EDA #5 using is_overperformer

```{r}
data |>
  ggplot(aes(x = factor(changed_team), fill = factor(is_overperformer))) +
  geom_bar(position = "fill") +
  labs(
    x = "changed team",
    y = "proportion",
    fill = "is overperformer"
  )
```

Try some of An's modelling

```{r}

clean_data <- data |>
  filter((!is.na(changed_team) & (!is.na(prev_ppg))))

#install.packages("MatchIt")
library(MatchIt)

match_simplest <- matchit(changed_team ~ prev_ppg,
                          data = clean_data, method = "nearest", distance = "glm",
                          replace = TRUE, ratio = 1)
match_simplest_summary <- summary(match_simplest)

match_simplest_summary
```
