```{r}
# Clear the environment and read the prepared dataset
rm(list = ls()) # Clear the environment
library(tidyverse)
library(cbbdata)

df <- read.csv("player_stats.csv")
df
```

#### 1. Plot the 1D and 2D distributions of variables

Summarize all variables in the dataset:

```{r}
summary(df)
```

Many quantitative variables look very left-skewed! How about examining the skewness of each variable?

```{r}
# I will compute them quantitatively later...
```


I will plot the histogram of most quantitative variable indicators of player performance in this dataset, as some inferece models ([like hypothesis testing in linear regression models](https://library.virginia.edu/data/articles/normality-assumption)) prefer normally distributed variables.

```{r}
quantitative_df <- df |>
  select(where(is.numeric))

quantitative_df
```


```{r}
# https://stackoverflow.com/questions/63334436/pivot-longer-on-all-columns
quantitative_df |>
  select(1:20) |>
  pivot_longer(cols = everything(), names_to = "metric",
               values_to = "value") |>
  ggplot(aes(x = value)) +
  geom_density() +
  facet_wrap(~ metric, scales = "free")
```

```{r}
# Continue plotting for other relevant quantitative variables
quantitative_df |>
  select(21:40) |>
  pivot_longer(cols = everything(), names_to = "metric",
               values_to = "value") |>
  ggplot(aes(x = value)) +
  geom_density() +
  facet_wrap(~ metric, scales = "free")
```

```{r}
# Continue plotting for other relevant quantitative variables
quantitative_df |>
  select(41:60) |>
  pivot_longer(cols = everything(), names_to = "metric",
               values_to = "value") |>
  ggplot(aes(x = value)) +
  geom_density() +
  facet_wrap(~ metric, scales = "free")
```

Check the correlation between relevant quantitative variables with a heat map:

```{r}
# https://www.sthda.com/english/wiki/ggplot2-quick-correlation-matrix-heatmap-r-software-and-data-visualization

# install.packages(reshape2)
library(reshape2)

corr_matrix <- quantitative_df |>
  cor(use = "pairwise.complete.obs") |>
  round(2) |>
  melt() |>
  # Highlight the correlations that are greater than 0.85 in absolute value
  # (both positive and negative)
  mutate(plot_value = ifelse(value > 0.85 | value < -0.85, value, NA))
  
corr_matrix
```

```{r}
corr_matrix |>
  ggplot(aes(x = Var1, y = Var2, fill = plot_value)) +
  geom_tile()
```

This one looks messy, but things will clear up when we know which variables to include in our model. For now, there are some pairs of variables with high correlation.

#### 2. Explore differences in player metrics based on their transfer status

As a refresher, the documentation for the original dataset, `torvik_player_season`, can be found [here](https://github.com/andreweatherman/cbbdata/blob/main/data/dictionary_torvik_player_season.csv).

Comparing players who have transferred in their careers and players who do not:

```{r}
# Note that we have removed some NA values to properly display the results
df |>
  select(g, min, mpg, ppg, has_transferred) |>
  group_by(has_transferred) |>
  summarize(median_g = median(g, na.rm = TRUE),
            median_min = median(min, na.rm = TRUE),
            median_mpg = median(mpg, na.rm = TRUE),
            median_ppg = median(ppg, na.rm = TRUE)) |>
  pivot_longer(median_g:median_ppg, names_to = "metric", values_to = "value") |>
  ggplot(aes(x = has_transferred, y = value)) +
  geom_col() + coord_flip() +
  scale_x_continuous(breaks = seq(0, 1, 1)) +
  facet_wrap(~ metric, scales = "free")
```

Are these differences statistically significant? Perhaps we can use a statistical test later...

```{r}
# Add code to perform the statistical test here
```

[MORE COMING SOON]

#### 3. Explore the transferring data for teams and conferences

Which specific team has the most number of transfer-ins and transfer-outs during the 2021-25 periods and for each season?

(Note that we do not have data outside of the 2021-25 range. The transfer-ins data starts from the 2022-23 season and the transfer-outs data ends at the 2023-24 season.)

```{r}
# Filter the dataset to only obtain athletes' first year after transfer
transfer_ins <- df |>
  filter(changed_team == 1) |>
  select(team, year)

transfer_ins
```

```{r}
# Teams with the most transfer-ins each year
transfer_ins |>
  group_by(team, year) |>
  count(team) |>
  ungroup() |>
  group_by(year) |>
  slice_max(n, n = 10) |>
  ggplot(aes(x = n, y = fct_reorder(team, n))) +
  geom_col() + facet_wrap(~ year, scales = "free_x")
```

What about the teams which have the most players transferring out?

```{r}
df |>
  filter(changed_team == 1) |>
  select(prev_team, prev_year) |>
  group_by(prev_team, prev_year) |>
  count(prev_team) |>
  ungroup() |>
  group_by(prev_year) |>
  slice_max(n, n = 5) |>
  ggplot(aes(x = n, y = fct_reorder(prev_team, n))) +
  geom_col() + facet_wrap(~ prev_year, scales = "free_x")
```

In the presentation, we might link these EDA graph's result to management or coaching changes. Additional research is needed to confirm if that is the case.

[MORE COMING SOON]

#### 4. Explore the general developments of player statistics over time ("time trends")

As a simple metric, I will compare the median points per game of players in freshman, sophomore, junior, and senior years, grouped by different metrics.

First, comparing between players who have transferred at least once in their career, and players who have not:

```{r}
# https://www.r-bloggers.com/2024/06/how-to-avoid-empty-line-charts/

df |>
  select(exp, has_transferred, ppg) |>
  group_by(exp, has_transferred) |>
  summarize(mean_ppg = mean(ppg, na.rm = TRUE)) |>
  mutate(exp = factor(exp, levels = c("Fr", "So", "Jr", "Sr"))) |>
  ggplot(aes(x = exp, y = mean_ppg, color = as.factor(has_transferred))) +
  geom_point() + geom_line(aes(group = has_transferred))
```

What about players of the same experience level (freshman up to senior), who have just transferred into a new team?

```{r}
df |>
  select(exp, changed_team, ppg) |>
  group_by(exp, changed_team) |>
  summarize(mean_ppg = mean(ppg, na.rm = TRUE)) |>
  mutate(exp = factor(exp, levels = c("Fr", "So", "Jr", "Sr"))) |>
  ggplot(aes(x = exp, y = mean_ppg, color = as.factor(changed_team))) +
  geom_point() + geom_line(aes(group = changed_team))
```

How about combining these plots together?

```{r}
df |>
  select(exp, has_transferred, changed_team, cumulative_career_transfers, ppg) |>
  mutate(first_transfer_made = ifelse(cumulative_career_transfers > 0, 1, 0)) |>
  select(-cumulative_career_transfers) |>
  
  group_by(exp, has_transferred, changed_team, first_transfer_made) |>
  summarize(mean_ppg = mean(ppg, na.rm = TRUE)) |>
  
  ungroup() |>
  mutate(transfer_type = case_when(
    has_transferred == 0 & changed_team == 0 ~ "Never transferred",
    has_transferred == 1 & changed_team == 1 ~ "Year immediately after transfer",
    has_transferred == 1 & changed_team == 0 & first_transfer_made == 0 ~ "Has not transferred yet",
    has_transferred == 1 & changed_team == 0 & first_transfer_made == 1 ~ "2+ years after last transfer"
    ),
    exp = factor(exp, levels = c("Fr", "So", "Jr", "Sr"))) |>
  
  ggplot(aes(x = exp, y = mean_ppg, color = transfer_type)) +
  geom_point() + geom_line(aes(group = transfer_type)) +
  theme(legend.position = "bottom")
```
