Some useful resources:

- Paper mentioning the package: [(link)](https://imai.fas.harvard.edu/research/files/matchit.pdf)

- More on ICC and a section that mentions singular model fit, where the between-group variance is estimated as 0: [(link)](https://library.virginia.edu/data/articles/understanding-icc)

- Generate publication-like regression tables in R: [(link)](https://r-statistics.co/Regression-Tables-in-R.html)

- Sample economics paper on how to assemble a mixed effects model: [(link)](https://link.springer.com/article/10.1007/s10834-026-10076-w/tables/4)

```{r}
# Clear the environment and read the prepared dataset
rm(list = ls()) # Clear the environment

library(tidyverse)
library(broom)
library(lme4) # For mixed effects models
library(MatchIt) # For propensity matching
library(modelsummary) # Display the results of regression models
library(clubSandwich) # For adjusting standard errors (observations are not independent)
library(merDeriv)

df <- read.csv("player_stats_hoopR.csv")
df
```

```{r}
# Filter out for NA values in the dataset (check later)
# Change necessary variable data types for the models to work properly
df <- df |>
  mutate(id = as.factor(id),
         player = as.factor(player),
         pos = as.factor(pos),
         team = as.factor(team),
         conf = as.factor(conf)) |>
  # Filter out players who have changed team and skipped the year before
  # (they are likely to transfer out of DI to lower divisions or JUCOs)
  filter(!(skipped_year_before == 1 & changed_team == 1)) |>
  drop_na(changed_team, prev_ppg, pos, exp, year, current_barthag, current_barthag_conf,
          id, team, conf)
  # We don't use data influenced by the pandemic period
```

```{r}
# How many observations remain?
df # 12613 observations
df |>
  count(changed_team) # 4278 transferred, 8335 not transferred
```

More on arguments of the `matchit` command: [(link)](https://cran.r-project.org/web/packages/MatchIt/vignettes/matching-methods.html)

Selecting variables for propensity score matching: [(link)](https://link.springer.com/article/10.1007/s10654-019-00494-6)

#### 1. Propensity matching throughout all the dataset

Some variables that affects performance (i.e. previous points per game) but does not necessarily affect players' decisions to transfer include:

- Position
- Experience level (freshman to super-senior)
- Conference classification: whether they are low, high, or mid-major
- Additional variables: team and conference competitiveness

*(cite the variable selection paper, 2006)*

Here, we are currently using sampling with replacement to ensure that every transferred players find a match.

```{r}
match_object <- matchit(changed_team ~ pos + exp + current_barthag + current_barthag_conf,
                        data = df, method = "nearest", distance = "glm",
                        replace = FALSE, ratio = 2)

match_object_summary <- summary(match_object)

match_object_summary
```

Not so many observations are matched, even with replacement... We might be losing a lot of useful trends and information among NCAA Division I men's basketball players!

```{r}
# How balanced are the matching?
plot(match_object, type = "density")
```

```{r}
# Extract the largest standardized mean differences in a matchit summary dataset
max_std_mean_diff <- function(matchit_summary) {
  
  # Extract the standardized mean differences of covariates among all data and matched data
  sum_all_mean_diff <- as.data.frame(matchit_summary$sum.all) |>
    select(`Std. Mean Diff.`)
  sum_matched_mean_diff <- as.data.frame(matchit_summary$sum.matched) |>
    select(`Std. Mean Diff.`)
  
  # Combine two datasets together and return the maximum value of the column
  result_df <- bind_rows(sum_all_mean_diff, sum_matched_mean_diff)
  result <- max(result_df$`Std. Mean Diff.`)
  
  return(result)
}

max_std_mean_diff(match_object_summary)
```

Checking the balance in covariates between matched and unmatched data:

```{r}
plot(match_object_summary, var.order = "unmatched",
     xlim = c(0, max_std_mean_diff(match_object_summary)))
```

```{r}
# match.data(<matchit object>) represents the dataframe that has been matched
match_object_data <- match.data(match_object)

match_object_data
```

```{r}
# Use the propensity matching weights in downstream regression
# Note that if we have random intercepts for conferences, summaries just take ages to run!
mixed_effects_matched <- lmer(ppg ~ changed_team + prev_ppg + pos + exp + year + conf +
                                current_barthag + current_barthag_conf + (1|id/team),
                              data = match_object_data, weights = weights)

# https://modelsummary.com/vignettes/modelsummary_extension.html
# https://modelsummary.com/vignettes/appearance.html
summary(mixed_effects_matched)
# modelsummary(mixed_effects_matched, stars = TRUE)
```

#### 2. Controls for transfers as an underperformer

```{r}
df_underperf <- df |>
  mutate(underperf_transfer = changed_team * is_prev_underperformer) |>
  select(-changed_team) |>
  rename(changed_team = underperf_transfer)

df_underperf
```

We will still do the same matching and regression workflow as mentioned earlier:

```{r}
# Generate a match object and extract data that has been matched
match_object_underperf <- matchit(changed_team ~ pos + exp + current_barthag +
                                    current_barthag_conf,
                                  data = df_underperf, method = "nearest", distance = "glm",
                                  replace = FALSE, ratio = 2)
match_object_underperf_data <- match.data(match_object_underperf)

# Implement a mixed effects regression model
mixed_effects_underperf <- lmer(ppg ~ changed_team + prev_ppg + pos + exp + year + conf +
                                  current_barthag + current_barthag_conf + (1|id/team),
                                data = match_object_underperf_data, weights = weights)

# https://modelsummary.com/vignettes/modelsummary_extension.html
# https://modelsummary.com/vignettes/appearance.html
summary(mixed_effects_underperf)
# modelsummary(mixed_effects_underperf, stars = TRUE)
```

#### 3. Does the result change if we only count transfers as an overperformer?

```{r}
# Adjust the changed_team variable: 1 if transferred as an overperformer, 0 if not
df_overperf <- df |>
  mutate(overperf_transfer = changed_team * is_prev_overperformer) |>
  select(-changed_team) |>
  rename(changed_team = overperf_transfer)

# Generate a match object and extract data that has been matched
match_object_overperf <- matchit(changed_team ~ pos + exp + current_barthag +
                                    current_barthag_conf,
                                  data = df_overperf, method = "nearest", distance = "glm",
                                  replace = FALSE, ratio = 2)
match_object_overperf_data <- match.data(match_object_overperf)

# Implement a mixed effects regression model
mixed_effects_overperf <- lmer(ppg ~ changed_team + prev_ppg + pos + exp + year + conf +
                                  current_barthag + current_barthag_conf + (1|id/team),
                                data = match_object_overperf_data, weights = weights)

# https://modelsummary.com/vignettes/modelsummary_extension.html
# https://modelsummary.com/vignettes/appearance.html
summary(mixed_effects_overperf)
# modelsummary(mixed_effects_underperf, stars = TRUE)
```

#### The entire regression table

```{r}
models <- list(
  "(1)" = mixed_effects_matched,
  "(2)" = mixed_effects_underperf,
  "(3)" = mixed_effects_overperf
)

# modelsummary(models, stars = TRUE)
modelsummary(models, stars = TRUE,
             coef_map = c(
               changed_team = "Transfer",
               prev_ppg = "Previous points per game",
               posforward = "Position is Forward",
               posguard = "Position is Guard",
               exp = "Experience level",
               year = "Season (ending year)",
               current_barthag = "Team rating",
               current_barthag_conf = "Conference rating"
             ))
```

#### DRAFT AND APPENDIX

```{r, eval = FALSE}
match_object <- matchit(changed_team ~ id + player + pos + exp + team + conf +
                          prev_ppg + current_barthag,
                        data = df, method = "nearest", distance = "glm",
                        replace = TRUE, ratio = 1)
saveRDS(match_object, file = "match_object.rds")
```

```{r, eval = FALSE}
# match_object <- readRDS("match_object.rds")
```

Coming soon: May consider inverse probability score weighting (not omitting so many observations).